---
title: "Built-in add_* functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Built-in add_* functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(data.table)
library(swereg)
```

## What `add_*` functions do

`add_*` functions are the workhorse of swereg: they take raw registry
data and fold it into a skeleton created by
`create_skeleton()`. Every `add_*` function shares the same shape:

- **First argument** is the skeleton.
- **Second argument** is a registry `data.table`.
- The skeleton is **mutated in place** via
  `skeleton[data, on = ..., := ...]`. No copy, no return.
- New columns are appended to the skeleton with names you choose (for
  the pattern-matching functions, names come from the `codes` list).

This vignette walks through every `add_*` function swereg ships with,
using the synthetic datasets included in the package. For writing your
*own* `add_*` for a registry swereg doesn't support, see
`vignette("custom-add-functions")`.

## Setup: a skeleton we can reuse

```{r}
data("fake_person_ids", package = "swereg")

skeleton <- create_skeleton(
  ids      = fake_person_ids[1:25],
  date_min = "2019-01-01",
  date_max = "2020-12-31"
)

head(skeleton, 3)
```

The skeleton has one row per (person, week) plus one "annual" row per
(person, year) marked by `is_isoyear == TRUE`. All `add_*` functions
respect this layout.

## `add_onetime()` — baseline / demographic data

Use `add_onetime()` for a single value per person that never changes
(date of birth, sex at birth, country of origin, etc.). The function
broadcasts the value to every row belonging to that person.

```{r}
data("fake_demographics", package = "swereg")
swereg::make_lowercase_names(fake_demographics)

add_onetime(skeleton, fake_demographics, id_name = "lopnr")

skeleton[, .(id, isoyearweek, fodelseman, doddatum)] |> head(3)
```

Every row for a given `id` now carries the same `fodelseman` and
`doddatum` values.

**Column-name collision policy**: if the skeleton already has a
column with the same name as something in `data`, `add_onetime()`
keeps the existing skeleton column and prefixes the incoming one with
`i.` (data.table's standard collision marker).

## `add_annual()` — yearly data for a specific year

Registries that report once per calendar year (SCB LISA income, family
composition, etc.) go through `add_annual()`. You call it *once per
year*, and values are attached to `(id, isoyear)` rows.

```{r}
data("fake_annual_family", package = "swereg")
swereg::make_lowercase_names(fake_annual_family)

add_annual(skeleton, fake_annual_family, id_name = "lopnr", isoyear = 2020L)

skeleton[isoyear == 2020 & is_isoyear == TRUE, .(id, isoyear, famtyp)] |> head(3)
```

**Gotcha**: the value only lands on `is_isoyear == TRUE` rows for the
matching year. The weekly rows (`is_isoyear == FALSE`) keep `NA`. If
you want the value forward-filled onto the weeks, do that explicitly
with a `nafill`-style step after all `add_annual()` calls finish.

## `add_diagnoses()` — ICD codes from the NPR

The National Patient Register holds main (`hdia`) and secondary
(`dia1`, `dia2`, …) diagnoses per hospital contact. `add_diagnoses()`
takes a named list of ICD-code prefixes and writes one boolean
column per entry, `TRUE` on every week that person had a matching
diagnosis.

```{r}
data("fake_diagnoses", package = "swereg")
swereg::make_lowercase_names(fake_diagnoses, date_columns = "indatum")

add_diagnoses(
  skeleton,
  fake_diagnoses,
  id_name  = "lopnr",
  diag_type = "both",
  codes = list(
    "dep"     = c("F32", "F33"),
    "anxiety" = c("F40", "F41")
  )
)

skeleton[dep == TRUE, .(id, isoyearweek, dep)] |> head(3)
```

**Pattern syntax** (shared with `add_operations()`, `add_cods()`,
`add_icdo3s()`, `add_snomed3s()`, `add_snomedo10s()`):

- `"F32"` — prefix match (all codes starting with F32).
- `"^F320"` — the leading `^` is implicit; written out for emphasis.
- `"!F321"` — **negation**: drops matches starting with F321 from the
  positive set. Use this to carve out a specific sub-code.
- Multiple entries in one vector are OR'd together.

**`diag_type`**: `"both"` searches `hdia`, `dia*`, `ekod*`, `icd7*`,
and `icd9*`. `"main"` searches `hdia` only — useful when you want the
primary reason for admission, not every code listed.

**Splitting by source** (inpatient vs outpatient): the dataset
includes a `source` column. Filter first, then call `add_diagnoses()`
twice:

```r
in_only <- fake_diagnoses[source == "inpatient"]
add_diagnoses(skeleton, in_only, "lopnr", codes = list("dep_inp" = c("F32", "F33")))
```

## `add_operations()` — surgical procedure codes

Same pattern as `add_diagnoses()`, but searches the `op1`, `op2`, …
columns of the NPR.

```{r}
add_operations(
  skeleton,
  fake_diagnoses,
  id_name = "lopnr",
  codes = list("mastectomy" = c("HAC10", "HAC20"))
)
```

## `add_rx()` — prescription drugs (LMED)

Prescriptions aren't instantaneous: a dispense covers a treatment
period from the prescription date (`edatum`) to `edatum + fddd` days
(`fddd` = defined daily doses dispensed). `add_rx()` converts these
intervals to ISO week ranges and flags every week the patient was
covered.

```{r}
data("fake_prescriptions", package = "swereg")
swereg::make_lowercase_names(fake_prescriptions, date_columns = "edatum")

add_rx(
  skeleton,
  fake_prescriptions,
  id_name = "p444_lopnr_personnr",
  codes = list(
    "antidep" = "N06A",
    "hormones" = c("G03", "L02AE")
  ),
  source = "atc"
)

skeleton[antidep == TRUE, .(id, isoyearweek, antidep)] |> head(3)
```

**`source = "atc"`** (default) uses prefix matching via `startsWith()`
on the `atc` column. **`source = "produkt"`** uses exact matching via
`%chin%` on the `produkt` (product name) column — useful when ATC
granularity isn't enough (e.g. distinguishing formulations).

**Gotcha**: prescriptions with `NA` `fddd` or negative durations are
dropped with a warning. They can't be placed on the week grid.

## `add_cods()` — cause of death

Cause-of-death records have one "underlying" cause (`ulorsak`) and
multiple contributing causes (`morsak1`, `morsak2`, …). `add_cods()`
lets you search either or both.

```{r}
data("fake_cod", package = "swereg")
swereg::make_lowercase_names(fake_cod, date_columns = "dodsdat")

add_cods(
  skeleton,
  fake_cod,
  id_name = "lopnr",
  cod_type = "both",
  codes = list(
    "cardio_death"   = c("I21", "I22"),
    "external_death" = c("X60", "X70")
  )
)
```

## `add_quality_registry()` — arbitrary event registries

Quality registries (Riksstroke, NKBC, …) don't follow the ICD-prefix
model. `add_quality_registry()` takes **quoted R expressions** evaluated
against the dataset columns; each expression yields one boolean flag per
week.

```{r}
fake_registry <- data.table::data.table(
  lopnr      = fake_person_ids[1:3],
  event_date = as.Date(c("2019-03-15", "2020-06-20", "2020-11-01")),
  severity   = c(5L, 18L, 22L),
  treated    = c(1L, 2L, 1L)
)

add_quality_registry(
  skeleton,
  fake_registry,
  id_name = "lopnr",
  date_col = "event_date",
  codes = list(
    "reg_event"   = TRUE,
    "reg_severe"  = quote(severity >= 15),
    "reg_treated" = quote(treated == 1)
  )
)
```

`TRUE` marks every week with any record. `quote(...)` flags the weeks
where the expression holds. Aggregation within a week is `any()`, and
`NA` is coerced to `FALSE` ("no evidence" is not the same as "true").

## Typical end-to-end workflow

The built-ins layer on top of each other. A realistic ordering:

```{r, eval = FALSE}
skeleton <- create_skeleton(ids, "2005-01-01", "2023-12-31")

add_onetime(skeleton, demographics, "lopnr")            # baseline
for (y in 2005:2023) {
  add_annual(skeleton, lisa[year == y], "lopnr", y)     # annual
}
add_diagnoses(skeleton, npr,     "lopnr", codes = ...)  # events
add_operations(skeleton, npr,    "lopnr", codes = ...)
add_rx(skeleton, lmed, "p444_lopnr_personnr", codes = ...)
add_cods(skeleton, cod,          "lopnr", codes = ...)
```

After this you typically derive row-independent (`ri_*`) variables
with `make_rowind_first_occurrence()` — see the "rowdep vs rowind"
vignette.

## If swereg doesn't ship an `add_*` for your registry

Write your own. The contract is small and the pipeline enforces it
automatically when you register your function via
`RegistryStudy$register_codes()`. See `vignette("custom-add-functions")`
— same mechanism works for Norwegian registries, regional Swedish
cohorts, payer claims, and anything else with a longitudinal
event-per-row shape.
