---
title: "Writing your own add_* function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing your own add_* function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(data.table)
library(swereg)
```

## When you need a custom `add_*`

swereg ships `add_*` functions for the common Swedish registries (NPR,
LMED, cause of death, quality registries, SCB one-time / annual). If
your registry doesn't fit those patterns — a regional cohort, a
research registry, a non-Swedish source — you'll need to write your
own. The good news is that the contract is small and mechanical.

This vignette walks through the contract, shows a complete
worked example, and documents the lessons learned from the built-ins
so your function avoids the sharp edges.

## The `add_*` contract

A well-behaved `add_*` function has a short list of hard
requirements (violations are bugs):

1. **Takes `skeleton` as its first argument** and mutates it in place
   using `skeleton[data, on = ..., := ...]`. It does **not** return a
   new skeleton, does **not** use `merge()`, does **not** assign back
   with `<-`.
2. **Preserves row count**. If `nrow(skeleton)` changes, something
   went wrong.
3. **Preserves the four structural columns** (`id`, `isoyear`,
   `isoyearweek`, `is_isoyear`). These are read-only.
4. **Adds one column per entry in the user's `codes` list**, named
   exactly as the list is named. Users expect
   `add_foo(skeleton, ..., codes = list(x = ..., y = ...))` to produce
   columns `x` and `y`.
5. **Takes `id_name` as an explicit argument**. Different Swedish
   registries use different ID column names after
   `make_lowercase_names()` (`lopnr`, `p444_lopnr_personnr`, …), so
   don't hard-code.

These are enforced by `skeleton_snapshot()` +
`validate_skeleton_after_add()`.

There's also a **soft** requirement worth being aware of but not
usually worth spending RAM to avoid:

- **Input-data mutation is fine in the batched pipeline, messy in
  manual workflows.** Many built-ins add scratch columns to the input
  `data.table` (`add_rx()` adds `start_date`/`stop_date`, etc.). In
  the `RegistryStudy` pipeline this is invisible — the batch data is
  loaded, fed to one `add_*`, then discarded. But in an interactive
  or test workflow where you hold the data in a variable and reuse
  it, stray columns are surprising. The validator can flag this
  **opt-in** — see the "Mutation warnings" section below. Avoiding
  mutation costs you a `copy()` of the input, which on millions of
  registry rows is not free.

## The validator

`skeleton_snapshot()` captures a cheap fingerprint of the skeleton
before your function runs (row count, column set, and — optionally —
the same for the input dataset). `validate_skeleton_after_add()`
compares that fingerprint to the post-state and errors / warns on any
violation of the contract.

```{r}
data("fake_person_ids", package = "swereg")
skeleton <- create_skeleton(
  ids      = fake_person_ids[1:5],
  date_min = "2020-01-01",
  date_max = "2020-12-31"
)

# Take the snapshot before running an add_*
snap <- skeleton_snapshot(skeleton)

# ... your add_* function runs here and mutates skeleton in place ...

# After your add_* returns, validate:
validate_skeleton_after_add(
  skeleton,
  snap,
  expected_new_cols = character(0),   # nothing expected in this toy example
  context           = "demo"
)
```

The checks (and when they fire):

| Check | Outcome | When it fires |
|---|---|---|
| `skeleton` still a `data.table` | **error** | You did `skeleton <- merge(...)` or similar. |
| `nrow` unchanged | **error** | Non-equi join multiplied rows, or inner join dropped rows. |
| Structural columns present | **error** | You accidentally dropped `id`, `isoyear`, `isoyearweek`, or `is_isoyear`. |
| `expected_new_cols` all added | **error** | Your loop ran but didn't actually write columns. |
| Some new columns added | warning | `codes` was non-empty but no columns appeared. |
| Input data not mutated | warning (**opt-in**) | Only when the caller passed `input_data = ...`. |

Errors indicate bugs. The input-mutation check only fires when the
caller explicitly passes `input_data` to both
`skeleton_snapshot()` and `validate_skeleton_after_add()` — it's
useful in tests and manual workflows, but deliberately silent by
default since input mutation is harmless in the batched pipeline.

## A complete worked example: `add_vaccinations()`

Suppose you have a regional vaccination registry. Each row is one
dose, with columns for person ID (`personnummer`), dose date
(`vac_date`), and vaccine code (`atc_vac`). You want to flag weeks in
which a person received any dose matching a user-supplied ATC prefix
list.

### Design decisions up front

- **Point events, not intervals**: unlike `add_rx()`, a vaccine dose
  is a single date. So we snap `vac_date` to an `isoyearweek` and
  write `TRUE` on that week only.
- **Collision policy**: overwrite, matching `add_diagnoses()` and
  friends. Initialise new columns to `FALSE`, then set `TRUE` on
  matching weeks.
- **Pattern semantics**: ATC prefixes, same as `add_rx(source = "atc")`.
  Use `startsWith()` for speed.

### The function

```{r}
add_vaccinations <- function(
  skeleton,
  dataset,
  id_name,
  date_col,
  codes = list()
) {
  # 1. Capture pre-state for post-hoc validation. This also asserts
  #    that `skeleton` is a data.table with the four structural columns,
  #    so we don't need a separate check for that.
  snap <- swereg::skeleton_snapshot(skeleton, input_data = dataset)

  # 2. Check everything specific to this registry.
  stopifnot(
    data.table::is.data.table(dataset),
    id_name %in% names(dataset),
    date_col %in% names(dataset),
    inherits(dataset[[date_col]], "Date"),
    is.list(codes),
    length(codes) > 0L,
    !is.null(names(codes)),
    "atc_vac" %in% names(dataset)
  )

  # 3. Initialise output columns to FALSE.
  for (nm in names(codes)) skeleton[, (nm) := FALSE]

  # 4. Work on a local copy so we don't mutate the caller's dataset.
  local_ds <- dataset[, .(
    id   = get(id_name),
    iyw  = cstime::date_to_isoyearweek_c(get(date_col)),
    atc  = atc_vac
  )]

  # 5. For each code entry, find matching rows and flag weeks.
  for (nm in names(codes)) {
    patterns <- codes[[nm]]
    hits <- Reduce(`|`, lapply(patterns, function(p) startsWith(local_ds$atc, p)))
    matches <- unique(local_ds[which(hits), .(id, isoyearweek = iyw)])
    skeleton[matches, on = .(id, isoyearweek), (nm) := TRUE]
  }

  # 6. Enforce the contract.
  swereg::validate_skeleton_after_add(
    skeleton,
    snap,
    expected_new_cols = names(codes),
    input_data        = dataset,
    context           = "add_vaccinations()"
  )

  invisible(skeleton)
}
```

### Using it

```{r}
fake_vax <- data.table::data.table(
  personnummer = rep(fake_person_ids[1:3], each = 2),
  vac_date     = as.Date(c(
    "2020-03-12", "2020-10-05",
    "2020-04-20", "2020-11-02",
    "2020-06-15", "2020-12-14"
  )),
  atc_vac      = c("J07BB02", "J07BX03", "J07BB02", "J07BB02", "J07CA02", "J07BX03")
)

add_vaccinations(
  skeleton,
  fake_vax,
  id_name  = "personnummer",
  date_col = "vac_date",
  codes    = list(
    "flu_vax"   = "J07BB",
    "covid_vax" = "J07BX"
  )
)

skeleton[flu_vax == TRUE | covid_vax == TRUE,
         .(id, isoyearweek, flu_vax, covid_vax)] |> head()
```

### Mutation warnings: when and when not to opt in

If you call `validate_skeleton_after_add()` with `input_data =
fake_vax` *and* the snapshot was taken with the same object, the
validator warns on any scratch columns left behind. In the
`add_vaccinations()` above we computed `local_ds` as a copy and
never touched `dataset`, so no warning fires.

If you *don't* pass `input_data` (or pass it to only one side), no
input check runs. That's the right default in the
`RegistryStudy` batched pipeline where the input is thrown away
after each call. Opt in when:

- Writing tests that assert the registry data stays untouched.
- Iterating interactively and reusing the dataset variable.
- Tracking down a bug where "my registry data has weird columns"
  is a suspected root cause.

Several built-ins (`add_annual`, `add_diagnoses`, `add_rx`,
`add_cods`) *do* mutate their inputs — they use the registry
data.table as scratch space. That's not technical debt; it's a
RAM / CPU tradeoff (copying millions of registry rows to avoid a
temporary column isn't free). If you're writing a new `add_*` and
you're willing to pay the copy cost for hygiene, do it. If you're
writing for the production pipeline, don't.

### What the validator caught during development

When writing this for real, you'll hit at least one of these. All
come from mistakes the built-ins have historically made:

- **Forgetting to initialise columns** → `skeleton[matches, ..., (nm) := TRUE]`
  leaves non-matching weeks as `NA` instead of `FALSE`. The validator
  won't catch this (column is present), but downstream code
  `sum(flu_vax)` will propagate `NA`.
- **Mutating the input dataset (opt-in check)** →
  `dataset[, iyw := cstime::date_to_isoyearweek_c(get(date_col))]`
  leaves a stray `iyw` column on the caller's `fake_vax`. Whether
  this matters depends on usage: harmless in the batched pipeline,
  confusing in manual / test workflows. The validator warns *only*
  when the caller passes `input_data = dataset` on both sides.
- **Accidentally returning a `merge()` result** → assigning
  `skeleton <- merge(skeleton, ...)` replaces `skeleton` in *your*
  environment but the caller's binding still points at the original.
  Their `skeleton` silently lacks the new columns. The validator
  catches this because the caller's `skeleton` (passed back to the
  validator) is still the unmutated original; `expected_new_cols` are
  missing.
- **Non-equi join multiplies rows** → joining on ATC prefix without
  `unique()`-ing the matches first can insert duplicate
  `(id, isoyearweek)` pairs and expand the skeleton. The validator
  catches this via the row-count check.

## Lessons from the built-ins — design cheat sheet

These are patterns the built-in `add_*` functions have converged on
(sometimes after making the mistake first). When in doubt, copy one
of the built-ins.

### Do

- **Snapshot first, validate last.** The snapshot/validate pair is
  ~3 lines at the top and 7 at the bottom. Cheap insurance.
- **Use update-by-reference joins.**
  `skeleton[data, on = c("id==<id_name>", "isoyearweek"), (cols) := ...]`
  is the canonical pattern. Never `merge()`, never `left_join()`.
- **Initialise all new columns to `FALSE` (or another sentinel)
  before the join.** Otherwise rows without matches get `NA` and
  propagate through downstream logic.
- **Respect `is_isoyear`.** The skeleton has both weekly
  (`is_isoyear == FALSE`) and yearly (`is_isoyear == TRUE`) rows.
  Decide explicitly whether each new column belongs on weekly rows,
  yearly rows, or both.
- **Use `startsWith()` over regex for prefix matching.** Roughly 5x
  faster and clearer to read.
- **Take `id_name` as a parameter.** Don't hard-code `"lopnr"` — LMED
  uses `p444_lopnr_personnr`.
- **Warn on partial ID matches.** If most of the skeleton's people
  aren't in the registry data, the user probably made a mistake.

### Don't

- **Don't reassign `skeleton`.** Any line that looks like
  `skeleton <- something` inside an `add_*` is a bug.
- **Don't assume the user ran `make_lowercase_names()`.** Check.
  Fail with an actionable error if required columns are missing.
- **Don't silently drop rows with `NA` in critical columns.** Warn,
  and state how many were dropped. (See `add_rx()` for an example
  of the right behaviour.)

### Judgment call: input-data mutation

Mutating the input `data.table` (adding scratch columns, coercing
date formats, etc.) is **fine** in the `RegistryStudy` batched
pipeline — the input is discarded after each call. It's **messy**
in manual or interactive workflows where the caller keeps the
dataset around. Several built-ins mutate; several don't. If you're
writing a new `add_*` and the cost of a `copy()` of the input is
tolerable, prefer the cleaner approach. If the input is large and
you're writing for the pipeline, don't bother. Either way, be
consistent within a single function.

## Summary

1. Take a snapshot: `snap <- skeleton_snapshot(skeleton, input_data = dataset)`.
2. Validate the inputs with the usual structural checks.
3. Do the join in-place with `skeleton[data, on = ..., := ...]`.
4. Validate the outputs:
   `validate_skeleton_after_add(skeleton, snap, expected_new_cols = names(codes),
   input_data = dataset, context = "my_add_fn()")`.
5. Return `invisible(skeleton)`.

See `vignette("builtin-add-functions")` for every `add_*` swereg
ships, and read the source of any of them for a battle-tested
reference implementation.
