---
title: "Writing your own add_* function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing your own add_* function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(data.table)
library(swereg)
```

## Who this is for

swereg's built-in `add_*` functions cover the common Swedish
registries (NPR, LMED, cause of death, quality registries, SCB
one-time and annual). They do not cover everything. You will need a
custom `add_*` if you work with:

- **Non-Swedish registries** — Norwegian NPR, Danish LPR, FinDep,
  etc. They follow similar longitudinal patterns but use different
  column names and coding systems.
- **Swedish registries swereg doesn't ship hard-coded support for** —
  SWEDEHEART, RIKSHIP, regional cohorts, research registries.
- **In-house data** — payer claims, EHR extracts, institutional
  biobanks with time-stamped events.

The mechanism is open-ended by design. You cannot write code for
every registry anyone might want to ingest; the right answer is a
clean contract that user-written functions can plug into.

## Two ways to use a custom `add_*`

1. **Register it via `RegistryStudy$register_codes()`.** The pipeline
   auto-validates the contract on every call, fingerprints your
   function, and replays it on change for incremental rebuilds. This
   is the recommended path for any production-style use.
2. **Call it directly as a free function** in the manual workflow.
   No automatic validation — you catch mistakes through your own
   tests. Fine for one-off exploration, risky at scale.

The rest of this vignette focuses on path (1), because swereg
enforces the contract for you there. The function body is the same
either way.

## The `add_*` contract

A well-behaved `add_*` function:

1. **Takes `skeleton` as its first argument** and mutates it in place
   using `skeleton[data, on = ..., := ...]`. No `merge()`, no
   `left_join()`, no `skeleton <- ...` reassignment.
2. **Preserves row count.** If `nrow(skeleton)` changes, the
   pipeline stops the run. Phase 1b, the trim registered with
   `RegistryStudy$register_trim()`, is the one phase that MAY delete
   skeleton rows.
3. **Preserves the four structural columns** (`id`, `isoyear`,
   `isoyearweek`, `is_isoyear`). Read-only.
4. **Adds one column per entry in the `codes` list**, named exactly
   as the list is named. If the caller passes
   `codes = list(flu_vax = ..., covid_vax = ...)`, the skeleton
   gains columns `flu_vax` and `covid_vax`.
5. **Takes `id_name` as an explicit argument.** Different registries
   use different ID column names; don't hard-code.
6. **Accepts `codes` as a named list** and receives any extra
   configuration through positional / named arguments the caller
   passes via `fn_args` (see below).

## Column-slot headroom

A table's headroom is its count of free column slots. data.table over-allocates
the list that holds a table's columns, and `:=` writes a new column into a free
slot, in place. Past the last free slot data.table allocates a longer list. A
longer list is a new R object, so `[.data.table` rebinds the grown table in the
frame it ran in.

Inside your `add_*` that frame belongs to your function. The caller keeps the
old table and never gets the new columns. data.table 1.18.4 reports nothing: no
error, no warning.

The failure is silent where it happens and loud somewhere else. The column
count does not move, every call returns, and the run stops later on a name
that is missing. Read the free slot count first when a column you asked for
is not there.

data.table gives a new table 1024 free slots on top of its column count, and
restores 1024 whenever it grows one. qs2 keeps none of them, so
`swereg::qs2_read()` restores them: 4096 for a skeleton, 1024 for every other
table. So the hazard needs a table that spent its slack, or a table that
reached you from `qs2::qs_read()` rather than from `swereg::qs2_read()`.

Every shipped `add_*` grows the list first, writes the grown table back to its
caller where it can, and returns it either way. `.ensure_dt_alloc()` in
`R/dt_alloc.R` does that. A function you write yourself carries the hazard.

**The rest of this section is about a function you call yourself.** A function
registered with `RegistryStudy$register_codes()` runs on slots the pipeline has
already reserved.

### Reproducing it

Two steps set a table to an exact free slot count.
`data.table::setalloccol()` never shrinks, so it cannot do it alone.

```{r}
fresh <- data.table(a = 1:3)
truelength(fresh) - ncol(fresh)
truelength(setalloccol(fresh, 3L)) - ncol(fresh)
```

Serialization drops the over-allocation, which is what a `.qs2` file on disk
does to a table. `setalloccol()` then sets the exact number back. Its `n` is the
free slot count on top of `ncol()`, not the total.

```{r}
no_headroom <- function(dt, free = 0L) {
  return(setalloccol(unserialize(serialize(dt, NULL)), free))
}
```

Now write an `add_*` with no guard, and give it a skeleton with no free slots.

```{r}
add_flag <- function(skeleton, dataset, id_name, codes = list()) {
  for (nm in names(codes)) skeleton[, (nm) := FALSE]
  return(invisible(NULL))
}

sk_tight <- no_headroom(
  create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")
)
truelength(sk_tight) - ncol(sk_tight)

add_flag(sk_tight, vax_batch, "id", codes = list(flu_vax = TRUE))
"flu_vax" %in% names(sk_tight)
```

The call returns normally and the column is missing.

### Reserve the slots in the caller

`.ensure_dt_alloc()` is internal, so your own function cannot call it. Your tool
is `data.table::setalloccol()`, and where you call it decides whether it works.

`setalloccol()` rebinds in its own calling frame. Called inside your `add_*`, it
grows that function's local and leaves the caller's table alone.

```{r}
inside_only <- function(skeleton, new_col) {
  setalloccol(skeleton, 1000L)
  skeleton[, (new_col) := TRUE]
  return(invisible(NULL))
}

sk_inside <- no_headroom(
  create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")
)
inside_only(sk_inside, "flu_vax")
"flu_vax" %in% names(sk_inside)
```

Call it in the frame that owns the table, before the add call.

```{r}
sk_caller <- no_headroom(
  create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")
)
sk_caller <- setalloccol(sk_caller, 1000L)

add_flag(sk_caller, vax_batch, "id", codes = list(flu_vax = TRUE))
"flu_vax" %in% names(sk_caller)
```

Reserve generously. One free slot costs 16 bytes, measured on R 4.5.2 and
data.table 1.18.4 at 3, 10,000 and 200,000 rows alike. A thousand free slots
therefore cost 16 KB per table.

**The return value is the other remedy, and since 26.10.14 it covers every
shipped function.** All eight exported `add_*` return the skeleton, invisibly.
So do `make_rowind_first_occurrence()` and the six `skeleton_eligible_*()`
functions. Write `sk <- add_diagnoses(sk, ...)` and the reservation stops
mattering to you. swereg exports no helper for the reservation itself, because
`data.table::setalloccol()` already is one.

Pass a variable, never an expression. The guard writes the grown table back to a
name, or to a `$`, `[[` or `@` chain that ends at a name. An expression gives it
no binding to write to, so it warns instead.

```{r}
tight <- function() {
  return(no_headroom(
    create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")
  ))
}

add_annual(
  tight(),
  data.table(lopnr = fake_person_ids[1:3], lisa_x = 1:3),
  id_name = "lopnr",
  isoyear = 2020
)
```

### Count the temporaries

`make_rowind_first_occurrence()` needs two free slots, not one. It writes a
`temp` column, then the new column, then deletes `temp`. Both columns sit on the
table at the same time, so one free slot is not enough. At one free slot the
caller keeps the orphan `temp` and never gets the new column. The next call then
collides with the leftover `temp`.
`skeleton_eligible_no_observation_in_window_excluding_wk0()` holds a
`.temp_obs_*` column the same way.

Any scratch column you build carries the same cost. This function needs two free
slots for one output column:

```{r}
add_first_year <- function(skeleton, flag_col, new_var) {
  skeleton[get(flag_col) == TRUE, temp := isoyear]
  skeleton[, (new_var) := first_non_na(temp), by = .(id)]
  skeleton[, temp := NULL]
  return(invisible(NULL))
}

build_hits <- function(free) {
  sk <- create_skeleton(fake_person_ids[1:3], "2019-01-01", "2020-12-31")
  sk[, hit := isoyear == 2020]
  return(no_headroom(sk, free))
}

one_slot <- build_hits(1L)
add_first_year(one_slot, "hit", "ri_year_first_hit")
c(
  temp = "temp" %in% names(one_slot),
  new = "ri_year_first_hit" %in% names(one_slot)
)
```

Two free slots reach the caller with the output column and no orphan.

```{r}
two_slots <- build_hits(2L)
add_first_year(two_slots, "hit", "ri_year_first_hit")
c(
  temp = "temp" %in% names(two_slots),
  new = "ri_year_first_hit" %in% names(two_slots)
)
unique(two_slots$ri_year_first_hit)
```

### The rebind reaches one frame

A helper that wraps an `add_*` call keeps the grown table in its own local. The
helper's own caller sees nothing, and nothing warns, because the rebind
succeeded. data.table 1.18.4 behaves this way.

```{r}
via_helper <- function(skeleton) {
  swereg::skeleton_eligible_isoyears(skeleton, 2020)
  return("eligible_isoyears" %in% names(skeleton))
}

sk_wrapped <- no_headroom(
  create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")
)
via_helper(sk_wrapped)
"eligible_isoyears" %in% names(sk_wrapped)
```

A helper like `via_helper()` MUST reserve the slots itself. Its own caller MAY
reserve them instead, before the helper runs.

## How the pipeline enforces the contract

When you pass a function to `RegistryStudy$register_codes(..., fn =
my_add)`, the pipeline reserves the column slots your codes need and
then wraps the call with a pre/post check. If your function writes
into the table it was given, leaves the structural columns alone, and
adds every expected new column, you see nothing. If it misbehaves,
`$process_skeletons()` (or `$apply_codes_to_skeleton()` directly)
errors with a pointer back to the registration:

```
$register_codes(my_add): did not add the expected columns: covid_vax.
Check that your loop over `names(codes)` actually writes to the
skeleton (e.g. `skeleton[..., (nm) := TRUE]`).
```

You don't call any validation helpers yourself. Just write the
function and register it.

## Common case: reusing a built-in via registration

Most "custom" work isn't actually custom code — it's using a
built-in through the registration path so the pipeline's hash
tracking and incremental rebuild apply. If the registry you're
ingesting looks like Swedish NPR data (columns for `hdia`, `dia*`,
`ekod*`, etc. after `make_lowercase_names()`), you don't need your
own function at all:

```r
study$register_codes(
  codes  = list("depression" = c("F32", "F33"), "anxiety" = c("F40", "F41")),
  fn     = swereg::add_diagnoses,
  groups = list("npr" = c("inpatient", "outpatient")),
  label  = "add_diagnoses_mood"
)
```

If the built-in takes extra knobs — e.g. `diag_type = "main"` to
search only the primary diagnosis — pass them through `fn_args`:

```r
study$register_codes(
  codes   = list("primary_mi" = "I21"),
  fn      = swereg::add_diagnoses,
  groups  = list("npr" = c("inpatient")),
  fn_args = list(diag_type = "main"),
  label   = "add_diagnoses_mi_primary"
)
```

`fn_args` is the bucket for anything the fn accepts beyond the four
mandatory arguments (`skeleton`, `data`, `id_name`, `codes`). It
participates in the entry's fingerprint, so changing
`diag_type = "main"` to `diag_type = "both"` triggers replay.

Reach for a custom `add_*` only when the built-ins genuinely can't
express what you need (different column names, different matching
semantics, a non-ICD coding system, etc.).

## A complete worked example: `add_vaccinations()`

Suppose you have a regional vaccination registry (Swedish, Norwegian,
or other — doesn't matter). Each row is one dose, with columns
`personnummer` (person ID), `vac_date` (Date), and `atc_vac` (vaccine
ATC code). You want one boolean column per user-supplied ATC prefix,
`TRUE` on every week a person received a matching dose.

### Design decisions

- **Point events, not intervals**: a vaccine dose is a single date.
  Snap `vac_date` to an `isoyearweek` and write `TRUE` on that week.
- **Collision policy**: overwrite. Match `add_diagnoses()` — init
  columns to `FALSE`, then set `TRUE` on matching weeks.
- **Pattern semantics**: ATC prefixes (same as
  `add_rx(source = "atc")`). Use `startsWith()` for speed.

### The function

```{r}
add_vaccinations <- function(
  skeleton,
  dataset,
  id_name,
  codes = list()
) {
  stopifnot(
    data.table::is.data.table(dataset),
    id_name %in% names(dataset),
    "vac_date" %in% names(dataset),
    inherits(dataset$vac_date, "Date"),
    "atc_vac" %in% names(dataset),
    is.list(codes),
    length(codes) > 0L,
    !is.null(names(codes))
  )

  # Initialise output columns to FALSE so non-matching weeks aren't NA.
  for (nm in names(codes)) skeleton[, (nm) := FALSE]

  # Work on a local projection so we don't mutate the caller's dataset.
  local_ds <- dataset[, .(
    id  = get(id_name),
    iyw = cstime::date_to_isoyearweek_c(vac_date),
    atc = atc_vac
  )]

  for (nm in names(codes)) {
    patterns <- codes[[nm]]
    hits <- Reduce(`|`, lapply(patterns, function(p) startsWith(local_ds$atc, p)))
    matches <- unique(local_ds[which(hits), .(id, isoyearweek = iyw)])
    skeleton[matches, on = .(id, isoyearweek), (nm) := TRUE]
  }
}
```

### Registering it with a `RegistryStudy`

A minimal `RegistryStudy` scoped to `tempdir()` so the vignette is
self-contained:

```{r}
study <- RegistryStudy$new(
  data_rawbatch_dir = tempfile("raw"),
  data_skeleton_dir = tempfile("skel"),
  group_names       = c("vax_registry"),
  id_col            = "id"
)

study$register_codes(
  codes  = list("flu_vax" = "J07BB", "covid_vax" = "J07BX"),
  fn     = add_vaccinations,
  groups = list("vax_registry"),
  label  = "add_vaccinations"
)
```

### Applying it

```{r}
data("fake_person_ids", package = "swereg")

skeleton <- create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")

vax_batch <- data.table::data.table(
  id       = rep(fake_person_ids[1:3], each = 2),
  vac_date = as.Date(c(
    "2020-03-12", "2020-10-05",
    "2020-04-20", "2020-11-02",
    "2020-06-15", "2020-12-14"
  )),
  atc_vac  = c("J07BB02", "J07BX03", "J07BB02", "J07BB02", "J07CA02", "J07BX03")
)

batch_data <- list(vax_registry = vax_batch)

study$apply_codes_to_skeleton(skeleton, batch_data)

skeleton[flu_vax | covid_vax, .(id, isoyearweek, flu_vax, covid_vax)] |> head()
```

Note there is no explicit snapshot/validate call in user code. The
wrapper runs automatically inside `$apply_codes_to_skeleton()` — any
contract violation would have errored here.

### Seeing the wrapper fire on a broken function

Let's deliberately write a broken `add_*` that forgets to create the
requested columns and see what happens:

```{r}
broken_add_vax <- function(skeleton, dataset, id_name, codes = list()) {
  # BUG: loops but writes to a typo'd column name
  for (nm in names(codes)) skeleton[, (paste0(nm, "_typo")) := FALSE]
}

study2 <- RegistryStudy$new(
  data_rawbatch_dir = tempfile("raw"),
  data_skeleton_dir = tempfile("skel"),
  group_names       = c("vax_registry"),
  id_col            = "id"
)

study2$register_codes(
  codes  = list("flu_vax" = "J07BB"),
  fn     = broken_add_vax,
  groups = list("vax_registry"),
  label  = "broken_add_vax"
)

sk2 <- create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")

tryCatch(
  study2$apply_codes_to_skeleton(sk2, batch_data),
  error = function(e) cat("ERROR caught:\n", conditionMessage(e), "\n")
)
```

The wrapper named the registration (`broken_add_vax`) and the missing
column (`flu_vax`) and told us what to look for. Four more errors fire
the same way:

- a changed row count
- a changed row order
- a dropped column
- a returned table that is not the one the wrapper passed in

### Registration removes the hazard

`RegistryStudy$register_codes()` reserves the slots for you, before your
function runs. Your function then writes in place, and the pipeline keeps what
it wrote. A function that needs more slots than the reservation still works:
the pipeline compares column vectors, so it can tell a growth from a
replacement.

Here is a skeleton with no free slot, the shape that lost its column above,
applied through the registry:

```{r}
sk_reg <- no_headroom(
  create_skeleton(fake_person_ids[1:3], "2020-01-01", "2020-12-31")
)
truelength(sk_reg) - ncol(sk_reg)

study$apply_codes_to_skeleton(sk_reg, batch_data)
c(flu = "flu_vax" %in% names(sk_reg), covid = "covid_vax" %in% names(sk_reg))
```

Both columns are there, from a table that started with no free slot at all.
This is why registration is the recommended path.


## Failure modes the wrapper catches

All of these come from real bugs people hit writing their first
custom `add_*`:

- **Building a new table instead of writing into the one you were
  given.** Each of `copy()`, `merge()`, `rbind()`,
  `dplyr::left_join()` and a `.()` projection builds new column
  vectors. The wrapper records the address of every column vector
  before your function runs and compares them after. It refuses a
  replacement however that replacement reaches the pipeline.
- **Writing your columns onto a table you built, then returning it.**
  The wrapper ignores that table, so the skeleton never gets the
  columns and the expected-column check reports them missing.
- **Non-equi joins multiplying rows**. Joining on ATC prefix without
  `unique()`-ing the matches can insert duplicate `(id, isoyearweek)`
  pairs. The wrapper catches the row-count change.
- **Writing to a structural column.** The wrapper compares `id`,
  `isoyear`, `isoyearweek` and `is_isoyear` by value against the ones
  it took before the call. That catches a row swap that keeps the row
  count, and a `skeleton[i, isoyear := x]` that keeps the type.
- **Looping over `names(codes)` but writing to a typo'd column name**.
  As demonstrated above.
- **Dropping any column**, structural or not. The wrapper checks that
  every column present before the call is present after it.
- **Running out of column slots and returning `NULL`.** The grown
  table is then the only one with your columns, and you threw it
  away. Return the skeleton.

One failure mode the wrapper does **not** catch: initialising to
`NA` instead of `FALSE`. The column exists, so the wrapper is happy,
but downstream code that does `sum(col)` or `col & other_col` will
propagate `NA`. Always initialise to `FALSE` (or whatever non-`NA`
sentinel makes sense for your type).

## What the wrapper guarantees, and what it cannot

It guarantees three things after your function runs:

1. Every column your `codes` list names is present.
2. `id`, `isoyear`, `isoyearweek` and `is_isoyear` satisfy
   `identical()` against the ones your function started with,
   attributes included.
3. No column that was there before is dropped, and none is replaced by
   a new vector.

**It cannot see a write INTO an existing column vector.** Your function
receives the skeleton by reference, so
`set(skeleton, i, "an_old_column", value)` changes that column and
moves nothing the wrapper measures. A new table header that shares the
original vectors passes the same way. Neither of those is new in
26.10.14: a registered function could always do it. Write only the
columns your `codes` list names.

An error from the wrapper stops the run and records nothing.
`$process_skeletons()` never writes that batch to disk. Treat the
in-memory skeleton as unusable afterwards. It MAY carry your writes and
it MAY not: the answer depends on whether the entry grew the table.

A table you return is ignored unless it is the skeleton itself. That
means it holds every column the skeleton held before, each at the same
vector address. A table the pipeline grew for you satisfies that. A
table you built does not.

## Design cheat sheet

Patterns the built-in `add_*` functions have converged on (sometimes
after making the mistake first). Copy them when in doubt.

### Do

- **Use update-by-reference joins.**
  `skeleton[data, on = c("id==<id_name>", "isoyearweek"), (cols) := ...]`
  is the canonical pattern. Never `merge()`, never `left_join()`.
- **Initialise new columns to `FALSE`** (or another non-`NA`
  sentinel) before the join.
- **Respect `is_isoyear`.** The skeleton has both weekly and annual
  rows. Decide explicitly whether each new column belongs on weekly
  rows, annual rows, or both.
- **Use `startsWith()` over regex for prefix matching.** ~5x faster.
- **Take `id_name` as a parameter.** Don't hard-code `"lopnr"`.
- **Warn on partial ID matches.** If most skeleton IDs aren't in the
  registry data, the user probably made a mistake.
- **Count every column your function holds at once**, temporaries
  included. See the section "Column-slot headroom".

### Don't

- **Don't reassign `skeleton`.** Any line that looks like
  `skeleton <- something` inside an `add_*` is a bug.
- **Don't assume the user ran `make_lowercase_names()`.** Check.
  Fail with an actionable error if required columns are missing.
- **Don't silently drop rows with `NA` in critical columns.** Warn
  with a count. (See `add_rx()` for the right pattern.)

### Judgment call: input-data mutation

Mutating the input `data.table` (adding scratch columns, coercing
date formats) is **fine** in the `RegistryStudy` batched pipeline —
the input is discarded after each call. It's **messy** in manual or
interactive workflows where the caller reuses the dataset variable.
Several built-ins mutate; several don't. If copy-cost is tolerable,
prefer the cleaner approach. If the input is huge and you're writing
for the pipeline, don't bother.

## Summary

1. Check whether a built-in already does what you need, and wrap
   *that* in `register_codes(codes = ..., fn = swereg::add_diagnoses,
   ...)` instead of writing a custom function.
2. If you genuinely need custom code, write it following the 6-point
   contract above.
3. Register it via
   `RegistryStudy$register_codes(codes = ..., fn = my_add, groups = ...)`.
   The pipeline validates the contract on every call and replays
   your function when the registration changes.
4. Read the source of any built-in `add_*` for a battle-tested
   reference implementation.

See `vignette("builtin-add-functions")` for every shipped `add_*`,
and `vignette("skeleton-pipeline")` for the full four-phase
pipeline.
