---
title: "Cleaning and deriving variables (skeleton2_clean)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning and deriving variables (skeleton2_clean)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
```

## Introduction

This vignette demonstrates **skeleton2_clean** - the second stage of the swereg workflow where raw integrated data is cleaned and analysis-ready variables are created.

**Prerequisites**: Complete the "Building the data skeleton (skeleton1_create)" vignette first, as this stage builds directly on skeleton1_create output.

## What is skeleton2_clean?

The skeleton2_clean stage focuses on:

- **Data cleaning**: Using only data within the skeleton (no external joins)
- **Variable derivation**: Creating composite indicators and summary variables
- **Quality filters**: Removing invalid observations and applying study criteria
- **Analysis preparation**: Creating variables ready for statistical modeling

This stage transforms raw integrated data into clean, analysis-ready variables.

## Step 1: load skeleton1_create output

Start with a completed skeleton1_create (we'll recreate it for demonstration):

```{r}
# Load all fake data
data("fake_person_ids", package = "swereg")
data("fake_demographics", package = "swereg")
data("fake_annual_family", package = "swereg") 
data("fake_inpatient_diagnoses", package = "swereg")
data("fake_outpatient_diagnoses", package = "swereg")
data("fake_prescriptions", package = "swereg")
data("fake_cod", package = "swereg")

# Apply make_lowercase_names to all datasets
fake_demographics <- copy(fake_demographics)
fake_annual_family <- copy(fake_annual_family)
fake_inpatient_diagnoses <- copy(fake_inpatient_diagnoses)
fake_outpatient_diagnoses <- copy(fake_outpatient_diagnoses)
fake_prescriptions <- copy(fake_prescriptions)
fake_cod <- copy(fake_cod)
swereg::make_lowercase_names(fake_demographics, date_column = "fodelseman")
swereg::make_lowercase_names(fake_annual_family)
swereg::make_lowercase_names(fake_inpatient_diagnoses, date_column = "INDATUM")
swereg::make_lowercase_names(fake_outpatient_diagnoses, date_column = "INDATUM")
swereg::make_lowercase_names(fake_prescriptions, date_column = "EDATUM")
swereg::make_lowercase_names(fake_cod, date_column = "dodsdat")

# Create skeleton covering 2015-2020
skeleton <- swereg::create_skeleton(
  ids = fake_person_ids,
  date_min = "2015-01-01",
  date_max = "2020-12-31"
)

# Add demographics
swereg::add_onetime(skeleton, fake_demographics, id_name = "lopnr")

# Add annual data
swereg::add_annual(skeleton, fake_annual_family, id_name = "lopnr", isoyear = 2015)

# Add diagnoses
diagnoses_combined <- rbindlist(list(
  fake_inpatient_diagnoses,
  fake_outpatient_diagnoses
), use.names = TRUE, fill = TRUE)

swereg::add_diagnoses(
  skeleton,
  diagnoses_combined,
  id_name = "lopnr",
  diags = list(
    "depression" = c("F32", "F33"),
    "anxiety" = c("F40", "F41"),
    "gender_dysphoria" = c("F64"),
    "psychosis" = c("F20", "F25")
  )
)

# Add prescriptions
swereg::add_rx(
  skeleton,
  fake_prescriptions,
  id_name = "p444_lopnr_personnr",
  rxs = list(
    "antidepressants" = c("N06A"),
    "antipsychotics" = c("N05A"),
    "hormones" = c("G03")
  )
)

# Add cause of death
swereg::add_cods(
  skeleton,
  fake_cod,
  id_name = "lopnr",
  cods = list(
    "external_death" = c("X60", "X70"),
    "cardiovascular_death" = c("I21", "I22")
  )
)

cat("skeleton1_create completed:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
```

## Step 2: data cleaning operations

Now clean and derive variables using only data within the skeleton:

### Create age variable

```{r}
# Create age variable
skeleton[, birth_year := as.numeric(substr(fodelseman, 1, 4))]
skeleton[, age := isoyear - birth_year]

cat("Age variable created\n")
```

### Create mental health composite variables

```{r}
# Create mental health composite variables
skeleton[, any_mental_health := depression | anxiety | psychosis]
skeleton[, severe_mental_illness := psychosis | gender_dysphoria]

# Check mental health prevalence
cat("Any mental health condition:", sum(skeleton$any_mental_health, na.rm = TRUE), "person-periods\n")
cat("Severe mental illness:", sum(skeleton$severe_mental_illness, na.rm = TRUE), "person-periods\n")
```

### Create medication concordance variables

```{r}
# Create medication concordance variables
skeleton[, depression_treated := depression & antidepressants]
skeleton[, psychosis_treated := psychosis & antipsychotics]

# Check treatment patterns
cat("Depression with treatment:", sum(skeleton$depression_treated, na.rm = TRUE), "periods\n")
cat("Psychosis with treatment:", sum(skeleton$psychosis_treated, na.rm = TRUE), "periods\n")
```

### Create life stage variables

```{r}
# Create life stage variables
skeleton[, life_stage := fcase(
  age < 18, "child",
  age >= 18 & age < 65, "adult", 
  age >= 65, "elderly",
  default = "unknown"
)]

# Check life stage distribution
cat("Life stage distribution:\n")
print(table(skeleton[is_isoyear == TRUE]$life_stage, useNA = "ifany"))
```

### Create outcome variables

```{r}
# Create outcome variables
skeleton[, death_any := external_death | cardiovascular_death]

cat("Any death:", sum(skeleton$death_any, na.rm = TRUE), "deaths\n")
```

## Step 3: quality filters and validation

Apply study criteria and quality filters:

```{r}
# Filter to valid ages and reasonable time periods
cat("Before filtering:", nrow(skeleton), "rows\n")

skeleton <- skeleton[age >= 0 & age <= 100]
skeleton <- skeleton[isoyear >= 2015]  # Remove historical rows

cat("After filtering:", nrow(skeleton), "rows\n")
```

## Step 4: create study design variables

Create variables for case-control or cohort study designs:

```{r}
# Create registry tag variables (simulate case-control study)
skeleton[, register_tag := fcase(
  gender_dysphoria == TRUE, "case",
  id %% 3 == 0, "control_matched",
  default = "control_population"
)]

# Create shared case variables (for matched studies)
# Find first gender dysphoria diagnosis for cases
gd_first <- skeleton[gender_dysphoria == TRUE & register_tag == "case", 
                     .(first_gd_year = min(isoyear, na.rm = TRUE)), 
                     by = .(id)]

# Add to skeleton
skeleton[gd_first, on = "id", first_gd_year := first_gd_year]

# For controls, assign their matched case's first GD year (simplified)
skeleton[register_tag != "case", first_gd_year := 2016]  # Simplified for demo

cat("Study design variables created\n")
print(table(skeleton[is_isoyear == TRUE]$register_tag))
```

## Step 5: remove temporary variables

Clean up intermediate variables:

```{r}
# Remove temporary variables
skeleton[, c("fodelseman", "birth_year") := NULL]

cat("skeleton2_clean completed:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
```

## Cleaned dataset summary

The cleaned skeleton2 now contains derived variables ready for analysis:

```{r}
# Show structure
cat("Variables:", paste(names(skeleton), collapse = ", "), "\n")

# Example analysis: Depression prevalence by life stage (filter to years with data)
depression_summary <- skeleton[is_isoyear == TRUE & isoyear >= 2015, .(
  n_person_years = .N,
  depression_prev = mean(depression, na.rm = TRUE),
  treatment_rate = ifelse(sum(depression, na.rm = TRUE) > 0,
                         mean(depression_treated[depression == TRUE], na.rm = TRUE),
                         NA_real_)
), by = .(life_stage, register_tag)]

print(depression_summary[n_person_years > 0])  # Only show non-empty groups
```

```{r}
# Example: Mental health treatment patterns (filter to valid data)
treatment_summary <- skeleton[any_mental_health == TRUE & is_isoyear == TRUE & !is.na(register_tag), .(
  antidepressant_use = mean(antidepressants, na.rm = TRUE),
  antipsychotic_use = mean(antipsychotics, na.rm = TRUE),
  hormone_use = mean(hormones, na.rm = TRUE),
  mean_age = mean(age, na.rm = TRUE),
  n_observations = .N
), by = register_tag]

print(treatment_summary[n_observations > 0])  # Only show groups with data
```

## Key principles for skeleton2_clean

### Data cleaning strategy
1. **Self-contained**: skeleton2_clean uses only data within skeleton
2. **Derived variables**: Create analysis variables from raw data
3. **Quality filters**: Remove invalid observations
4. **Clinical indicators**: Create meaningful composite variables

### Variable creation patterns
- **Composite indicators**: Combine multiple boolean variables (e.g., `any_mental_health`)
- **Treatment concordance**: Match diagnoses with treatments (e.g., `depression_treated`)
- **Life course variables**: Create age-based categories and life stages
- **Study design variables**: Create case/control indicators and matched cohort variables

## Understanding the skeleton2_clean output

The skeleton2_clean output contains:

- **Clean variables**: Validated age, filtered time periods
- **Derived indicators**: Composite mental health variables, treatment patterns
- **Study variables**: Case-control tags, cohort definitions
- **Summary measures**: Person-level aggregations for annual data
- **Analysis-ready format**: Variables ready for statistical modeling

## Next steps

This **skeleton2_clean** provides clean, analysis-ready variables for statistical modeling. The data has been validated, derived variables created, and study design implemented.

**For large datasets**: If you have huge datasets (>100,000 individuals) and limited RAM, see the "Batching (skeleton3_analyze)" vignette to learn memory-efficient processing techniques.

**Analysis ready**: For most analyses, you can proceed directly with descriptive statistics, regression modeling, or survival analysis using the cleaned skeleton2 data.