---
title: "Analysing the skeleton"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing the skeleton}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
```

## Introduction

This vignette demonstrates how to work with a finished skeleton --
collapsing weekly data to yearly summaries, creating analysis datasets,
and running common epidemiological queries.

**Prerequisites**: Complete `vignette("skeleton-create")` first, or
use the `RegistryStudy` pipeline from
`vignette("skeleton-pipeline")`.

## Build a skeleton for this vignette

We create a small skeleton with the same steps as
`vignette("skeleton-create")`, condensed here for convenience:

```{r}
skeleton <- swereg::create_skeleton(swereg::fake_person_ids, "2015-01-01", "2020-12-31")

# Demographics
fake_demographics <- swereg::fake_demographics |>
  data.table::copy() |>
  swereg::make_lowercase_names(date_columns = "fodelseman")
swereg::add_onetime(skeleton, fake_demographics, id_name = "lopnr")

# Diagnoses
fake_diagnoses <- swereg::fake_diagnoses |>
  data.table::copy() |>
  swereg::make_lowercase_names(date_columns = "indatum")
swereg::add_diagnoses(skeleton, fake_diagnoses, id_name = "lopnr",
  diags = list(
    "depression" = c("F32", "F33"),
    "anxiety" = c("F40", "F41"),
    "gender_dysphoria" = c("F64"),
    "psychosis" = c("F20", "F25")
  ))

# Prescriptions
fake_prescriptions <- swereg::fake_prescriptions |>
  data.table::copy() |>
  swereg::make_lowercase_names(date_columns = "edatum")
swereg::add_rx(skeleton, fake_prescriptions, id_name = "p444_lopnr_personnr",
  rxs = list(
    "antidepressants" = c("N06A"),
    "antipsychotics" = c("N05A"),
    "hormones" = c("G03")
  ))

# Cause of death
fake_cod <- swereg::fake_cod |>
  data.table::copy() |>
  swereg::make_lowercase_names(date_columns = "dodsdat")
swereg::add_cods(skeleton, fake_cod, id_name = "lopnr",
  cods = list(
    "external_death" = c("X60", "X70"),
    "cardiovascular_death" = c("I21", "I22")
  ))

# Derived variables
skeleton[, birth_year := as.numeric(substr(fodelseman, 1, 4))]
skeleton[, age := isoyear - birth_year]
skeleton[, any_mental_health := depression | anxiety | psychosis]
skeleton[, depression_treated := depression & antidepressants]
skeleton[, death_any := external_death | cardiovascular_death]
skeleton[, life_stage := fcase(
  age < 18, "child",
  age >= 18 & age < 65, "adult",
  age >= 65, "elderly",
  default = "unknown"
)]
skeleton[, c("fodelseman", "birth_year") := NULL]

# Filter
skeleton <- skeleton[age >= 0 & age <= 100 & isoyear >= 2015]

cat("Skeleton ready:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
```

## Weekly vs yearly rows

The skeleton contains both weekly and yearly rows. The `is_isoyear`
column distinguishes them:

```{r}
cat("Weekly rows:", sum(!skeleton$is_isoyear), "\n")
cat("Yearly rows:", sum(skeleton$is_isoyear), "\n")
```

Weekly rows carry event-level precision (diagnosis this week, prescription
active this week). Yearly rows carry annually-updated data (income,
education, family type). Most analyses need the data collapsed to one
granularity.

## Collapsing weekly data to person-years

The most common aggregation: collapse weekly boolean indicators into
yearly summaries answering "did anything happen this year?"

Use `swereg::max_with_infinite_as_na()` for boolean/logical columns --
it returns `TRUE` if any week was `TRUE`, `FALSE` if all were `FALSE`,
and `NA` if all were `NA`:

```{r}
person_years <- skeleton[, .(
  age            = swereg::first_non_na(age),
  life_stage     = swereg::first_non_na(life_stage),

  # "Did this happen at any point during the year?"
  depression     = swereg::max_with_infinite_as_na(depression),
  anxiety        = swereg::max_with_infinite_as_na(anxiety),
  psychosis      = swereg::max_with_infinite_as_na(psychosis),
  any_mental_health   = swereg::max_with_infinite_as_na(any_mental_health),
  gender_dysphoria    = swereg::max_with_infinite_as_na(gender_dysphoria),
  antidepressants     = swereg::max_with_infinite_as_na(antidepressants),
  antipsychotics      = swereg::max_with_infinite_as_na(antipsychotics),
  hormones            = swereg::max_with_infinite_as_na(hormones),
  depression_treated  = swereg::max_with_infinite_as_na(depression_treated),
  death_any           = swereg::max_with_infinite_as_na(death_any)
), by = .(id, isoyear)]

cat("Person-year dataset:", nrow(person_years), "rows\n")
head(person_years)
```

## Example analyses on the person-year dataset

### Prevalence by year

```{r}
prevalence <- person_years[, .(
  n = .N,
  depression_prev = mean(depression, na.rm = TRUE),
  anxiety_prev    = mean(anxiety, na.rm = TRUE),
  treatment_rate  = ifelse(
    sum(depression, na.rm = TRUE) > 0,
    mean(depression_treated[depression == TRUE], na.rm = TRUE),
    NA_real_
  )
), by = .(isoyear)]

print(prevalence[order(isoyear)])
```

### Prevalence by life stage

```{r}
by_stage <- person_years[, .(
  n = .N,
  depression_prev     = mean(depression, na.rm = TRUE),
  antidepressant_use  = mean(antidepressants, na.rm = TRUE),
  mean_age            = mean(age, na.rm = TRUE)
), by = .(life_stage)]

print(by_stage)
```

### Treatment patterns among those with any mental health condition

```{r}
treatment <- person_years[any_mental_health == TRUE, .(
  antidepressant_use = mean(antidepressants, na.rm = TRUE),
  antipsychotic_use  = mean(antipsychotics, na.rm = TRUE),
  hormone_use        = mean(hormones, na.rm = TRUE),
  mean_age           = mean(age, na.rm = TRUE),
  n                  = .N
), by = .(isoyear)]

print(treatment[order(isoyear)])
```

## Working directly with weekly data

Some analyses need weekly resolution -- for example, time-to-event
models or exposure-lag analyses. Use the weekly rows directly:

```{r}
weekly <- skeleton[is_isoyear == FALSE]
cat("Weekly dataset:", nrow(weekly), "rows\n")

# Weeks with active depression treatment
cat("Person-weeks on antidepressants:",
    sum(weekly$antidepressants, na.rm = TRUE), "\n")
```

## Creating a person-level baseline table

Collapse to one row per person for baseline characteristics:

```{r}
baseline <- person_years[, .(
  first_year     = min(isoyear),
  last_year      = max(isoyear),
  follow_up_years = .N,
  ever_depression = any(depression, na.rm = TRUE),
  ever_anxiety    = any(anxiety, na.rm = TRUE),
  ever_gd         = any(gender_dysphoria, na.rm = TRUE),
  died            = any(death_any, na.rm = TRUE)
), by = .(id)]

cat("Baseline table:", nrow(baseline), "individuals\n")
print(head(baseline))
```

## Summary

The two-step manual workflow:

1. **Create the skeleton** (`vignette("skeleton-create")`): build the
   time grid, integrate registry data, derive variables
2. **Analyse the skeleton** (this vignette): collapse to the right
   granularity and run queries

For production-scale pipelines with incremental rebuilds, see
`vignette("skeleton-pipeline")`.
