---
title: "The three-phase skeleton pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The three-phase skeleton pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The three-phase skeleton pipeline

This vignette describes how `swereg` builds and incrementally
rebuilds per-batch skeleton files in production. If you just want to
create a small skeleton and attach some diagnoses by hand, read
`vignette("skeleton1-create")` first. This vignette is for people
running full-registry pipelines where total build time is measured in
days and every iteration matters.

## The problem

A production swereg pipeline on Swedish registry data typically has
two dominant costs:

1. **Rawbatch creation** (14h on real data): split the raw registry
   files into per-person-batch subsets, save as `.qs2`.
2. **Skeleton processing** (48h on real data): for each batch, build a
   person-week time grid, apply censoring, add demographics from LISA,
   classify MHT exposure from LMED, and match ICD-10 / ATC / operation
   codes across the subset.

That is roughly 62 hours per full run. Before the three-phase
migration, editing one ICD-10 code in `code_definitions.R` meant a
full 62h rebuild, because the old pipeline ran the whole
`skeleton_create()` function end-to-end per batch and serialized the
bare `data.table`.

The goal of the three-phase pipeline is to make that 62h cost
*proportional to the size of the change*: editing one code pattern
should cost a few minutes, not days.

## Three phases, three cadences of change

The pipeline splits skeleton construction into three phases, each
with its own invalidation strategy. Per batch,
`RegistryStudy$process_skeletons()` runs the phases in order, but
only re-runs the parts that actually changed.

### Phase 1 -- framework

A single user-supplied function, registered via
`$register_framework(fn)`, with signature
`function(batch_data, config)`. It returns a fresh `data.table`
containing the base time grid plus structural censoring (immigration,
emigration, death). Nothing downstream can exist without it.

**Invalidation:** full rebuild whenever
`digest::digest(list(body(fn), formals(fn)), "xxhash64")` changes.
Framework edits are expected to be rare (you don't change the time
grid on a whim), so a conservative "always rebuild on hash change" is
fine. When phase 1 re-runs, phase 2 and phase 3 state are discarded
and re-applied from scratch on the fresh base.

### Phase 3 -- randvars

An ordered named list of user-supplied functions, each registered via
`$register_randvars(name, fn)`. Each has signature
`function(skeleton, batch_data, config)` and mutates the skeleton in
place (or, less commonly, returns a new `data.table` if it filters
rows).

Registration order is execution order. Editing one step should replay
that step and everything downstream of it but leave upstream steps
untouched.

**Invalidation:** *divergence-point rewind-and-replay*. swereg scans
the stored `(name, hash)` sequence against the current one, finds the
first position where either the name or the hash differs, drops the
stored `added_columns` of every step from that point forward, and
then replays current steps from that point. Add, remove, edit, and
reorder are all handled uniformly because any of those operations
changes either the name sequence or the hash sequence.

For the MHT project, phase 3 looks like:

```r
study$register_randvars("lisa_demographics",  skeleton_randvars_lisa)
study$register_randvars("lmed_exposure",      skeleton_randvars_lmed)
study$register_randvars("pre_2006_exclusion", skeleton_randvars_exclusion)
```

`pre_2006_exclusion` reads `approach1` from `lmed_exposure`, so when
you edit `lmed_exposure`'s body, the divergence point lands at step 2
and both step 2 and step 3 replay. That dependency is implicit in the
order -- there is no explicit dep graph.

### Phase 2 -- code registry

A list of code registrations, each added via `$register_codes()`
(primary) or `$register_derived_codes()` (derived). Each entry
contributes one or more columns to the skeleton; the shape of those
columns is predicted from metadata, not observed from running `fn`.

**Invalidation:** per-entry fingerprint diff.
`Skeleton$sync_with_registry()` computes
`to_drop = stored_fingerprints - current_fingerprints` and
`to_add = current_fingerprints - stored_fingerprints`, drops stale
entries' columns, then re-applies new entries in registration order.
When you change one code in one entry, every other entry's
fingerprint is unchanged, so only that one entry re-runs.

Phase 2 runs AFTER phase 3 in the current orchestration. Phase-3
steps CANNOT depend on phase-2 columns. If you find yourself wanting
that, push the logic into phase 3 (read raw data yourself) or
restructure.

## The `Skeleton` R6 class

Each per-batch file on disk is a serialized `Skeleton` object, not a
bare `data.table`. A `Skeleton` carries its own provenance:

```{r, eval = FALSE}
sk <- study$load_skeleton(batch_number = 1L)
sk
#> <Skeleton batch 1>
#>   rows:             8,234,112
#>   cols:             287
#>   framework_hash:   abcd1234efgh
#>   randvars steps:   3
#>   applied codes:    127
#>   pipeline_hash:    4d8f99af7b2c

sk$data                          # the underlying data.table
sk$framework_fn_hash             # phase-1 hash
names(sk$randvars_state)         # applied phase-3 step names in order
length(sk$applied_registry)      # phase-2 entries currently materialized
sk$pipeline_hash()               # rolled-up provenance scalar
```

### Why R6 and not a bare `data.table`?

Because the provenance IS the interesting state. Without it, you
can't answer "is this skeleton still valid for the current pipeline?"
without either rebuilding from scratch or trusting a separate
sidecar file. With it, every batch file can answer the question
locally via `sk$pipeline_hash() == study$pipeline_hash()`.

### The qs2 over-allocation gotcha

`data.table` maintains over-allocated pointer slots ("truelength")
so that in-place `:=` doesn't have to reallocate. qs2 serialization
drops those slots, so after `qs_read()` the table has `truelength == 0`
and the next `:=` inside a function triggers a silent reallocation
that breaks the caller's reference.

`RegistryStudy$load_skeleton()` fixes this by calling
`data.table::setalloccol()` on the freshly-loaded table:

```r
obj$data <- data.table::setalloccol(
  obj$data, n = getOption("datatable.alloccol", 4096L)
)
```

The reassignment is mandatory -- `setalloccol()` returns a new
pointer. `data.table::copy()` would also work but doubles memory
use, which matters when skeletons are multi-GB per batch.

### Legacy bare-data.table auto-wrap

If `load_skeleton()` encounters a legacy pre-migration bare
`data.table` file, it wraps it in a new `Skeleton` with empty
provenance fields. The next pipeline run then sees
`framework_fn_hash == NULL` and triggers a full rebuild for that
batch -- the safe fallback for unknown files.

## The code registry: primary and derived entries

### Primary entries via `register_codes()`

A primary entry ties a code list to a `fn` (e.g. `add_diagnoses`,
`add_cods`, `add_rx`) and a set of raw-data groups to apply it to.
Optional `combine_as` runs `fn` once more on the rbind of all groups
to produce a combined column.

```r
study$register_codes(
  codes      = list(f20 = c("F20"), vte = c("I26", "I80")),
  fn         = swereg::add_diagnoses,
  groups     = list(ov = "outpatient", sv = "inpatient"),
  combine_as = "os"
)
```

For each `(group_prefix, code_name)` pair, a column is written:
`ov_f20`, `sv_f20`, `ov_vte`, `sv_vte`. Because `combine_as = "os"`
is set, two additional columns are produced:
`os_f20 = ov_f20 | sv_f20` and `os_vte = ov_vte | sv_vte`
(computed by re-running `add_diagnoses` on the rbind of outpatient
and inpatient rawbatch data, not by ORing the columns).

### The DORS gotcha and `register_derived_codes()`

The pre-derived-codes approach tried to be clever: register ICD-10
codes once with `groups = list(ov, sv, dors)` and
`combine_as = "osd"`, hoping to get one column covering all three
sources. This failed silently because `add_diagnoses` searches for
`hdia`/`dia*`/`ekod*`/`icd*` columns -- none of which exist in the
cause-of-death registry. DORS uses `ulorsak` (underlying cause) and
`morsak*` (contributing causes). So `dors_f20` was always `FALSE`
and `osd_f20 = ov_f20 | sv_f20 | FALSE` was effectively just hospital
data. **Deaths never contributed to outcomes.**

The fix needs two different functions -- `add_diagnoses` for hospital,
`add_cods` for DORS -- sharing one logical combined column. `combine_as`
can't express that: it reruns the SAME `fn` on rbind data. So we need
a way to OR together columns that were produced by different
registrations.

Enter `register_derived_codes()`:

```r
# 1. Hospital (OV + SV) -> ov_*, sv_*, os_*
study$register_codes(
  codes      = ICD10_CODES,
  fn         = swereg::add_diagnoses,
  groups     = list(ov = "outpatient", sv = "inpatient"),
  combine_as = "os"
)

# 2. DORS underlying cause -> dorsu_*
study$register_codes(
  codes   = ICD10_CODES,
  fn      = swereg::add_cods,
  fn_args = list(cod_type = "underlying"),
  groups  = list(dorsu = "dors")
)

# 3. DORS contributing causes -> dorsm_*
study$register_codes(
  codes   = ICD10_CODES,
  fn      = swereg::add_cods,
  fn_args = list(cod_type = "multiple"),
  groups  = list(dorsm = "dors")
)

# 4. Derived: osd_<nm> = os_<nm> | dorsu_<nm> | dorsm_<nm>
study$register_derived_codes(
  codes = ICD10_CODES,
  from  = c("os", "dorsu", "dorsm"),
  as    = "osd"
)
```

A derived entry doesn't read rawbatch data at all. It iterates its
own `codes` list, looks up
`get("<from[1]>_<nm>")`, `get("<from[2]>_<nm>")`, etc. on the
skeleton, ORs them, and writes `<as>_<nm>`. Missing source columns
raise a loud error. Because it runs in registration order during
phase 2 sync, upstream primary entries are guaranteed to have been
applied already.

### Fingerprint cascade for derived entries

Derived entries fold the fingerprints of their upstream primary
entries into their own fingerprint. That means editing ANY upstream
field -- `fn_args`, `groups`, even a single code pattern -- cascades
into derived replay without the user having to touch the derived
entry. If you flip `cod_type = "underlying"` to
`cod_type = "both"` on the DORS primary, the next
`process_skeletons()` run drops the derived entry's columns and
rewrites them from the new dorsu output, all on every batch, in
order, without rebuilding anything else.

Cascade is implemented in `code_registry_fingerprints()` as a
two-pass walk:

1. Compute every primary entry's fingerprint (depends only on
   `codes, label, groups, fn_args, combine_as`, exactly as in the
   pre-derived release).
2. For each derived entry, walk entries registered before it,
   collect the primary fingerprints whose `groups` prefixes or
   `combine_as` appear in the derived's `from`, and hash
   `(kind="derived", codes, from, as, upstream_fps)`.

Primary fingerprint stability is preserved (same payload shape as
before the derived feature landed) so the release doesn't
accidentally invalidate any existing skeletons.

## Registering the pipeline

A complete runner script has the shape:

```r
study <- swereg::RegistryStudy$new(
  data_rawbatch_dir = c("/linux/.../rawbatch/", "C:/.../rawbatch/"),
  data_skeleton_dir = c("/linux/.../skeleton/", "C:/.../skeleton/"),
  data_raw_dir      = c("/linux/.../raw/",      "C:/.../raw/"),
  group_names       = c("lmed", "inpatient", "outpatient",
                        "cancer", "dors", "other")
)

# Phase 1
study$register_framework(skeleton_framework)

# Phase 3 (registration order = execution order)
study$register_randvars("lisa_demographics",  skeleton_randvars_lisa)
study$register_randvars("lmed_exposure",      skeleton_randvars_lmed)
study$register_randvars("pre_2006_exclusion", skeleton_randvars_exclusion)

# Phase 2: primaries first, then derived
study$register_codes(ICD10_CODES, swereg::add_diagnoses,
  groups = list(ov = "outpatient", sv = "inpatient"), combine_as = "os")
study$register_codes(ICD10_CODES, swereg::add_cods,
  fn_args = list(cod_type = "underlying"),
  groups = list(dorsu = "dors"))
study$register_codes(ICD10_CODES, swereg::add_cods,
  fn_args = list(cod_type = "multiple"),
  groups = list(dorsm = "dors"))
study$register_derived_codes(ICD10_CODES,
  from = c("os", "dorsu", "dorsm"), as = "osd")

# ... rawbatch creation (one-time) ...

study$process_skeletons(n_workers = 4L)
```

The critical invariant: the framework, randvars, and code registry
must all be registered on `study` BEFORE `$process_skeletons()` is
called. There is no callback argument on `$process_skeletons()` --
the pipeline is declared on the study itself, not passed in per call.

## The reload-clobber anti-pattern

A subtle failure mode when running the pipeline repeatedly:

```r
# WRONG
study <- swereg::RegistryStudy$new(...)
study$register_framework(current_framework_fn)     # fresh in-memory
study$register_codes(current_code_list, ...)       # fresh in-memory

if (file.exists(study$meta_file)) {
  study <- swereg::qs2_read(study$meta_file)       # CLOBBER
}
```

The reload path replaces the freshly-registered pipeline with the
stale on-disk version, silently reverting any edits you just made.
The fix is to absorb only the persisted *runtime* state (IDs,
batches, groups_saved) into the fresh study without touching its
config:

```r
# CORRECT
if (file.exists(study$meta_file)) {
  study$adopt_runtime_state_from(swereg::qs2_read(study$meta_file))
}
```

## Provenance verification

Each `Skeleton` can compute its own `pipeline_hash()` from its stored
phase provenance. The study exposes the expected value:

```r
sk <- study$load_skeleton(1L)
sk$pipeline_hash() == study$pipeline_hash()
# TRUE iff this batch is fully synced with the current pipeline
```

Two summary helpers:

```r
# Per-batch table with framework hash, randvars count, code count, saved_at
study$skeleton_pipeline_hashes()

# Errors loudly if any persisted skeleton is out of sync
study$assert_skeletons_consistent()
```

Downstream TTE consumers call `$assert_skeletons_consistent()`
automatically near the top of
`tteplan_from_spec_and_registrystudy()`, so you can't accidentally
build a plan against a half-upgraded skeleton set.

## Cross-host pipeline snapshots

If `data_pipeline_snapshot_dir` is configured at construction time,
`$process_skeletons()` writes one TSV row per host to
`{snapshot_dir}/{host_label}.tsv` after each successful run. The
snapshot file is meant to be `git add`-ed and committed, so the
history of who ran what version of the pipeline on which host lives
in `git log`.

Why per-host files and not a single append-only log: the user
typically runs on 3+ hosts in parallel. A shared append-only file
would conflict on every concurrent run. Per-host files never have
the same name, so concurrent runs produce separate commits that
merge without conflict.

A tiny helper at `dev/commit_pipeline_snapshot.sh` in the MHT repo
wraps the `git add` + `git commit` into a one-liner the operator
runs after the generator finishes. swereg never runs git commands
itself -- that would be a side effect the user hasn't explicitly
asked for.

## Expected re-run costs

After the first full build:

| Edit | Cost |
|---|---|
| Nothing | Per-batch no-op: load `Skeleton`, see nothing changed, save. Seconds per batch. |
| One code pattern in one primary entry | Drop that entry's columns on all batches, re-apply it. Roughly `(phase-2 total) / (N primaries)`. |
| One primary's `fn_args` (e.g. `cod_type`) | Same as above, plus the cascade re-applies any derived entries that depend on it. |
| One phase-3 step's body | Drop that step's columns on all batches, replay it and everything downstream of it in the randvars sequence. |
| The framework function's body | Full rebuild from scratch, all phases. Rare. |

For the MHT project, the expected "edit one ICD-10 code" cost drops
from ~62 hours to ~minutes. That's the headline win.

## Summary

- **Per-batch skeleton files are `Skeleton` R6 objects** carrying
  framework hash, ordered randvars state, and the applied code
  registry fingerprint map.
- **Three phases with independent invalidation**: framework (full
  rebuild on hash change), randvars (divergence-point rewind and
  replay), codes (per-entry fingerprint diff).
- **Primary and derived code entries** let you combine outputs from
  different registration functions into one logical column. Derived
  entries cascade upstream fingerprint changes automatically.
- **Provenance verification** via `pipeline_hash()` lets any
  consumer check whether a skeleton matches the current pipeline.
- **`adopt_runtime_state_from()`** is the correct reload pattern --
  never clobber the freshly-registered pipeline by reassigning from
  disk.

For further context:

- `?RegistryStudy` -- full method reference
- `?Skeleton` -- per-batch provenance object
- `vignette("skeleton-concept")` -- why the person-week time grid
- `vignette("rowdep-rowind-concept")` -- variable-type conventions
