---
title: "Basic Longitudinal Skeletons from Swedish Registry Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Longitudinal Skeletons from Swedish Registry Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
```

## Introduction

**swereg** is an R package designed for manipulating and analyzing Swedish healthcare registry data in epidemiological research. It provides a comprehensive framework for creating longitudinal data skeletons and integrating multiple Swedish health registries.

This vignette demonstrates the basic workflow for creating `skeleton1_create` - the foundation of all swereg analyses.

## Core Concept: The Longitudinal Skeleton

Think of the skeleton concept like building a body: you create **'good bones'** (the skeleton structure) and then attach the **'muscles'** (the data) to it.

### The Three-Stage Skeleton Workflow

swereg uses a systematic three-stage approach for registry data analysis:

1. **skeleton1_create**: Raw data attachment stage
   - Create the time-structured framework (the 'bones')
   - Attach raw registry data from multiple sources
   - Focus on data integration, not cleaning

2. **skeleton2_clean**: Data cleaning and derivation stage  
   - Clean and standardize variables
   - Create derived variables and indicators
   - Handle missing data and outliers

3. **skeleton3_analyze**: Analysis preparation stage
   - Minimize to only rows/columns needed for specific analysis
   - Apply inclusion/exclusion criteria
   - Create final analysis-ready dataset

### skeleton1_create: The Foundation

The skeleton1_create stage provides a time-structured framework with:

- **Individual IDs**: Each person in your study
- **Time structure**: Both ISO years and ISO year-weeks 
- **Data integration points**: Framework for merging registry data from NPR, LMED, SCB, cause of death registries

## Step 1: Create the Basic Skeleton

Start by creating a skeleton with your study population and time period:

```{r}
# Load example data
data("fake_person_ids", package = "swereg")

# Create skeleton covering 2015-2020
skeleton <- swereg::create_skeleton(
  ids = fake_person_ids,
  date_min = "2015-01-01",
  date_max = "2020-12-31"
)

# Examine the structure
head(skeleton)
cat("Skeleton dimensions:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
```

The skeleton contains:

- `id`: Individual identifier
- `isoyear`: ISO year (for annual data)
- `isoyearweek`: ISO year-week (for weekly data)
- `is_isoyear`: Boolean indicating annual vs weekly rows

## Step 2: Add Demographic Data (One-time)

Demographics don't change over time, so we add them once per person:

```{r}
# Load and prepare demographic data
data("fake_demographics", package = "swereg")
swereg::make_lowercase_names(fake_demographics)

# Add to skeleton
swereg::add_onetime(skeleton, fake_demographics, id_name = "lopnr")

# Check what was added
new_vars <- setdiff(names(skeleton), c("id", "isoyear", "isoyearweek", "is_isoyear"))
cat("Added demographic variables:", paste(new_vars, collapse = ", "), "\n")
```

## Step 3: Add Annual Data

Some data varies by year (e.g., family status, income):

```{r}
# Load annual family data
data("fake_annual_family", package = "swereg")
swereg::make_lowercase_names(fake_annual_family)

# Add annual data for 2015
swereg::add_annual(skeleton, fake_annual_family, id_name = "lopnr", isoyear = 2015)

cat("Annual data added for 2015\n")
```

## Step 4: Add Diagnosis Data

Hospital diagnoses are the core of most epidemiological studies:

```{r}
# Load and prepare diagnosis data
data("fake_inpatient_diagnoses", package = "swereg")
data("fake_outpatient_diagnoses", package = "swereg")

# CRITICAL: Apply make_lowercase_names FIRST
swereg::make_lowercase_names(fake_inpatient_diagnoses)
swereg::make_lowercase_names(fake_outpatient_diagnoses)

# Combine inpatient and outpatient
diagnoses_combined <- rbindlist(list(
  fake_inpatient_diagnoses,
  fake_outpatient_diagnoses
), use.names = TRUE, fill = TRUE)

# Define diagnosis patterns to search for (^ prefix automatically added)
diagnosis_patterns <- list(
  "depression" = c("F32", "F33"),
  "anxiety" = c("F40", "F41"), 
  "gender_dysphoria" = c("F64"),
  "cardiovascular" = c("I10", "I20", "I21")
)

# Add diagnoses to skeleton
swereg::add_diagnoses(
  skeleton,
  diagnoses_combined,
  id_name = "lopnr",
  diags = diagnosis_patterns
)

# Check results
diag_vars <- names(diagnosis_patterns)
for(var in diag_vars) {
  count <- sum(skeleton[[var]], na.rm = TRUE)
  cat("-", var, ":", count, "positive cases\n")
}
```

## Step 5: Add Prescription Data

Medication data from the Swedish Prescribed Drug Register:

```{r}
# Load prescription data
data("fake_prescriptions", package = "swereg")
swereg::make_lowercase_names(fake_prescriptions)

# Define drug patterns (ATC codes, ^ prefix automatically added)
drug_patterns <- list(
  "antidepressants" = c("N06A"),
  "hormones" = c("G03"),
  "cardiovascular_drugs" = c("C07", "C08", "C09")
)

# Add prescriptions to skeleton
swereg::add_rx(
  skeleton,
  fake_prescriptions,
  id_name = "p444_lopnr_personnr",
  rxs = drug_patterns
)

# Check prescription usage
rx_vars <- names(drug_patterns)
for(var in rx_vars) {
  count <- sum(skeleton[[var]], na.rm = TRUE)
  cat("-", var, ":", count, "prescription periods\n")
}
```

## Step 6: Add Cause of Death Data

For mortality studies:

```{r}
# Load cause of death data
data("fake_cod", package = "swereg")
swereg::make_lowercase_names(fake_cod)

# Define cause of death patterns (^ prefix automatically added)
cod_patterns <- list(
  "cardiovascular_death" = c("I21", "I22"),
  "external_causes" = c("X60", "X70")
)

# Add to skeleton
swereg::add_cods(
  skeleton,
  fake_cod,
  id_name = "lopnr",
  cods = cod_patterns
)

# Check mortality
cod_vars <- names(cod_patterns)
for(var in cod_vars) {
  count <- sum(skeleton[[var]], na.rm = TRUE)
  cat("-", var, ":", count, "deaths\n")
}
```

## Completed skeleton1_create

Your completed **skeleton1_create** now contains all the raw registry data integrated into the time structure:

```{r}
cat("Final skeleton dimensions:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
cat("Variables:", paste(names(skeleton), collapse = ", "), "\n")

# Example: Show data for one person
person_data <- skeleton[id == fake_person_ids[1] & isoyear == 2018]
cat("\nExample data for person", fake_person_ids[1], "in 2018:\n")
print(head(person_data[, .(id, isoyearweek, depression, antidepressants, cardiovascular)]))
```

## Key Principles

1. **Always use `make_lowercase_names()`** after reading registry data
2. **Sequential integration**: Add data types in logical order
3. **Pattern matching**: Use regex patterns for medical codes
4. **Time structure**: Leverage ISO year-weeks for precise temporal analysis

## Next Steps: From skeleton1_create to Analysis

This **skeleton1_create** forms the foundation for the complete workflow:

- **skeleton2_clean**: Clean variables, create derived indicators, handle missing data
- **skeleton3_analyze**: Apply study criteria and create analysis-ready datasets

The raw data is now integrated - next you'll clean it (skeleton2) and then prepare it for analysis (skeleton3).

See the "Advanced Workflow" vignette (`vignette("advanced-workflow")`) for the complete three-stage pipeline including skeleton2_clean and skeleton3_analyze.
