---
title: "Basic Longitudinal Skeletons from Swedish Registry Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Longitudinal Skeletons from Swedish Registry Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
```

## Introduction

**swereg** is an R package designed for manipulating and analyzing Swedish healthcare registry data in epidemiological research. It provides a comprehensive framework for creating longitudinal data skeletons and integrating multiple Swedish health registries.

This vignette demonstrates the basic workflow for creating `skeleton1_create` - the foundation of all swereg analyses.

## Core Concept: The Longitudinal Skeleton

The skeleton is a time-structured framework that provides:

- **Individual IDs**: Each person in your study
- **Time structure**: Both ISO years and ISO year-weeks 
- **Data integration points**: Framework for merging registry data

## Step 1: Create the Basic Skeleton

Start by creating a skeleton with your study population and time period:

```{r}
# Load example data
data("fake_person_ids", package = "swereg")

# Select a subset for demonstration
ids <- fake_person_ids[1:20]

# Create skeleton covering 2015-2020
skeleton <- swereg::create_skeleton(
  ids = ids,
  date_min = "2015-01-01",
  date_max = "2020-12-31"
)

# Examine the structure
head(skeleton)
cat("Skeleton dimensions:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
```

The skeleton contains:
- `id`: Individual identifier
- `isoyear`: ISO year (for annual data)
- `isoyearweek`: ISO year-week (for weekly data)
- `is_isoyear`: Boolean indicating annual vs weekly rows

## Step 2: Add Demographic Data (One-time)

Demographics don't change over time, so we add them once per person:

```{r}
# Load and prepare demographic data
data("fake_demographics", package = "swereg")
swereg::make_lowercase_names(fake_demographics)

# Filter to our study population
demographics_subset <- fake_demographics[lopnr %in% ids]

# Add to skeleton
swereg::add_onetime(skeleton, demographics_subset, id_name = "lopnr")

# Check what was added
new_vars <- setdiff(names(skeleton), c("id", "isoyear", "isoyearweek", "is_isoyear"))
cat("Added demographic variables:", paste(new_vars, collapse = ", "), "\n")
```

## Step 3: Add Annual Data

Some data varies by year (e.g., family status, income):

```{r}
# Load annual family data
data("fake_annual_family", package = "swereg")
swereg::make_lowercase_names(fake_annual_family)

annual_subset <- fake_annual_family[lopnr %in% ids]

# Add annual data for 2015
swereg::add_annual(skeleton, annual_subset, id_name = "lopnr", isoyear = 2015)

cat("Annual data added for 2015\n")
```

## Step 4: Add Diagnosis Data

Hospital diagnoses are the core of most epidemiological studies:

```{r}
# Load and prepare diagnosis data
data("fake_inpatient_diagnoses", package = "swereg")
data("fake_outpatient_diagnoses", package = "swereg")

# CRITICAL: Apply make_lowercase_names FIRST
swereg::make_lowercase_names(fake_inpatient_diagnoses)
swereg::make_lowercase_names(fake_outpatient_diagnoses)

# Combine inpatient and outpatient
diagnoses_combined <- rbindlist(list(
  fake_inpatient_diagnoses[lopnr %in% ids],
  fake_outpatient_diagnoses[lopnr %in% ids]
), use.names = TRUE, fill = TRUE)

# Define diagnosis patterns to search for (^ prefix automatically added)
diagnosis_patterns <- list(
  "depression" = c("F32", "F33"),
  "anxiety" = c("F40", "F41"), 
  "gender_dysphoria" = c("F64"),
  "cardiovascular" = c("I10", "I20", "I21")
)

# Add diagnoses to skeleton
swereg::add_diagnoses(
  skeleton,
  diagnoses_combined,
  id_name = "lopnr",
  diags = diagnosis_patterns
)

# Check results
diag_vars <- names(diagnosis_patterns)
for(var in diag_vars) {
  count <- sum(skeleton[[var]], na.rm = TRUE)
  cat("-", var, ":", count, "positive cases\n")
}
```

## Step 5: Add Prescription Data

Medication data from the Swedish Prescribed Drug Register:

```{r}
# Load prescription data
data("fake_prescriptions", package = "swereg")
swereg::make_lowercase_names(fake_prescriptions)

prescriptions_subset <- fake_prescriptions[p444_lopnr_personnr %in% ids]

# Define drug patterns (ATC codes, ^ prefix automatically added)
drug_patterns <- list(
  "antidepressants" = c("N06A"),
  "hormones" = c("G03"),
  "cardiovascular_drugs" = c("C07", "C08", "C09")
)

# Add prescriptions to skeleton
swereg::add_rx(
  skeleton,
  prescriptions_subset,
  id_name = "p444_lopnr_personnr",
  rxs = drug_patterns
)

# Check prescription usage
rx_vars <- names(drug_patterns)
for(var in rx_vars) {
  count <- sum(skeleton[[var]], na.rm = TRUE)
  cat("-", var, ":", count, "prescription periods\n")
}
```

## Step 6: Add Cause of Death Data

For mortality studies:

```{r}
# Load cause of death data
data("fake_cod", package = "swereg")
swereg::make_lowercase_names(fake_cod)

cod_subset <- fake_cod[lopnr %in% ids]

# Define cause of death patterns (^ prefix automatically added)
cod_patterns <- list(
  "cardiovascular_death" = c("I21", "I22"),
  "external_causes" = c("X60", "X70")
)

# Add to skeleton
swereg::add_cods(
  skeleton,
  cod_subset,
  id_name = "lopnr",
  cods = cod_patterns
)

# Check mortality
cod_vars <- names(cod_patterns)
for(var in cod_vars) {
  count <- sum(skeleton[[var]], na.rm = TRUE)
  cat("-", var, ":", count, "deaths\n")
}
```

## Final skeleton1_create

Your completed skeleton1_create now contains:

```{r}
cat("Final skeleton dimensions:", nrow(skeleton), "rows,", ncol(skeleton), "columns\n")
cat("Variables:", paste(names(skeleton), collapse = ", "), "\n")

# Example: Show data for one person
person_data <- skeleton[id == ids[1] & isoyear == 2018]
cat("\nExample data for person", ids[1], "in 2018:\n")
print(head(person_data[, .(id, isoyearweek, depression, antidepressants, cardiovascular)]))
```

## Key Principles

1. **Always use `make_lowercase_names()`** after reading registry data
2. **Sequential integration**: Add data types in logical order
3. **Pattern matching**: Use regex patterns for medical codes
4. **Time structure**: Leverage ISO year-weeks for precise temporal analysis

## Next Steps

This basic skeleton1_create forms the foundation for:
- **skeleton2_clean**: Creating derived variables and cleaning data
- **skeleton3_analyze**: Extracting analysis-ready datasets

See the "Advanced Workflow" vignette for the complete pipeline.
