---
title: "The specification schema"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The specification schema}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Why the schema exists

`tteplan_read_spec()` refuses every key the schema does not name. Before that,
a key swereg did not read was ignored in silence.

A collaborator wrote a cohort restriction under
`inclusion_criteria$additional_inclusion`. swereg read `inclusion_criteria` for
`isoyears` and nothing else. The restriction never reached the eligibility
filter. The study population stayed unrestricted, and no message said so. The
specification looked exactly like one that worked.

One table names every legal key path. It is `.TTE_SPEC_SCHEMA`, in
`R/tteplan_spec_schema.R`, and nothing else in swereg holds that list. Every
table below is generated from that object, so no second copy of the list can
drift away from it.

# How a key path is written

A path is written the way the walker over the parsed YAML produces it.

- The root is `$`.
- A mapping key appends `/<key>`.
- A sequence index becomes `[]`.

A treatment block therefore has the path `$/enrollments[]/treatment`. Every
element of a sequence shares one path, so the schema states one rule for all of
them.

The path is what separates two keys that share a name.
`$/enrollments[]/additional_inclusion` is accepted.
`$/inclusion_criteria/additional_inclusion` is refused. A table keyed by the
key name alone cannot state that.

Each context declares its own children. There is no generic rule for an
`implementation` block. A treatment implementation, an outcome implementation
and a confounder implementation accept different keys.

# The three key classes

Each key path carries one class.

```{r schema}
schema <- swereg:::.TTE_SPEC_SCHEMA

rows <- do.call(rbind, lapply(names(schema), function(context) {
  node <- schema[[context]]
  keys <- list(
    consumed = node[["consumed"]],
    metadata = node[["metadata"]],
    legacy = names(node[["legacy"]])
  )
  data.frame(
    context = context,
    key = unlist(keys, use.names = FALSE),
    class = rep(names(keys), lengths(keys)),
    stringsAsFactors = FALSE
  )
}))
rows <- rows[order(rows$context, rows$key), ]

# Wrap a path in backticks. A bare `$` opens inline maths in a pipe table, and
# pandoc then merges the cells it spans.
as_code <- function(x) paste0("`", x, "`")

knitr::kable(
  as.data.frame(table(rows$class), stringsAsFactors = FALSE),
  col.names = c("class", "key paths")
)
```

| class | what swereg does with the key | a specification that carries it |
|:---|:---|:---|
| `consumed` | swereg reads it. | accepted |
| `metadata` | swereg accepts it and never reads it. | accepted |
| `legacy` | swereg refuses it. The message names the replacement. | refused |

## Why two classes are not enough

A two-class table holds "swereg reads it" and "swereg refuses it". Under that
table, every key swereg does not read is refused. These key paths are then
refused, and every one of them is correct today.

```{r metadata}
meta <- rows[rows$class == "metadata", ]
knitr::kable(
  data.frame(context = as_code(meta$context), key = as_code(meta$key)),
  row.names = FALSE
)
```

Each one records a decision for a human reader. `$/open_questions[]/resolution`
records how the study team settled an open question.
`$/standing_methods/admin_censoring` records the administrative censoring rule.
`R/` names both only in the schema, and reads neither value. The `metadata`
class is what lets the gate refuse an unknown key without refusing these.

# Every declared key path

The schema declares `r nrow(rows)` key paths across `r length(schema)` mapping
contexts.

```{r all-paths}
knitr::kable(
  data.frame(
    context = as_code(rows$context),
    key = as_code(rows$key),
    class = rows$class
  ),
  row.names = FALSE
)
```

# The global inclusion container

`inclusion_criteria` is a fixed container. It holds `isoyears` and `criteria`,
and nothing else.

```yaml
inclusion_criteria:
  isoyears: [2010, 2023]
  criteria:
    - name: "Recorded gender dysphoria diagnosis"
      rationale: "The study population is people with the diagnosis"
      type: has_event
      implementation:
        source_variable: osd_f64
        window: lifetime_before_baseline
```

Each entry MUST declare `type: has_event`. `tteplan_read_spec()` refuses any
other type. A criterion swereg reads and ignores never restricts the study
population, and it looks exactly like one that does.

`tteplan_apply_exclusions()` adds one `eligible_has_<variable>_<window>` column
for each criterion. It then combines every eligibility column into `eligible`.

## How it differs from a per-enrollment `additional_inclusion`

| | `inclusion_criteria$criteria` | `enrollments[]$additional_inclusion` |
|:---|:---|:---|
| scope | every enrollment | the one enrollment that declares it |
| path | `$/inclusion_criteria/criteria[]` | `$/enrollments[]/additional_inclusion[]` |
| types accepted | `has_event` | `has_event` and `age_range` |

A `has_event` entry generates the same column name in both places. Two entries
that generate one column name are one criterion written twice, so
`tteplan_read_spec()` refuses the pair. That holds when both entries are
global, and when one is global and one is per-enrollment.

`inclusion_criteria$additional_inclusion` is refused. That name belongs to an
enrollment. Copied to the container, it named a path swereg never read.

# What you see when a key is refused

`tteplan_read_spec()` reports every refused key in one message, and reads no
data first. A legacy key carries its migration message.

```{r legacy}
spec_path <- file.path(tempdir(), "spec_v001.yaml")
writeLines(
  c(
    "inclusion_criteria:",
    "  isoyears: [2010, 2023]",
    "  additional_inclusion:",
    "    - name: \"Recorded gender dysphoria diagnosis\"",
    "      type: has_event",
    "      implementation:",
    "        source_variable: osd_f64",
    "        window: lifetime_before_baseline"
  ),
  spec_path
)
msg <- tryCatch(swereg::tteplan_read_spec(spec_path), error = conditionMessage)
cat(sub(tempdir(), "<tempdir>", msg, fixed = TRUE))
```

The report names one key, not six. The walk stops at a refused key, because the
schema declares no context below it. Move the entries into `criteria`, and
swereg reads each one.

A key the schema does not name anywhere carries the list of keys its context
accepts.

```{r unknown}
writeLines(
  c("inclusion_criteria:", "  isoyear: 2010"),
  spec_path
)
msg <- tryCatch(swereg::tteplan_read_spec(spec_path), error = conditionMessage)
cat(sub(tempdir(), "<tempdir>", msg, fixed = TRUE))
```

# Where the gate runs

`tteplan_read_spec()` calls the gate on the specification as written, before
any normalisation. swereg writes derived keys back into the specification as it
reads it: `window_weeks`, `source_variable_combined`, `variable_combined`, and
the two-field `observed_var`. The schema names none of them, because the schema
describes the input.

`tteplan_read_spec()` holds the one call to `yaml::yaml.load()` in `R/`. Two
functions call it: `tteplan_from_spec()`, and the plan's own `$reload_spec()`
method. No path into a specification skips the gate.

# See also

- `vignette("tte-workflow", package = "swereg")` for the whole specification,
  and for the pipeline that reads it.
- `?tteplan_read_spec` for the observation contract and the window conversion
  rules.
