---
title: "TTE methods: manuscript and statistical analysis plan text"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TTE methods: manuscript and statistical analysis plan text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(data.table)
ev <- readRDS("tte-validation-evidence.rds")
ci3 <- function(est, lo, hi) sprintf("%.3f [%.3f, %.3f]", est, lo, hi)
ci2 <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)
sg3 <- function(x) sprintf("%+.3f", x)
```

This vignette provides drop-in text describing the target trial emulation
(TTE) methodology implemented in the `swereg-TTE` family of functions
(`TTEEnrollment`, `TTEPlan`, and friends), for **both** the intention-to-treat
(ITT) and the per-protocol (PP) estimand. It has four sections:

- **Manuscript methods** (Section 1) — short, prose-only, suitable for the
  main body of a journal article. Copy, paste, and replace the `treatment` /
  `outcome` / `confounder` placeholders.
- **Statistical analysis plan** (Section 2) — detailed and
  implementation-agnostic: formulas, per-step model specifications,
  conventions, identifying assumptions, and known limitations, written so a
  statistician who has never seen this package could reimplement the
  estimator. Suitable for a pre-registered SAP, a protocol appendix, or a
  methods supplement.
- **Validation evidence** (Section 3) — the numbers. Every table is rendered
  from a results artifact produced by rerunning the full validation battery
  (simulations with known planted truth, cross-checked against the
  `TrialEmulation` package), not from prose claims.
- **Implementation mapping** (Section 4) — the code behind the SAP: which
  function, argument, option, and test file realises each SAP step, plus
  provenance notes on estimator changes.

Every formula and convention in Section 2 describes what the code actually
computes; where the implementation deviates from a canonical reference
construction, the deviation and its rationale are stated explicitly. The
implementation follows the sequential-trial-emulation literature (Hernán and
Robins 2008; Danaei et al. 2013; Hernán and Robins 2016; Caniglia et al.
2023; Cashin et al. 2025).

---

# 1. Manuscript methods

We applied target trial emulation, a framework for analysing observational
data under explicit protocols that mirror a hypothetical randomized trial, to
estimate the effect of `treatment` on `outcome` in the Swedish national health
registries (Hernán and Robins 2008, 2016; Cashin et al. 2025).

## Sequential trials design

Because eligible individuals can initiate treatment at many different calendar
times, we emulated a sequence of target trials rather than a single trial
(Hernán et al. 2008; Danaei et al. 2013; Caniglia et al. 2023). A new trial
opens every `period` weeks of calendar time. At each trial's baseline, all
eligibility criteria are re-evaluated; eligible individuals enter as
initiators (treatment begins in that trial's baseline period) or as
non-initiators (eligible and untreated). A new-user (washout) criterion
requires no use of the study treatment within a pre-specified washout window
before baseline (a fixed window, e.g. two years as in Danaei et al. 2013, or
the entire observable history for a never-user design), so each person
initiates in at most one trial while contributing eligible person-time as a
non-initiator to earlier trials. Anchoring time zero at eligibility and assignment — rather
than at eventual exposure — prevents immortal time bias (Hernán and Robins
2016). To bound computation, `k` non-initiators were sampled per initiator
within each trial; confounding adjustment is by weighting (below), not by
matching on covariates.

## Estimands

We report two estimands (Danaei et al. 2013). The observational analogue of
the intention-to-treat effect compares initiators with non-initiators as
classified at each trial's baseline, ignoring subsequent changes in
treatment. The per-protocol effect is the effect of sustained treatment
versus sustained non-treatment; for this estimand, follow-up is artificially
censored when a participant's treatment status first deviates from the
baseline-assigned strategy. Both are reported as marginal incidence rate
ratios (IRRs), with weighted event counts and rates per 100,000 person-years
by arm.

## Confounding and censoring adjustment

Baseline treatment assignment is not random: we adjusted for measured baseline
confounders (`confounders`) by stabilised inverse probability of treatment
weighting, estimated from a logistic model at each trial's baseline (Hernán
and Robins 2008). For the per-protocol estimand, artificial censoring at
protocol deviation is informative whenever time-varying factors predict both
adherence and the outcome; we therefore additionally applied stabilised
inverse probability of censoring weights from discrete-time censoring models
fit separately by assigned arm, with a smooth function of follow-up time and
the most recently updated covariate values (Hernán and Robins 2008; Danaei et
al. 2013). Weights were truncated at the 1st and 99th percentiles to limit
the influence of extreme values (Danaei et al. 2013).

## Outcome model and inference

We fit a weighted quasi-Poisson marginal structural model of the event
indicator on assigned baseline treatment with log person-time as offset,
including natural splines of follow-up time and of the trial (calendar)
index; the exponentiated treatment coefficient estimates the marginal IRR
pooled across sequential trials. Because individuals contribute repeated
observations within and across trials, confidence intervals use
cluster-robust (sandwich) standard errors clustered on the person (Hernán
and Robins 2008; Danaei et al. 2013). Effect heterogeneity across calendar
time and pre-specified subgroups was assessed by Wald tests of the
corresponding interaction terms.

## Software

Analyses used R with the `swereg` package, which implements the sequential
enrollment, weighting, and estimation pipeline described above;
censoring-weight models were fit with `mgcv`, and the final weighted
regression with cluster-robust variance with `survey::svyglm()`. The
implementation is validated against simulated data with known true effects
and against the `TrialEmulation` package (Su et al. 2024); the validation
suite runs in continuous integration.

---

# 2. Statistical analysis plan

This section specifies the estimators exactly as implemented, in pipeline
order, in implementation-agnostic terms (the mapping from each step to the
software that executes it is in Section 4). Notation: individuals
$i = 1, \ldots, N$; sequential trials indexed by their calendar baseline band
$m$; follow-up bands within a trial $j = 0, 1, \ldots, K$, each of fixed
width $w$ weeks (four by default). Let $A_{i,m,j}$ indicate being on the
protocol-defined treatment in band $(m,j)$, with $A_{i,m,0}$ the assigned
baseline arm; $L_{i,m,j}$ the confounder vector as most recently updated at
band $(m,j)$, with $L_{i,m,0}$ its baseline value; $Y_{i,m,j}$ the outcome
indicator; and $C_{i,m,j}$ the indicator of artificial censoring (protocol
deviation or loss to follow-up) in band $(m,j)$.

## 2.1 Sequential enrollment, new-user requirement, and matching

Calendar time is partitioned into consecutive bands of width $w$; each band
opens one trial. Within a band, a person's arm is *intervention* if they are
on the intervention treatment in **any** eligible week of that band, and
*comparator* if on the comparator treatment throughout; person-bands with
treatment status outside the two protocol arms are ineligible for that band.
The band width is a bias–feasibility tradeoff: coarser bands admit residual
within-band immortal time (Caniglia et al. 2023), which shrinks as $w$
decreases.

Eligibility (inclusion windows, exclusion criteria with lifetime or
fixed-width look-back windows) is re-evaluated at every band. The design does
not impose a new-user rule automatically: the incident-user design is
produced by a protocol-specified washout exclusion on the treatment history —
either a finite look-back window (e.g. 104 weeks, the Danaei et al. 2013
convention) or the entire observable history, for a never-user design. A
lifetime washout makes each person eligible to initiate in at most one band
and removes them from later trials; a finite washout additionally allows
re-qualification after sufficient time off treatment. A protocol without any
washout exclusion enrols prevalent users as "initiators" at every band and
re-enrols discontinuers as comparators — a prevalent-user design that is
almost never the intended estimand (the software warns when a specification
omits the washout; Section 4.1).

Within each band, all intervention person-trials are enrolled and comparators
are randomly downsampled at a fixed matching ratio per initiator (2:1 by
default, with a pre-specified seed). This sampling bounds computation and is
**not** covariate matching; all confounding adjustment is deferred to the
weights (2.4). Each enrolled person-trial is expanded to $K$ follow-up bands;
within each band, confounders take their first-week value, outcomes are the
within-band maximum, and person-time is the number of observed source weeks
in the band (so partially observed bands contribute their true person-time).

## 2.2 Estimands

Both estimands are marginal incidence rate ratios, standardised over the
enrolled trials' baseline covariate distribution through the weights:

- **ITT analogue**: the contrast of initiating versus not initiating at
  baseline, ignoring subsequent switching. Identified under baseline
  exchangeability given $L_{m,0}$ (plus assumptions in 2.9); estimated with
  the treatment weight alone.
- **Per-protocol**: the contrast of sustained treatment versus sustained
  non-treatment. Follow-up is censored at the first deviation from the
  assigned strategy; identified under the additional assumption that the
  censoring model captures all joint determinants of deviation and outcome;
  estimated with the product of treatment and censoring weights.

**Interpretation under non-proportional hazards.** The reported IRR is the
coefficient of a proportional-rates working model. When the true marginal
rate ratio varies over follow-up (e.g., depletion of susceptibles, or effects
that accumulate over time), the single IRR is a person-time-weighted average
of the time-varying rate ratio — a well-defined summary, but one that can
differ from, say, the ratio of cumulative incidences over the full horizon.
Simulations with strongly time-varying effects show swereg and
`TrialEmulation` produce the same weighted-average summary (Section 3.2,
harmful-effect cell). Where time-varying effects are of scientific interest,
report follow-up-specific estimates (e.g., by follow-up horizon) rather than
the pooled IRR alone.

## 2.3 Follow-up construction and censoring events

For each person-trial, follow-up stops at the earliest of: (1) the first
outcome event; (2) the first protocol deviation (PP only; the ITT panel never
censors at switching); (3) the person's end of observed data, when that occurs
before any planned stop; (4) the pre-specified administrative end of study;
and (5) the pre-specified analysis horizon.

Deviation is the first band in which the observed treatment status differs
from the assigned arm (initiators off treatment; comparators on treatment);
a band with missing on-treatment status counts as deviation. There is no
grace period: deviation censors at the first mismatched band (a grace-period
design would require cloning, which this pipeline does not implement; Hernán
and Robins 2016).

**Event-priority convention.** If the first event and the first deviation
fall in the same band, the person-trial exits through the *event*: the
outcome is measured over the interval before within-interval censoring is
applied, so the band counts as an event, not a censoring, in both the
censoring model and the analysis data. (The alternative — treating collision
bands as censorings — discards real events and undercounts the per-protocol
outcome in switching-heavy data; see the provenance note in Section 4.2.)

Rows at and before the stop band are retained; censoring-event rows (their
band person-time included) are removed from the analysis data after the
censoring weights are estimated (2.5), so the analysis panel contains only
protocol-consistent, at-risk person-time.

## 2.4 Baseline treatment weights (IPW)

On the baseline row of each person-trial, a logistic regression of assignment
on the baseline confounders (main effects) is fit:

$$
\mathrm{logit}\,\mathrm{Pr}(A_{m,0} = 1 \mid L_{m,0})
  = \gamma_0 + \gamma^\top L_{m,0},
$$

and the stabilised weight uses the **marginal** initiation fraction
$\bar{p} = \widehat{\mathrm{Pr}}(A_{m,0}=1)$ as numerator:

$$
SW^A =
  A_{m,0}\,\frac{\bar{p}}{\widehat{ps}} +
  (1 - A_{m,0})\,\frac{1-\bar{p}}{1-\widehat{ps}},
\qquad \widehat{ps} = \widehat{\mathrm{Pr}}(A_{m,0}=1 \mid L_{m,0}).
$$

The weight is constant across a person-trial's follow-up rows. Missing
baseline confounders are singly imputed by hot-deck sampling from observed
values (fixed seed) before the model is fit; imputation uncertainty is not
propagated (2.8). The propensity model is main-effects only: if strong
non-linearity or interactions are suspected, they must be encoded as derived
confounder variables in the protocol specification.

## 2.5 Per-protocol censoring weights (IPCW)

Censoring ($C_{m,j} = 1$: deviation or loss in band $j$) is modelled on the
panel before censoring-event rows are dropped. The default censoring model,
fit **separately by assigned arm** $a$, is a discrete-time logistic
generalized additive model:

$$
\mathrm{logit}\,\mathrm{Pr}(C_{m,j} = 1 \mid \text{at risk},\ A_{m,0}=a)
  = s_a(j) + s_a(m) + \alpha_a^\top L_{m,j},
$$

with penalised-spline smooth functions $s_a(\cdot)$ of follow-up band and of
the trial index (a linear trial term when there are few bands; a fully
linear-in-time specification is available as a pre-specified sensitivity
option). The confounder columns carry their per-band updated values, so
time-varying confounders — where available in the source data — inform the
censoring model. Arm-specific fits fall back to the arm's marginal censoring
rate when a stratum has no (or all) censoring events or too few rows to
support the model.

The stabilised weight for the row in band $k$ is a ratio of cumulative
uncensored probabilities **through band $k$ inclusive**:

$$
SW^C_{m,k}
 = \frac{\prod_{j=0}^{k} \bar{q}_{a}(j)}
        {\prod_{j=0}^{k} \widehat{\mathrm{Pr}}(C_{m,j}=0 \mid \cdot)},
$$

where $\bar{q}_a(j)$ is the numerator: the **marginal** mean uncensored
probability at band $j$ within arm $a$ (by band only, when arms are pooled).

Two deliberate deviations from the textbook construction, and why:

- **Inclusive cumulative product.** Because the censoring-event row is
  subsequently removed, a row present at band $k$ exists if and only if the
  person-trial is uncensored *through* $k$; the weight therefore includes
  band $k$'s own uncensoring probability. (With the convention that censored
  bands stay in the risk set, the product would stop at $k-1$; the two
  conventions must not be mixed.)
- **Marginal numerator.** Canonical stabilisation (Danaei et al. 2013) uses a
  numerator model conditional on *baseline* covariates, which then requires
  those covariates in the outcome model. Here the outcome model is
  covariate-free (marginal MSM, 2.7), so the numerator is the marginal
  uncensored probability by band and arm. This preserves consistency of the
  marginal estimand; it stabilises slightly less aggressively when baseline
  covariates strongly predict censoring.

## 2.6 Final analysis weights and truncation

$$
W_{i,m,j} = SW^A_{i,m} \times SW^C_{i,m,j}
$$

for the per-protocol panel; $W_{i,m,j} = SW^A_{i,m}$ for the ITT panel.
Weights are truncated at percentiles (1st/99th by default) of the pooled
person-band rows: the ITT weight directly, and the PP weight as the truncated
*product* (component-wise truncation is not applied, so extreme components
can offset — sensitivity analyses may truncate components separately).
Primary analyses use truncated weights; untruncated PP results are exported
alongside as a sensitivity analysis.

**Positivity and the truncation tradeoff.** Truncation trades variance for
bias toward the null: in a simulation with near-violation of positivity
(propensity scores spanning 0–1), attenuation of the log-IRR grows
monotonically with truncation severity (measured in Section 3.2, Table 4).
Inspect the propensity-score overlap and the raw weight distribution; if
extreme weights are structural (deterministic treatment within a stratum),
restrict the eligible population instead of truncating harder.

## 2.7 Outcome model

The IRR is estimated by a weighted quasi-Poisson marginal structural model on
the analysis panel:

$$
\log \mathrm{E}[Y_{i,m,j}]
  = \beta_0 + \beta_1 A_{i,m,0}
  + \mathrm{ns}(j, 3)
  + f(m)
  + \log(\text{person-weeks}_{i,m,j}),
$$

fit by weighted quasi-Poisson regression with survey-linearised variance
(Section 4.1). $\mathrm{ns}(j, 3)$ is a natural cubic spline of follow-up
band (the discrete-time baseline-rate analogue); $f(m)$ is a natural spline
of the trial index with 3 df (linear when 2–4 bands; omitted for a single
band), adjusting smoothly for calendar trends while sharing one treatment
coefficient across trials (Danaei et al. 2013; Caniglia et al. 2023). No
confounders enter the outcome model: $\exp(\beta_1)$ is the marginal IRR
(Section 3.5 discusses marginal versus conditional estimands). Descriptive
weighted event counts, person-years (52.25 weeks/year), and rates per
100,000 person-years accompany each IRR.

## 2.8 Inference

Standard errors are survey-linearised (Huber–White sandwich) with clustering
on the **person** identifier — not the person-trial — accounting for repeated
person-trials and repeated bands within person (Hernán and Robins 2008;
Danaei et al. 2013; Su et al. 2024). Confidence intervals are Wald on the log
scale, $\exp(\hat\beta_1 \pm 1.96\,\widehat{se})$. Two caveats for the SAP:

- The variance treats the estimated weights (and the hot-deck imputation) as
  fixed. For stabilised weights this is typically slightly conservative for
  the treatment coefficient, but it is not exact; a person-level bootstrap of
  the entire pipeline is the fuller alternative for definitive reporting.
- Monte-Carlo calibration (Section 3.4, Table 8) shows near-nominal coverage
  where the estimand's assumptions hold, mild undercoverage under confounding
  with independent loss, and coverage degradation driven by bias — not by the
  variance estimator — when an estimand ignores informative loss.

## 2.9 Identifying assumptions

For the **ITT analogue**: (1) consistency; (2) no unmeasured confounding of
baseline assignment given the baseline confounders at each trial's baseline;
(3) positivity of assignment within confounder strata; (4) loss to follow-up
independent of the outcome (no censoring weights are applied to the ITT
panel — shown by simulation to hold under independent loss and to bias
the ITT estimate under informative loss, in swereg and `TrialEmulation`
alike; Section 3.1, scenario s3).

For the **per-protocol** estimand, additionally: (5) the censoring model
(2.5) captures all joint determinants of protocol deviation/loss and the
outcome, including their time-varying values as materialised in the source
data; (6) positivity of continued adherence. Under strong
treatment–confounder feedback the single-model IPCW approach retains residual
bias: time-updated censoring covariates remove part — not all — of the
deviation selection bias relative to freezing them at baseline (measured in
Section 3.2, Table 5); where feedback is central, g-methods beyond this
pipeline are indicated.

## 2.10 Heterogeneity, subgroups, and small cells

Effect heterogeneity across calendar time is tested by a joint Wald test of
the treatment × trial-index spline interaction; effect modification by
pre-specified baseline subgroups by treatment × subgroup interaction, with
stratified IRRs per level. Zero-event strata return no estimate rather than
an unstable one. Enrollments and outcomes are pre-specified in a
machine-readable study specification; results tables report weighted events,
person-years, rates, IRR, CI, and p-value per estimand, plus CONSORT-style
attrition (unique persons and person-trials separately, per Cashin et al.
2025).

## 2.11 Known limitations

- No grace periods and no cloning; deviation censors at the first mismatched
  band (2.3).
- No as-treated estimand.
- Single hot-deck imputation of missing baseline confounders (no variance
  propagation).
- Comparator downsampling (2.1) discards comparator information (efficiency,
  not bias).
- The propensity and censoring models are main-effects (plus smooth time)
  specifications; non-linearities must be pre-encoded as derived variables.

---

# 3. Validation evidence

Every number in this section is computed, not asserted. The tables are
rendered from a results artifact regenerated by rerunning the complete
validation battery — the same data-generating processes, truth calculations,
and fit wrappers that the package's test suite enforces in continuous
integration (Section 4.3 maps each layer to its test file and explains how to
regenerate the artifact).

Provenance: generated `r ev$meta$generated_utc` with swereg
`r ev$meta$swereg`, TrialEmulation `r ev$meta$trialemulation`, under
`r ev$meta$r`.

Reading the tables: unless stated otherwise, estimates and biases are on the
**log-IRR scale**, truth is computed by simulating the counterfactual
contrast directly (200,000 persons per arm, no loss — loss is a nuisance the
estimator must survive, not part of the estimand), and each single-dataset
cell is one draw at a fixed seed, where Monte Carlo noise is roughly
0.03–0.05 on the log scale — single-run biases of that order are
indistinguishable from zero. The multi-replicate cells (Tables 3, 7, and 8)
quantify bias and coverage across repeated draws.

## 3.1 Cross-package validation matrix (enrollment layer)

Three escalating scenarios are fed through the full triangle — known
potential-outcome truth, swereg, and `TrialEmulation` — for both estimands,
comparing point estimates and CIs on a common rate-ratio scale
(`TrialEmulation`'s odds ratios converted per Zhang and Yu 1998).
`TrialEmulation` is a peer required to recover the truth itself, not an
oracle.

```{r}
tri <- copy(ev$triangle)
tri[, nuisance := fcase(
  scenario == "s1", "none",
  scenario == "s2", "confounding + independent loss",
  scenario == "s3", "confounding + informative loss"
)]
tri[, `:=`(
  sw = ci3(sw_est, sw_lo, sw_hi),
  te = ci3(te_est, te_lo, te_hi),
  bias_sw = sg3(sw_est - truth),
  bias_te = sg3(te_est - truth),
  gap = sg3(sw_est - te_est)
)]
knitr::kable(
  tri[, .(
    scenario, nuisance, estimand,
    truth = sprintf("%.3f", truth), sw, te, bias_sw, bias_te, gap
  )],
  col.names = c(
    "Scenario", "Nuisances", "Estimand", "True log-IRR",
    "swereg [95% CI]", "TrialEmulation [95% CI]",
    "swereg bias", "TE bias", "swereg − TE"
  ),
  align = "lllrrrrrr",
  caption = paste0(
    "Table 1. Cross-package validation matrix (N = 20,000, T = 20 bands, ",
    "fixed seed). Log-IRR scale."
  )
)
x31 <- list(
  max_ok = tri[scenario != "s3" | estimand != "itt", max(abs(sw_est - truth))],
  itt_sw = tri[scenario == "s3" & estimand == "itt", sg3(sw_est - truth)],
  itt_te = tri[scenario == "s3" & estimand == "itt", sg3(te_est - truth)],
  itt_gap = tri[scenario == "s3" & estimand == "itt", sprintf("%.3f", abs(sw_est - te_est))]
)
```

Three things to read off Table 1. First, in every cell where the estimand's
assumptions hold (all of s1 and s2, and PP in s3), swereg's absolute bias is
at most `r sprintf("%.3f", x31$max_ok)` — within single-dataset Monte Carlo
noise of zero. Second, in s3 the per-protocol estimator stays close to the
truth *because* its censoring weights model the informative loss. Third — the
reason this matrix exists — the s3 ITT row shows both packages biased by
design (swereg `r x31$itt_sw`, TrialEmulation `r x31$itt_te`) while agreeing
with each other to within `r x31$itt_gap`: ITT carries no loss weight, so
informative loss biases it in any correct implementation, and cross-package
agreement is **not** evidence of correctness.

## 3.2 Stress matrix (enrollment layer)

Adversarial scenarios with planted truth, each pushing an edge the standard
matrix does not reach.

```{r}
st <- copy(ev$stress)
st[, covers := fifelse(truth >= lo & truth <= hi, "yes", "no")]
knitr::kable(
  st[, .(
    cell, estimand, truth = sprintf("%.3f", truth),
    est = ci3(est, lo, hi), bias = sg3(est - truth), covers, note
  )],
  col.names = c(
    "Cell", "Estimand", "True log-IRR", "Estimate [95% CI]",
    "Bias", "CI covers truth", "Note"
  ),
  align = "llrrrcl",
  caption = "Table 2. Stress cells, single dataset at fixed seed. Log-IRR scale."
)
```

The informative-attrition ITT row is the designed failure: with no loss
weight, heavy confounder-driven attrition biases ITT (its CI need not cover),
while PP with IPCW absorbs the same attrition. Determinism: refitting the PP
estimator (including the spline-based censoring model) on identical data
reproduced the estimate to a maximum absolute difference of
`r format(ev$stress_determinism$max_abs_delta)` — the pipeline has no
uncontrolled stochastic step.

```{r}
sh <- copy(ev$stress_harmful)
knitr::kable(
  sh[, .(
    seed, truth = sprintf("%.3f", truth),
    sw = sprintf("%.3f", sw_est), te = sprintf("%.3f", te_est),
    bias_sw = sg3(sw_est - truth), gap = sg3(sw_est - te_est)
  )],
  col.names = c(
    "Seed", "True log-IRR (cumulative-rate)", "swereg", "TrialEmulation",
    "swereg − truth", "swereg − TE"
  ),
  align = "lrrrrr",
  caption = paste0(
    "Table 3. Harmful effect (true log-IRR > 0) with strong depletion of ",
    "susceptibles, ITT, three seeds. Log-IRR scale."
  )
)
x32 <- list(
  mb = sh[, sg3(mean(sw_est - truth))],
  mg = sh[, sprintf("%.3f", max(abs(sw_est - te_est)))]
)
```

Under depletion of susceptibles the marginal hazard ratio declines over
follow-up, so the single pooled IRR (a person-time-weighted average, 2.2)
legitimately sits above the cumulative-rate truth — mean gap `r x32$mb`
across seeds. This is an estimand property, not an implementation defect:
swereg and `TrialEmulation` agree to within `r x32$mg` on every seed because
both target the same weighted-average summary.

```{r}
tc <- copy(ev$stress_trunc)
knitr::kable(
  tc[, .(
    trunc_percentiles, truth = sprintf("%.3f", truth),
    est = sprintf("%.3f", est), bias = sg3(est - truth),
    wmax = sprintf("%.0f", wmax_raw)
  )],
  col.names = c(
    "Truncation percentiles (%)", "True log-IRR", "Estimate",
    "Bias (attenuation)", "Max raw stabilised weight"
  ),
  align = "lrrrr",
  caption = paste0(
    "Table 4. Near-positivity violation: attenuation toward the null grows ",
    "monotonically with truncation severity. ITT, log-IRR scale."
  )
)
```

This quantifies the 2.6 tradeoff on a design whose propensity scores approach
the boundary (maximum raw stabilised weight
`r tc[1, sprintf("%.0f", wmax_raw)]`): each tightening of the truncation
percentiles buys variance at the price of measurable bias toward the null.

```{r}
tv <- copy(ev$stress_tv)
knitr::kable(
  tv[, .(
    fit, truth = sprintf("%.3f", truth), est = ci3(est, lo, hi),
    bias = sg3(est - truth), abs_bias = sprintf("%.3f", abs(est - truth))
  )],
  col.names = c(
    "Fit", "True log-IRR", "Estimate [95% CI]", "Bias", "|Bias|"
  ),
  align = "lrrrr",
  caption = paste0(
    "Table 5. Treatment–confounder feedback: the censoring model that sees ",
    "the time-updated confounder is strictly less biased than the one frozen ",
    "at baseline; residual PP bias remains (2.9). Log-IRR scale."
  )
)
```

## 3.3 Full-pipeline truth recovery (plan layer)

The enrollment layer above is validated in isolation; this layer drives the
complete production path — machine-readable specification → skeleton files
(including an ISO week-53 year) → sequential eligibility → per-band 2:1
matched enrollment → IPW → PP/ITT analysis files → weighted outcome model,
executed through the same worker subprocesses as a production run — against a
planted constant-hazard truth of IRR = 2.0 (marginally attenuated to 1.982 in
the confounded scenario by first-event frailty).

```{r}
pl <- copy(ev$plan)
pl[, covers_pp := fifelse(pp_lo <= truth_pp & truth_pp <= pp_hi, "yes", "no")]
pl[, covers_itt := fifelse(itt_lo <= truth_itt & truth_itt <= itt_hi, "yes", "no")]
knitr::kable(
  pl[, .(
    cell, loss, n = format(n_persons, big.mark = ","),
    truth_pp = sprintf("%.2f", truth_pp),
    pp = ci2(pp_irr, pp_lo, pp_hi), covers_pp,
    truth_itt = sprintf("%.2f", truth_itt),
    itt = ci2(itt_irr, itt_lo, itt_hi), covers_itt
  )],
  col.names = c(
    "Cell", "Loss", "Persons", "PP truth", "PP IRR [95% CI]", "covers",
    "ITT truth", "ITT IRR [95% CI]", "covers"
  ),
  align = "llrrrcrrc",
  caption = paste0(
    "Table 6. Plan-layer factorial (A = no confounding, B = baseline ",
    "confounding; each × no/independent/informative loss) plus a 4%/week ",
    "discontinuation cell. IRR scale."
  )
)
x33 <- list(
  crude = pl[cell == "B_none", sprintf("%.2f", crude_rr)],
  ipw = pl[cell == "B_none", sprintf("%.2f", ipw_rr)],
  truth_b = pl[cell == "B_none", sprintf("%.2f", truth_pp)],
  disc_gap = pl[cell == "DISC", sg3(log(pp_irr) - log(itt_irr))],
  disc_truth_gap = pl[cell == "DISC", sg3(log(truth_pp) - log(truth_itt))]
)
```

Two rows deserve comment. In the confounded no-loss cell (B_none) the
confounder is doing real work: the crude rate ratio is
`r x33$crude` versus an IPW-weighted `r x33$ipw` against a marginal truth of
`r x33$truth_b` — the weighting removes essentially all of the planted
confounding. In the discontinuation cell (DISC) the two estimands genuinely
separate, PP − ITT = `r x33$disc_gap` on the log scale against a true
separation of `r x33$disc_truth_gap`: PP censors at deviation and
IPCW-reweights back to the sustained-treatment truth of 2.0, while ITT
attenuates toward the do(initiate) truth, exactly as designed. This cell also
exercises the event-priority convention (2.3): events falling in the
deviation band are counted as events.

```{r}
mc <- copy(ev$plan_mc)
mcs <- rbind(
  mc[, .(
    estimand = "pp",
    mean_bias = mean(log(pp_irr) - log(truth)),
    mc_sd = sd(log(pp_irr)),
    covered = sum(pp_lo <= truth & truth <= pp_hi),
    n_seeds = .N
  ), by = scenario],
  mc[, .(
    estimand = "itt",
    mean_bias = mean(log(itt_irr) - log(truth)),
    mc_sd = sd(log(itt_irr)),
    covered = sum(itt_lo <= truth & truth <= itt_hi),
    n_seeds = .N
  ), by = scenario]
)
setorder(mcs, scenario, -estimand)
knitr::kable(
  mcs[, .(
    scenario, estimand, mean_bias = sg3(mean_bias),
    mc_sd = sprintf("%.3f", mc_sd),
    coverage = sprintf("%d/%d", covered, n_seeds)
  )],
  col.names = c(
    "Scenario", "Estimand", "Mean log bias", "MC sd", "95% CI coverage"
  ),
  align = "llrrr",
  caption = paste0(
    "Table 7. Plan-layer Monte Carlo: 8 independent seeds per scenario at ",
    "6,000 persons, full pipeline per replicate. Log-IRR scale."
  )
)
```

## 3.4 Coverage calibration

A Monte-Carlo coverage study for the ITT sandwich standard errors: 200
replicate draws per scenario, each refit end to end, asking what fraction of
95% CIs cover the population truth.

```{r}
cv <- copy(ev$coverage)
cv[, nuisance := fcase(
  scenario == "s1", "none",
  scenario == "s2", "confounding + independent loss",
  scenario == "s3", "confounding + informative loss"
)]
knitr::kable(
  cv[, .(
    scenario, nuisance,
    reps = sprintf("%d/%d", n_fit, M),
    mean_bias = sg3(mc_mean_bias), mc_sd = sprintf("%.3f", mc_sd),
    coverage = sprintf("%d/%d (%.1f%%)", covered_n, n_fit, 100 * coverage)
  )],
  col.names = c(
    "Scenario", "Nuisances", "Replicates fit", "Mean log bias",
    "MC sd", "95% CI coverage"
  ),
  align = "llrrrr",
  caption = paste0(
    "Table 8. ITT coverage calibration, M = 200 replicates per scenario at ",
    "N = 3,000. Log-IRR scale."
  )
)
x34 <- list(
  c1 = cv[scenario == "s1", sprintf("%.1f%%", 100 * coverage)],
  c2 = cv[scenario == "s2", sprintf("%.1f%%", 100 * coverage)],
  c3 = cv[scenario == "s3", sprintf("%.1f%%", 100 * coverage)],
  b3 = cv[scenario == "s3", sg3(mc_mean_bias)]
)
```

Where the estimand's assumptions hold (s1) coverage is `r x34$c1` — the
sandwich variance is calibrated. Under confounding with independent loss (s2)
coverage dips mildly to `r x34$c2`, typical for IPW with estimated weights
treated as fixed (2.8). Under informative loss (s3) coverage degrades to
`r x34$c3` — but Table 8 shows why: the mean bias (`r x34$b3`) is what eats
the coverage, not a failing variance estimator. A correct SE cannot rescue a
biased point estimate; the remedy is the per-protocol estimand with censoring
weights, not a wider interval.

## 3.5 Marginal versus conditional estimands

swereg and `TrialEmulation` both remove baseline confounding, but by
different routes, producing two distinct — each valid — estimands. swereg
weights and fits a covariate-free model: a *marginal* effect. `TrialEmulation`
conventionally adjusts the outcome model: a *conditional* effect. Rate ratios
are collapsible, so these coincide for the IRR; odds ratios are not, so the
`TrialEmulation` OR is converted with the Zhang–Yu relation
$\mathrm{RR} = \mathrm{OR} / (1 - p_0 + p_0\,\mathrm{OR})$ ($p_0$ =
reference-arm per-band risk) before comparison. The conversion removes the
scale gap only; a residual conditional-versus-marginal difference remains
(visible in Table 1 as the small swereg − TE gaps in the confounded
scenarios, larger for ITT than PP). The primary correctness guarantee is
each implementation's agreement with the *known simulated truth* on its own
scale.

---

# 4. Implementation mapping

The SAP (Section 2) is deliberately implementation-agnostic. This section
reveals the code: which function, argument, and option realises each step,
the provenance of estimator-behaviour changes, and where the validation
evidence comes from.

## 4.1 SAP step → code

| SAP | Step | Implementation |
|:--|:--|:--|
| 2.1 | Band width $w$ | `period_width` (default 4 weeks) in the trial-band assignment inside `TTEPlan` |
| 2.1 | Sequential eligibility, enrollment, matching | `TTEPlan$s1_generate_enrollments_and_ipw()`; `matching_ratio` and `seed` from the YAML spec's `treatment.implementation` |
| 2.1 | Washout / new-user exclusion | Spec-level exclusion: `type: no_prior_intervention` with `window: lifetime_before_baseline`, or a finite `window` in weeks |
| 2.1 | Prevalent-user warning | `tteplan_read_spec()` warns when a spec lacks a washout exclusion; silence with `options(swereg.warn_prevalent_user = FALSE)` |
| 2.3 | Follow-up stop events, event priority | `TTEEnrollment$s5_prepare_outcome()`; horizon from `follow_up`, administrative end of study from `admin_censor_isoyearweek` |
| 2.4 | Hot-deck imputation | `TTEEnrollment$s1_impute_confounders(seed = 4)` |
| 2.4 | Stabilised IPW | `TTEEnrollment$s2_ipw(stabilize = TRUE)` |
| 2.5 | IPCW censoring model | `TTEEnrollment$s6_ipcw_pp()` via `s4_prepare_for_analysis(estimate_ipcw_pp_with_gam = TRUE, estimate_ipcw_pp_separately_by_treatment = TRUE)`; GAM engine `mgcv::bam(..., discrete = TRUE)`; `estimate_ipcw_pp_with_gam = FALSE` gives the linear-in-time sensitivity variant |
| 2.6 | Weight truncation | `TTEEnrollment$s3_truncate_weights(lower = 0.01, upper = 0.99)`; truncated columns `ipw_trunc` (ITT) and `analysis_weight_pp_trunc` (PP product weight); untruncated PP results exported as a sensitivity sheet |
| 2.7–2.8 | Outcome model + inference | `TTEEnrollment$irr(weight_col)`: `survey::svydesign(ids = ~person)` + `survey::svyglm(family = quasipoisson())` with `splines::ns()` terms for follow-up and trial index |
| 2.10 | Pre-specification | YAML spec parsed by `tteplan_read_spec()`; full grid run by `TTEPlan$s1_…`/`s2_…`/`s3_analyze()` |

## 4.2 Provenance notes

- **Event-priority convention (2.3).** Enforced since swereg 26.7.3.
  Previously, a first event falling in the same band as the first protocol
  deviation was dropped as a censoring, which undercounted per-protocol
  events in switching-heavy data (≈10% of PP events in a high-switching
  simulation; negligible under strong persistence).
- **Per-person administrative censoring.** `admin_censor_var` (a per-person
  censoring column) is accepted by the constructor for backward compatibility
  but is not implemented in outcome preparation and now errors loudly rather
  than silently doing nothing; use `admin_censor_isoyearweek`.

## 4.3 Where the validation numbers come from

The four evidence layers in Section 3 are permanent, executable tests:

| Section | Layer | Test file | Gate |
|:--|:--|:--|:--|
| 3.1 | Cross-package triangle | `tests/testthat/test-tte_validation_matrix.R` | runs in CI |
| 3.2 | Stress matrix | `tests/testthat/test-tte_stress_matrix.R` | fast subset in CI; full battery `SWEREG_RUN_STRESS=true` |
| 3.3 | Plan-layer truth matrix | `tests/testthat/test-tteplan_truth_matrix.R` | reduced-N subset in CI; full factorial `SWEREG_RUN_PLAN_MATRIX=true` |
| 3.4 | Coverage calibration | `tests/testthat/test-tte_coverage.R` | opt-in only, `SWEREG_RUN_COVERAGE=true` |

The tables themselves are rendered from
`vignettes/tte-validation-evidence.rds`, regenerated by
`dev/generate_validation_evidence.R` (in the source repository, not the
installed package). The script reruns every cell through the same
DGP/truth/fit helpers the tests source (`tests/testthat/helper-tte_*.R`), so
the vignette's numbers and the suite's assertions cannot drift apart; rerun
it after any estimator change and commit the refreshed artifact alongside.

## References

- Hernán MA, Robins JM. Observational studies analyzed like randomized
  experiments: an application to postmenopausal hormone therapy and
  coronary heart disease. *Epidemiology* 2008;19(6):766–779.
- Hernán MA, Robins JM. Using big data to emulate a target trial when a
  randomized trial is not available. *Am J Epidemiol* 2016;183(8):758–764.
- Danaei G, García Rodríguez LA, Cantero OF, Logan R, Hernán MA.
  Observational data for comparative effectiveness research: an emulation
  of randomised trials of statins and primary prevention of coronary heart
  disease. *Stat Methods Med Res* 2013;22(1):70–96.
- Caniglia EC, et al. Emulating a sequence of target trials to avoid
  immortal time bias: an application in pregnancy. *Am J Epidemiol* 2023.
- Cashin AG, et al. Emulating a target trial — the TARGET statement.
  *JAMA* 2025.
- Su L, Rezvani R, Seaman SR, Bartlett JW. *TrialEmulation: An R package
  to emulate target trials for time-to-event data from electronic health
  records.* arXiv:2402.12083, 2024.
- Zhang J, Yu KF. What's the relative risk? A method of correcting the odds
  ratio in cohort studies of common outcomes. *JAMA* 1998;280(19):1690–1691.
