---
title: "TTE methods: manuscript and statistical analysis plan text"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TTE methods: manuscript and statistical analysis plan text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(data.table)
library(ggplot2)
ev <- readRDS("tte-validation-evidence.rds")
ci3 <- function(est, lo, hi) sprintf("%.3f [%.3f, %.3f]", est, lo, hi)
ci2 <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)
sg3 <- function(x) sprintf("%+.3f", x)
scen_lab <- function(s) {
  fcase(
    s == "s1", "s1: no confounding, no loss",
    s == "s2", "s2: confounding, independent loss",
    s == "s3", "s3: confounding, informative loss"
  )
}
col_fit <- c(
  "swereg, truncated weights" = "#8a1c1c", # primary red
  "swereg, untruncated weights" = "#E69F00", # orange
  "TrialEmulation" = "#0072B2" # blue
)
col_var <- c(
  "truncated (primary)" = "#8a1c1c", # primary red
  "untruncated (sensitivity)" = "#E69F00", # orange
  "TrialEmulation (reference)" = "#0072B2" # blue
)
```

This vignette gives methods text for the target trial emulation (TTE)
pipeline in swereg (`TTEEnrollment`, `TTEPlan` and related classes). It
covers the intention-to-treat (ITT) and the per-protocol (PP) estimand. It has
four sections:

- **Statistical analysis plan** (Section 1). The values each study states, the
  formulas, model specifications, conventions, identifying assumptions and
  known limitations. It follows the target trial table of the TARGET
  guideline. It names no code, so a statistician can reimplement the
  estimators from it. The one exception is 1.8.9, which names the three time
  columns. Copy it whole into a protocol, a pre-registered
  statistical analysis plan or a methods supplement. The section numbers stay
  valid.
- **Manuscript methods** (Section 2). Short past-tense text for the methods
  section of a journal article, in the order of Section 1. Replace the
  placeholders that Section 2 lists.
- **Validation evidence** (Section 3). The design of the validation battery,
  the data-generating processes, and tables and figures of estimate against
  truth for every validation cell. The numbers are rendered from a saved
  results file.
- **Implementation mapping** (Section 4). The function, argument, option,
  column and test file behind each step of Section 1, and notes on changes to
  the estimators.

Section 1 describes what the code computes. Where the implementation differs
from a standard reference construction, Section 1 states the difference and
the reason. The implementation follows the sequential trial emulation
literature (Hernán et al. 2008; Danaei et al. 2013; Hernán and Robins 2016;
Caniglia et al. 2023; Cashin et al. 2025).

---

# 1. Statistical analysis plan

This section specifies the estimators as swereg implements them. It names no
code, so a statistician can reimplement the estimators from it. Sections 1.1
to 1.8 follow the rows of the target trial table in TARGET items 6 and 7
(Cashin et al. 2025). Sections 1.9 and 1.10 add the sensitivity analyses and
the limitations. The validation documentation of the software reports the
simulation evidence for its quantitative statements.

## Values your protocol must state

Each study states the values below in its protocol. Each item names the value
that swereg uses when the specification is silent, or says that there is none.

- The width of each trial's enrollment period. Default: 4 weeks. The same
  width sets each follow-up interval (1.4).
- The new-user washout window. Default: no washout. swereg then warns that no
  washout covers the intervention (1.1).
- The sampling ratio for comparators. No default: the specification states
  it, for example 2:1, or swereg stops (1.3).
- The analysis horizon and the administrative end of study. No default
  horizon: the specification lists each one. The administrative end defaults
  to the last week in the first batch of the data (1.4).
- The tolerance for each arm. Default: 0 weeks in each arm, so the first
  discordant week is a deviation (1.4).
- The confounders in the treatment weight model and in the censoring weight
  model. No default. One list of confounders serves both models (1.8.1,
  1.8.2).
- The truncation percentiles. Default: the 1st and 99th percentiles, which
  the pipeline holds fixed (1.8.3).
- Subgroups, and how heterogeneity is tested. Default: no subgroups. Each
  subgroup gets rate ratios within its levels and a Wald test of its
  interaction with treatment (1.8.8).
- Missing data, and how it is handled. Default: the rules of 1.8.7. They
  include a single hot-deck imputation of a missing confounder value at entry.

## 1.1 Eligibility criteria

TARGET items 6a and 7a.

Calendar time is cut into consecutive enrollment periods of equal width. Each
enrollment period opens one trial. Eligibility is assessed in every week of
the data. A person can therefore be eligible in some weeks of an enrollment
period and not in others.

The study states global criteria: a range of ISO years, inclusion criteria and
exclusion criteria. Each enrollment can add its own age range, range of ISO
years, inclusion criteria and exclusion criteria. These apply after the global
criteria. An enrollment takes one age range and one range of ISO years at
most.

An enrollment's range of ISO years limits trial entry week by week. An
enrollment period that crosses a new year therefore recruits from its
in-range weeks only. Follow-up continues past the end of the range. The range
lies inside the global range of ISO years.

A look-back window counts the calendar ISO weeks before the current week, not
the data rows. A week with no data row still counts toward the window. An
annual record covers every ISO week of its year. A lifetime window covers
every earlier week. The current week is never inside its own window.

One further exclusion window reads every week of the person, after baseline
as well. It removes a person from every trial because of a later event, so
eligibility then depends on the future.

swereg imposes no new-user rule of its own. A washout exclusion on the
treatment history makes the design a new-user design. Its window is a fixed
number of weeks, for example 104 weeks as in Danaei et al. (2013). For a
never-user design, the window is the whole earlier history.

A lifetime washout lets each person start treatment in one enrollment period
at most. Every later week is then ineligible, so the person leaves all later
trials. A finite washout expires. It stops excluding the person once its
window holds no week at the washed-out value.

A protocol without any washout enrolls prevalent users in the intervention
arm at every enrollment period. People who stop treatment can re-enter as
comparators. That is a prevalent-user design, and it is rarely the intended
estimand.

swereg warns when no washout covers the intervention. A prevalent week is a
week at the intervention value after an earlier week of the same person at
that value. A washout covers the enrollment when it makes every prevalent week
ineligible. A washout on a parent column can therefore cover a sub-type arm,
and a washout on the right column at the wrong value cannot. The check reads
the weekly rows of the first data batch only, so it misses a counterexample
that exists only in a later batch.

## 1.2 Treatment strategies

TARGET items 6b and 7b.

The protocol compares two strategies, the intervention and the comparator. One
column of the weekly data holds the treatment of each week. The protocol names
the value of that column that marks each arm. A week that holds neither value
is outside both arms.

Under the per-protocol estimand, a person follows the assigned strategy while
each week holds the value of the assigned arm. Section 1.4 states when a
departure ends follow-up.

One value per week cannot record two treatments in the same week. The washout
exclusions cannot resolve that case either, because each one reads earlier
weeks only. The protocol decides the rule when it builds the column. A week
coded as missing drops out of the arm classification, and it has further
effects under some washouts (1.8.7).

## 1.3 Assignment procedures

TARGET items 6c and 7c.

Arm classification in an enrollment period reads only the weeks in which the
person is eligible and holds one of the two arm values. It drops every other
week first. A person enters the intervention arm when at least one of those
weeks holds the intervention value. A person enters the comparator arm when
all of them hold the comparator value. A candidate person-trial with no such
week is ineligible for that trial.

A week outside both arms therefore does not prevent a comparator
classification. The recruiting week is the earliest week of the enrollment
period that is eligible and holds an arm value. The treatment weights read
each confounder at the recruiting week (1.8.1).

The worked example below shows the filter. A protocol compares intervention A
with comparator B. One column holds the treatment of each week, and it reads
A, B or missing. Two washout exclusions apply to both arms: no earlier A, and
no earlier B. Each one covers the whole earlier history. One person starts B
in week 1 of a four-week enrollment period, and starts A in week 3.

| Week | Treatment | Eligible | Why |
|---|---|---|---|
| 1 | B | yes | no earlier week holds A or B |
| 2 | missing | no | week 1 holds B |
| 3 | A | no | week 1 holds B |
| 4 | A | no | week 1 holds B |

The classification keeps week 1 alone. That week holds the comparator value,
so the person enters the comparator arm, and the recruiting week is week 1.
The two A weeks change nothing, because the filter already dropped them.

The current week is outside its own look-back window (1.1). The week of a
first start therefore stays eligible. Under a lifetime washout on both arms,
each person enters the arm of the treatment they started first. No separate
rule is needed for a person who starts both treatments in one enrollment
period.

All intervention person-trials of an enrollment period are enrolled.
Comparators enter by incidence density sampling from the same enrollment
period, with a seed stated in the specification. A specification without a
seed is refused, so every draw can be reproduced. The draw takes the
comparator-to-intervention ratio times the number of intervention
person-trials, rounded to a whole number. Where fewer qualified comparators
remain, the draw takes all of them.

The draw runs after the time-zero checks of 1.4, so it refills the ratio from
qualified comparators alone. It is stratified by the enrollment period, not by
the week, and it reads no other variable. It attaches no comparator to an
intervention person-trial, so it forms no matched set, and no later step
conditions on one. The weights carry all confounding adjustment (1.8.1). The
draw bounds computation.

## 1.4 Follow-up

TARGET items 6d and 7d.

Time zero is a landmark: the first week after the enrollment period closes. A
person enters the trial only if they reach that week under observation and
free of every enrollment outcome.

swereg checks both conditions on each candidate person-trial before the
comparator draw. The outcome check reads every outcome of the enrollment, over the whole earlier
history of the person. An outcome in the time-zero week is a follow-up event.
The enrollment period therefore contributes no follow-up and no immortal time
(Caniglia et al. 2023). Its width sets how long a newly eligible person waits
for the next time zero.

Each enrolled person-trial is split into follow-up intervals of the same width
as the enrollment period. Their number is the horizon divided by the width,
rounded up. The first interval opens at time zero, and the last one is clipped
at the horizon. Follow-up time counts weeks from time zero.

For each person-trial, follow-up stops at the first of these events, with
one exception for the per-protocol estimand:

1. the first outcome at or after time zero;
2. loss of observation: the first week after time zero without observation,
   whether a gap in observation or the end of the record;
3. under the per-protocol estimand only, a protocol deviation;
4. the administrative end of study;
5. the analysis horizon.

The exception: under the per-protocol estimand, an outcome can fall in the
same follow-up interval as an earlier deviation. The person-trial then exits
through the outcome, at its exact week. The row counts as an event and not as
a censoring, in the censoring model and in the analysis data. The opposite
convention discards real events and undercounts per-protocol outcomes in data
with frequent switching.

The exception never applies to a loss of observation. A gap is a week without
observation between two observed weeks. It stops follow-up under both
estimands at the start of its first absent week, and no tolerance applies. An
outcome in that week or later is never counted, even in the same follow-up
interval. The row that reaches the gap is clipped there and marked as censored
by loss of observation.

Every stop is exact to the week. None is rounded to the edge of a follow-up
interval. A trial whose time zero falls after the administrative end
contributes no follow-up.

Under the per-protocol estimand, deviation is read from the weekly records. A
week is discordant when it does not hold the value of the assigned arm. A week with a missing treatment value
and a week outside both arms are discordant in both arms.

Each arm has a tolerance ${k}$, the number of consecutive discordant weeks it
allows. A concordant week resets the run. The deviation falls at the right
edge of week $({k}+1)$ of a run of consecutive discordant weeks. A tolerance
of 0 therefore places it at the first discordant week.

A run that starts before time zero counts only its weeks from time zero. An
absent week breaks a run. No grace period applies, and the pipeline does not
clone person-trials. A treatment deviation never stops follow-up under the
intention-to-treat analogue.

Rows at and before the stop are kept. The row that reaches the stop is clipped
at that week, so it holds the person-time before the stop and none after it.
The outcome sits at its exact week: the row that holds it is clipped there and
carries the event. Person-time is the stop minus the start, in weeks, on a
half-open interval. A time-updated confounder takes its value in the first
week of the interval. The censoring model of 1.8.2 fits on the clipped rows.

## 1.5 Outcomes

TARGET items 6e and 7e.

Each outcome is a weekly indicator in the data, and the protocol names the
primary one. Every combination of enrollment, outcome and horizon is one
emulated trial, analysed on its own. Follow-up counts the first occurrence of
the outcome at or after time zero. An earlier occurrence of any outcome of the
enrollment excludes the candidate person-trial (1.4).

The risk is cause-specific. Death and the end of observation censor follow-up,
and no competing-risk model is fitted (1.6).

## 1.6 Causal contrasts

TARGET items 6f and 7f.

Both estimands condition on reaching time zero under observation and free of
every enrollment outcome. The target of inference is therefore a
landmark-survivor estimand. It says nothing about people who die, or have an
outcome, before time zero (Dafni 2011). Both estimands are marginal: the
weights standardise them over the baseline confounder distribution of the
enrolled person-trials.

- The intention-to-treat analogue is the effect of the intervention versus the
  comparator strategy as assigned in the enrollment period, ignoring later
  changes of treatment. The treatment weight alone estimates it.
- The per-protocol estimand is the effect of sustained intervention versus the
  sustained comparator strategy. Follow-up is censored at a deviation, when a
  run of discordant weeks exceeds the arm's tolerance (1.4). The product of
  the
  treatment and censoring weights estimates it.

Each estimand is reported on two scales. The relative scale is the marginal
incidence rate ratio (IRR) of 1.8.4. The absolute scale is the risk difference
at follow-up week ${t}$:

$$
\mathrm{RD}(t) = \{1 - S_{1}(t)\} - \{1 - S_{0}(t)\} = S_{0}(t) - S_{1}(t).
$$

Here $S_{a}(t)$ is the weighted survival of arm ${a}$ through week ${t}$
(1.8.5). Arm ${a} = 1$ is the intervention and arm ${a} = 0$ the comparator.
A protective intervention gives a negative risk difference. The risk
difference is estimated at every distinct stop time on the weekly grid, and
reported at the end of follow-up.

The number needed to treat is $-1/\mathrm{RD}(t)$. A positive value is the
number needed to treat for benefit (NNTB). A negative value is reported by its
magnitude as the number needed to treat for harm (NNTH) (Altman 1998). A risk
difference of exactly zero has no number needed to treat.

The risk is the net risk implied by the cause-specific hazard of the outcome.
It describes a world without death only if removing death would not change
that hazard. It is not the cumulative incidence with death as a competing
risk. It is at least as large as the cumulative incidence computed from the
same hazards.

The IRR is the coefficient of a proportional-rates working model. The
marginal rate ratio can change over follow-up, for example through depletion
of susceptible people or an effect that accumulates. The IRR is then a
person-time-weighted average of the time-varying rate ratio. That average can
differ from the ratio of the risks at the end of follow-up. In the simulations
of section 3, swereg and TrialEmulation estimate the same weighted average
under strongly time-varying effects.

The risk-difference curve shows how the effect develops over follow-up. Report
it beside the IRR when a time-varying effect is of scientific interest.

## 1.7 Identifying assumptions

TARGET items 6g, 7g.i and 7g.ii.

The intention-to-treat analogue rests on four assumptions:

1. consistency;
2. no unmeasured confounding of assignment, given the confounders at the
   recruiting week;
3. positivity of assignment within confounder strata;
4. loss to follow-up that is independent of the outcome.

No censoring weights apply to the intention-to-treat analysis. In the
simulations of section 3, the estimand holds under independent loss and is
biased under informative loss, in swereg and TrialEmulation alike.

The per-protocol estimand rests on two further assumptions:

5. the censoring model of 1.8.2 captures all common causes of deviation or
   loss and of the outcome, including their time-varying values in the data;
6. positivity of continued adherence.

The variables behind assumptions 2 and 5 are the confounders that the protocol
lists. Assumption 2 reads them at the recruiting week. Assumption 5 reads
their most recent value in each follow-up interval (1.8.7).

Under strong feedback between treatment and confounders, the single-model
censoring weights keep residual bias. Time-updated censoring covariates remove
part of the deviation selection bias, relative to covariates frozen at
baseline, and not all of it. Section 3 measures that residual bias. Where
feedback is central, g-methods are indicated: the parametric g-formula, or
g-estimation of structural nested models. This pipeline does not implement
them.

Unmeasured prognostic factors that drive adherence or loss violate assumption
5, for example a healthy-adherer mechanism. They bias the per-protocol
estimand in any implementation, and the weight diagnostics cannot detect them.
The design addresses them, for example through negative-control outcomes or
sensitivity analyses for unmeasured selection.

## 1.8 Data analysis

TARGET items 6h and 7h.i.

The study specification is machine-readable. It states the enrollments, the
outcomes, the horizons and the subgroups. For each estimand, the results
report weighted events, person-years and rates. They also report the IRR with
its interval and p-value, the risk difference and the number needed to treat.
The attrition counts persons and person-trials separately.

The formulas below index persons by ${i}$, trials by ${m}$ and the follow-up
intervals of a trial by ${j} = 0, \ldots, {K}-1$. Here ${K}$ is the horizon
divided by the width of the enrollment period, rounded up. $A_{i,m,0}$ is the
assigned arm, 1 for the intervention. $L_{i,m,j}$ is the confounder vector as
most recently updated in interval ${j}$, and $L_{i,m,0}$ is its value at the
recruiting week. $Y_{i,m,j}$ is the outcome indicator. $C_{i,m,j}$ marks
censoring by deviation or loss in interval ${j}$.

$c_{m,j}$ is the calendar period of interval ${j}$. Calendar periods are the
blocks of calendar time that define the enrollment periods, so each follow-up
interval lies in exactly one. A trial with several follow-up intervals
therefore spans several calendar periods. Section 1.8.9 states the three time
axes, and the reason for the time terms of each model.

### 1.8.1 Treatment weights

The baseline row of each person-trial is its first follow-up row, at
follow-up week 0. A logistic regression of the assigned arm on the confounders
at the recruiting week, as main effects, is fit there:

$$
\mathrm{logit}\,\mathrm{Pr}(A_{m,0} = 1 \mid L_{m,0})
  = \gamma_{0} + \gamma^\top L_{m,0}.
$$

The model reads the value at the recruiting week, because the value at time
zero would adjust for the wrong instant. One model covers all trials of the
enrollment, and it holds no term for the trial. The stabilised weight uses the
marginal fraction of intervention person-trials as numerator:

$$
SW^{A} =
  A_{m,0}\,\frac{\bar{p}}{\widehat{ps}} +
  (1 - A_{m,0})\,\frac{1-\bar{p}}{1-\widehat{ps}},
\qquad \bar{p} = \widehat{\mathrm{Pr}}(A_{m,0}=1),
\qquad \widehat{ps} = \widehat{\mathrm{Pr}}(A_{m,0}=1 \mid L_{m,0}).
$$

The weight is constant across the rows of a person-trial. The propensity model
holds main effects only. The protocol encodes strong non-linearity or
interactions as derived confounders.

### 1.8.2 Censoring weights

The censoring indicator marks the row at which a deviation or a loss of
observation stops follow-up. The row of an outcome is never marked as censored
(1.4). The model fits on the data that keep the censoring row. By default
it fits separately in each arm ${a}$, as a discrete-time complementary log-log
generalised additive model with a person-time offset:

$$
\mathrm{cloglog}\,\mathrm{Pr}(C_{m,j} = 1 \mid \text{at risk},\ A_{m,0}=a)
  = s_{a}(u_{m,j}) + g_{a}(c_{m,j}) + \alpha_{a}^\top L_{m,j}
  + \log \Delta_{m,j}.
$$

Here $u_{m,j}$ is the start of interval ${j}$ in weeks from time zero.
$\Delta_{m,j}$ is the width of the row in weeks, and $\eta_{m,j}$ is the
linear predictor without the offset. The uncensoring probability is then
$q_{m,j} = \widehat{\mathrm{Pr}}(C_{m,j}=0 \mid \cdot) = \exp\{-\exp(\eta_{m,j})\Delta_{m,j}\}$.
One linear predictor therefore gives ${q}(4) = {q}(1)^{4}$.

That identity makes a four-week row and a one-week row comparable. It matters
because the last row is clipped at its exact stop, so it can be narrower than
a whole interval. A logit link carries no such identity.

Both time terms follow the term ladder of 1.8.9. The follow-up term $s_{a}$
depends on the number of distinct interval starts in the data of the arm:

- 10 or more: a penalised spline;
- 4 to 9: a natural cubic spline of 3 degrees of freedom;
- 2 or 3: a factor;
- 1: no term.

A penalised spline asks for 10 basis functions, so it needs 10 distinct
values. The calendar term $g_{a}$ follows the same ladder on the number of
distinct calendar periods in the data of the arm. Both counts exclude the rows
of zero width, which hold no person-time.

The confounders carry their updated value in each interval. Time-varying
confounders, where the data hold them, therefore inform the censoring model.
Section 1.8.7 states how a missing value is filled.

A stratum with no censored row takes an uncensoring probability of exactly 1
on every row, and so a weight of 1. A stratum with no uncensored row stops the
run, and so does a model that cannot be fit. swereg substitutes no marginal
censoring rate for a model it could not fit.

The stabilised weight for the row in interval ${k}$ is a ratio of cumulative
uncensoring probabilities through the start of that interval:

$$
SW^{C}_{m,k}
 = \frac{\prod_{j=0}^{k-1} \bar{q}_{a}(j)}
        {\prod_{j=0}^{k-1} \widehat{\mathrm{Pr}}(C_{m,j}=0 \mid \cdot)}.
$$

Here $\bar{q}_{a}(j)$ is the numerator: a second fit in the same arm, which
keeps the follow-up term $s_{a}$ and drops the calendar term and the
confounders (1.8.9). Two rules govern the ratio.

- The product is lagged. It stops at interval ${k}-1$, so the uncensoring
  probability of a row does not enter its own weight. A censored interval
  stays in the risk set. The empty product gives the first interval of every
  person-trial a weight of exactly 1. An inclusive product belongs to data
  that delete the censoring row, and the two conventions are not mixed.
- The numerator is marginal. Canonical stabilisation (Danaei et al. 2013)
  uses a numerator model conditional on baseline covariates, which then enter
  the outcome model too. Here the outcome model holds no confounders (1.8.4),
  so the numerator model carries the follow-up term alone. This keeps the
  marginal estimand consistent. It stabilises less when baseline covariates
  strongly predict censoring.

### 1.8.3 Truncation

The analysis weight is the product of the two weights for the per-protocol
estimand, and the treatment weight alone for the intention-to-treat analogue:

$$
W_{i,m,j} = SW^{A}_{i,m} \times SW^{C}_{i,m,j} \;\; \text{(per-protocol)},
\qquad
W_{i,m,j} = SW^{A}_{i,m} \;\; \text{(intention-to-treat)}.
$$

Weights are truncated at the 1st and 99th percentiles of the pooled rows of
all person-trials. The intention-to-treat analysis truncates the treatment
weight. The per-protocol analysis truncates the product and not its
components, so extreme components can offset each other. Primary analyses use
truncated weights. The untruncated per-protocol result is reported beside them
as a sensitivity analysis (1.9).

Truncation trades bias for variance. Clipping the weight tails reduces the
variance of the estimator. It also under-corrects the confounding or selection
that the clipped weights carried, and that moves the estimate. Under
near-violations of treatment positivity, it moves the estimate toward the
null. The shift grows with how strongly measured covariates drive censoring.

The truncated weight is the primary analysis on simulation evidence, which
section 3 reports. The scenarios included heavy loss to follow-up strongly
driven by covariates. In every per-protocol scenario, the truncated fit had
the smaller sampling spread, and its bias stayed bounded. Its root-mean-squared
error was lower than, or about equal to, that of the untruncated fit. The
untruncated fit was less biased on average when the censoring weights were
heavy-tailed. Its sampling spread was several times larger, and in some
scenarios its bias was larger too.

### 1.8.4 Outcome model

The IRR comes from a weighted quasi-Poisson marginal structural model on the
analysis data:

$$
\log \mathrm{E}[Y_{i,m,j}]
  = \beta_{0} + \beta_{1} A_{i,m,0}
  + h(u_{m,j})
  + f(m)
  + \log(\text{person-weeks}_{i,m,j}).
$$

Here $u_{m,j}$ is the start of interval ${j}$ in weeks from time zero, and
${m}$ is the trial. The person-weeks are the width of the row after clipping.
Both time terms follow the term ladder of 1.8.9 without the penalised spline.
Each is a natural cubic spline of 3 degrees of freedom with 4 or more distinct
values in the analysis data, a factor with 2 or 3, and absent with 1.

The trial term ${f}$ adjusts for differences between trials, while one
treatment coefficient is shared across trials (Danaei et al. 2013; Caniglia et
al. 2023). Section 1.8.9 states why the model reads the trial and the interval
start, and not the calendar period of the row or the stop of the row. No
confounders enter the outcome model, so $\exp(\beta_{1})$ is the marginal IRR. The model is fit by weighted quasi-Poisson regression with
survey-linearised variance (1.8.6).

With rare events in each interval, as is typical in register data, the IRR
approximates the hazard ratio of a proportional-hazards model (Thompson 1977).
The Poisson working model stays feasible on data with millions of person-trial
intervals, where weighted Cox regression would not. The quasi-Poisson
variance function allows for overdispersion, including that from the weights.
Each IRR comes with weighted event counts, person-years at 52.25 weeks per
year, and rates per 100,000 person-years.

### 1.8.5 Absolute scale

The survival of arm ${a}$ is a weighted product-limit estimate on the distinct
stop times of the rows. Every stop is a whole number of weeks, so the estimate
moves on a weekly grid:

$$
\hat S_{a}(t) = \prod_{u \le t} \{1 - \hat h_{a}(u)\},
\qquad
\hat h_{a}(u) = \frac{\sum W_{i,m,j} \, Y_{i,m,j}}{\sum W_{i,m,j}}.
$$

Both sums run over the person-trials of arm ${a}$. The numerator holds the
events at the stop of their own row. The denominator holds the weight of every
row at risk at ${u}$, so the risk set spans the stop time. A row is at risk at
${u}$ when $t_{\text{start}} < u \le t_{\text{stop}}$.

$W_{i,m,j}$ is the analysis weight of 1.8.3. It is the truncated treatment
weight for the intention-to-treat analogue and the truncated product weight
for the per-protocol estimand. Covariates do not enter the estimator, so the
weights carry the whole adjustment, as in the IRR model. A stop time at which
an arm has nobody at risk leaves the survival of that arm unchanged. The risk
difference is $\hat S_{0}(t) - \hat S_{1}(t)$.

### 1.8.6 Inference

Standard errors are survey-linearised (a Huber–White sandwich) and clustered
on the person, not the person-trial. That accounts for the repeated
person-trials and the repeated intervals of one person (Hernán et al. 2008;
Danaei et al. 2013; Su et al. 2024). The interval of the IRR is a Wald
interval on the log scale, $\exp(\hat\beta_{1} \pm z\,\widehat{se})$. Here
${z}$ is the two-sided normal critical value at the study's confidence level,
95% unless the specification sets another, so ${z}$ is 1.96 at 95%. Two
caveats apply:

- The variance treats the estimated weights, the single imputation and the
  carry-forward as fixed. For stabilised weights this is usually slightly
  conservative for the treatment coefficient, but it is not exact. A
  person-level bootstrap of the whole pipeline, with every model refitted, is
  the fuller alternative.
- In the simulations of section 3, coverage is near nominal where the
  assumptions of the estimand hold. It is slightly below nominal under
  confounding with independent loss. When an estimand ignores informative
  loss, coverage falls because of bias, not because of the variance
  estimator.

The interval of the risk difference is a percentile interval from a
person-level (cluster) bootstrap, with 500 replicates and a fixed seed. Each
replicate draws ${N}$ persons with replacement, and a drawn person brings all
of their person-trials. One resample serves both arms. A person can be a
comparator in an early trial and in the intervention arm of a later one. The
survival estimates of the two arms are then correlated, and separate
resamples per arm would ignore that covariance.

The weights keep their estimated values in every replicate: the models of
1.8.1 and 1.8.2 are not refitted. The level of the interval is the study's
confidence level, the same level as the interval of the IRR.

At a stop time where either arm has no weighted event up to and including
that time, the risk difference has no interval. Every replicate that is not
missing would then give that arm a risk of exactly zero, and the percentiles
would reflect the other arm alone. The point estimate is still reported.

The interval of the number needed to treat is
$(-1/\mathrm{RD}_{\mathrm{lo}}, -1/\mathrm{RD}_{\mathrm{hi}})$. It exists only
when the risk-difference interval strictly excludes zero, because
${x} \mapsto -1/{x}$ is undefined at zero. Otherwise the number needed to treat
has no interval. Altman (1998) reports such an interval as running from an
NNTH through infinity to an NNTB, and this implementation does not. The sign
of the point estimate of the risk difference alone decides benefit or harm.

### 1.8.7 Missing data

Each kind of missing value has one rule:

- A missing confounder value at the recruiting week is singly imputed. The
  default draws one hot-deck value from the observed values, under a fixed
  seed.
- A missing confounder value during follow-up is carried forward from the
  last observed value of the same person-trial. The carry-forward starts from
  the value at the recruiting week. A person-trial with no observed value after entry
  therefore carries its entry value, imputed or not, through follow-up. swereg
  reports the filled rows and person-trials per confounder and per
  enrollment.
- A missing treatment value during follow-up is discordant in both arms
  (1.4).
- A missing treatment value in the enrollment period drops that week from the
  arm classification (1.3).
- A missing subgroup value removes the row from the analyses of that subgroup
  (1.8.8).

A missing treatment value also interacts with the washouts. One washout rule
excludes a week when an earlier week in its window holds the washed-out value.
Under that rule, a missing week makes every later week ineligible while the
window holds it. Under a lifetime washout of that kind, one missing week
removes the person from every later trial. The other washout rule excludes a
week when an earlier observed week holds another value, and it skips missing
weeks.

A week without observation is not a missing value. It is a loss of
observation, and it stops follow-up under both estimands (1.4). Neither the imputation nor the carry-forward
propagates its uncertainty into the variance (1.8.6).

### 1.8.8 Heterogeneity and subgroups

The protocol names each subgroup as a categorical baseline variable. For each
subgroup, the pipeline fits rate ratios within each level. It also runs a
joint Wald test of the interaction between treatment and the subgroup. Both
run for both estimands. A level with no events returns no estimate rather than
an unstable one.

The subgroup IRRs and the effect-modification test hold the time terms of the
IRR model (1.8.9). A joint Wald test of the interaction between treatment and
a natural spline of the trial tests heterogeneity across trials. The spline has
3 degrees of freedom, or one fewer than the number of trials where that is
smaller. swereg provides that test, and the pipeline does not run it. A study that wants it states it in the
protocol.

### 1.8.9 Time terms

Every time term of 1.8 reads one of three time axes. Each row of the analysis
data carries all three. This subsection names them by their columns, because
the reasons for each model's terms depend on what each column holds.

| Column | Symbol | Meaning |
|:--|:--|:--|
| `period_id` | $c_{m,j}$ | The calendar period of the row. |
| `enrollment_period_id` | ${m}$ | The trial. It is the enrollment period, which ends at time zero, and it is constant within a person-trial. |
| `tstart` | $u_{m,j}$ | The time since time zero at the start of the interval, in weeks. |

Follow-up interval ${j}$ of trial ${m}$ is interval number ${j}+1$. With an
enrollment period of ${w}$ weeks, the three axes obey one identity.

$$
c_{m,j} = m + (j + 1),
\qquad
u_{m,j} = j\,w .
$$

In column terms, `period_id` = `enrollment_period_id` + interval number, and
the interval number is `tstart` / ${w}$ + 1. Figure A shows the three axes on a
Lexis diagram.

```{r time-term-setup, include = FALSE}
tt_theme <- theme_classic(base_size = 11) +
  theme(
    axis.line = element_line(colour = "black"),
    axis.ticks = element_line(colour = "black"),
    axis.text = element_text(colour = "black"),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
tt_col <- c(
  "period_id" = "#0072B2",
  "enrollment_period_id" = "#8a1c1c",
  "tstart" = "#E69F00",
  "not in the model" = "grey80"
)
```

```{r time-term-lexis, fig.width = 7, fig.height = 5.2, fig.cap = "Figure A. Each line is one person-trial from its time zero, and the black lines are one person in three successive trials. The dashed lines are one calendar period, one trial and one time since time zero."}
w <- 4
pt <- data.table(
  who = c(rep("other person-trials", 2), rep("one person, three successive trials", 3), rep("other person-trials", 3)),
  enrollment_period_id = c(0, 1, 2, 3, 4, 5, 6, 1),
  n_intervals = c(6, 4, 1, 1, 5, 3, 4, 7)
)
pt[, `:=`(
  x = enrollment_period_id + 1,
  xend = enrollment_period_id + 1 + n_intervals,
  yend = n_intervals * w
)]
pt[, who := factor(who, levels = c("one person, three successive trials", "other person-trials"))]
ggplot(pt) +
  geom_vline(xintercept = 7, colour = tt_col[["period_id"]], linetype = "dashed") +
  geom_abline(slope = w, intercept = 0, colour = tt_col[["enrollment_period_id"]], linetype = "dashed") +
  geom_hline(yintercept = 12, colour = tt_col[["tstart"]], linetype = "dashed") +
  geom_segment(aes(x = x, y = 0, xend = xend, yend = yend, colour = who, linewidth = who)) +
  annotate("label", x = 7, y = 27, label = "vertical: one period_id", colour = tt_col[["period_id"]], size = 3.2, hjust = 0) +
  annotate("label", x = 4.7, y = 20, label = "diagonal: one enrollment_period_id", colour = tt_col[["enrollment_period_id"]], size = 3.2, hjust = 1) +
  annotate("label", x = 9.3, y = 12, label = "horizontal: one tstart", colour = tt_col[["tstart"]], size = 3.2, vjust = -0.3) +
  scale_colour_manual(NULL, values = c("black", "grey55")) +
  scale_linewidth_manual(NULL, values = c(1.3, 0.6)) +
  scale_x_continuous("Calendar period (period_id)", breaks = 0:12) +
  scale_y_continuous("Time since time zero (weeks)", breaks = seq(0, 28, 4)) +
  coord_fixed(ratio = 1 / w, xlim = c(0, 12), ylim = c(0, 28), expand = FALSE) +
  labs(title = "Three time axes on one Lexis diagram") +
  tt_theme
```

The models of 1.8 hold the time terms below. Here flex(${x}$) is the term
ladder below, and ${n}$ is the number of
trials.

| Model | Time terms | Other terms | Offset | Family | Engine |
|:--|:--|:--|:--|:--|:--|
| Treatment weights (1.8.1) | none | confounders at the recruiting week | none | binomial, logit link | logistic regression |
| Censoring denominator (1.8.2) | flex(`tstart`) + flex(`period_id`) | time-updated confounders | log person-weeks | binomial, cloglog link | penalised GAM, or GLM (1.9) |
| Censoring numerator (1.8.2) | flex(`tstart`) | none | log person-weeks | binomial, cloglog link | as the denominator |
| IRR and subgroup IRR (1.8.4, 1.8.8) | flex(`tstart`) + flex(`enrollment_period_id`) | arm | log person-weeks | quasi-Poisson | survey-weighted GLM |
| Effect modification (1.8.8) | flex(`tstart`) + flex(`enrollment_period_id`) | arm × factor(subgroup) | log person-weeks | quasi-Poisson | survey-weighted GLM |
| Heterogeneity across trials (1.8.8) | flex(`tstart`) | arm × ns(`enrollment_period_id`, df = min(3, ${n}$ − 1)) | log person-weeks | quasi-Poisson | survey-weighted GLM |

The person-weeks of the offset are the width of the row after clipping (1.8.4).

```{r time-term-models, fig.width = 8, fig.height = 3, fig.cap = "Figure B. Each panel is the Lexis diagram of Figure A. Coloured lines are the two axes that the model holds, and grey lines are the axis it leaves out."}
w <- 4
lexis_panel <- function(title, used, x_title = NULL, y_title = NULL, note = NULL) {
  axis <- c("period_id", "enrollment_period_id", "tstart")
  role <- ifelse(axis %in% used, axis, "not in the model")
  names(role) <- axis
  # One segment layer for the three families of lines, so that every panel
  # draws the same legend and patchwork can collect it.
  seg <- rbind(
    data.table(x = 1:11, y = 0, xend = 1:11, yend = 28, role = role[["period_id"]]),
    data.table(x = 0, y = seq(4, 24, w), xend = 12, yend = seq(4, 24, w), role = role[["tstart"]]),
    data.table(x = 0:11, y = 0, xend = 0:11 + 7, yend = 28, role = role[["enrollment_period_id"]])
  )
  seg <- seg[order(role != "not in the model")]
  ggplot(seg) +
    geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = role), linewidth = 0.6, show.legend = TRUE) +
    scale_colour_manual(NULL, values = tt_col, limits = names(tt_col), guide = guide_legend(nrow = 1)) +
    {
      if (is.null(note)) NULL else annotate("label", x = 6, y = 26, label = note, size = 3.2, colour = tt_col[["enrollment_period_id"]])
    } +
    scale_x_continuous(x_title, breaks = seq(0, 12, 4)) +
    scale_y_continuous(y_title, breaks = seq(0, 28, 8)) +
    coord_fixed(ratio = 1 / w, xlim = c(0, 12), ylim = c(0, 28), expand = FALSE) +
    labs(title = title) +
    tt_theme +
    theme(plot.title = element_text(size = 10))
}
patchwork::wrap_plots(
  # The shared axis titles are drawn once: y on the left panel, x on the middle.
  lexis_panel("Censoring denominator", c("period_id", "tstart"), y_title = "Weeks since time zero"),
  lexis_panel("Outcome model", c("enrollment_period_id", "tstart"), x_title = "Calendar period"),
  lexis_panel("Heterogeneity test", c("enrollment_period_id", "tstart"), note = "treatment × trial"),
  nrow = 1
) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

#### At most two axes in each model

The outcome models and the censoring denominator each hold two of the three
axes. The censoring numerator holds `tstart` only, and the treatment-weight
model holds no time term. The identity makes the third axis a linear function
of the other two. This is the age–period–cohort identity of
Lexis-diagram analysis. A linear term in the third axis therefore adds
nothing to a model that holds the other two. Only the curvature of the third
axis can add to the fit, and the data identify that curvature weakly.

A grid of 10 trials and 8 follow-up intervals shows this. Natural cubic
splines of 3 degrees of freedom in `enrollment_period_id` and `tstart`, with an
intercept, give a design matrix of rank 7. A linear `period_id` term leaves the
rank at 7. A natural cubic spline of 3 degrees of freedom in `period_id` raises
it to 9.

#### Outcome models: the trial and the time since time zero

The outcome models hold the trial and the time since time zero, as the
sequential trial literature does. Danaei et al. (2013, p. 76) pooled 83 trials
into one model. That model included "'month at the trial's baseline' (taking
values from 1 to 83) and month of follow-up in each 'trial'". Each entered as
a continuous covariate with its squared term. To test heterogeneity, they added "a product
term between the indicator for therapy initiation and the month of the
'trial'" (p. 76). The heterogeneity test of 1.8.8 is that test, with a natural
spline of the trial in place of the linear product term.

Caniglia et al. (2023, author manuscript p. 5) pooled 13 trials and included
"'trial' (taking values from 1-13 and modeled with restricted cubic splines)".
Their outcome model is a log-binomial model of a risk at delivery, so it holds
no follow-up time. TrialEmulation (Su et al. 2024, p. 17) adds `trial_period`
and `followup_time` to the outcome model by default, each as a linear and a
quadratic term. There, `trial_period` is the trial index and `followup_time`
is "the follow-up visit number within trials" (p. 16).

The trial term adjusts for differences in the event rate between trials. One
treatment coefficient is then shared by every trial.

#### Censoring denominator: the calendar period and the time since time zero

The censoring denominator holds the calendar period of the row and the time
since time zero. Deviation and loss respond to the time on the protocol, for
example a stop early in treatment. They also respond to calendar time, for
example a change in prescribing practice. Such a change acts on every open
trial in the same calendar period, so the calendar period is its axis.

Danaei et al. (2013, p. 79) used inverse probability weights against the
selection that the artificial censoring causes. Their model for treatment use
included "calendar month of follow-up and its squared term". The same model
also included the baseline calendar year.

#### Censoring numerator: the time since time zero only

The numerator model holds the time since time zero and nothing else. Su et al.
(2024, p. 6–7, the text after equation 3) state the condition. Covariates
"conditioned upon in the numerator terms of the stabilised weights must be
included as covariates in the marginal structural model". The outcome model
holds the trial and the time since time zero. A spline in `period_id` is not
in the span of those two terms, as the rank example above shows. A spline or
factor in `period_id` can fall outside that span, as the rank example shows,
so the numerator omits it.

By default the numerator fits in each arm, so it also conditions on the arm,
which the outcome model holds. A numerator that depends only on terms of the
outcome model leaves the estimand unchanged. The cost of the smaller numerator
is therefore more variable weights, and not bias.

#### Time read at the start of the interval

Every model with a time-since-entry term reads it at the interval start,
`tstart`, and not at the interval stop, `tstop`. The last interval of a person-trial is
cut short at an event or a censoring, so its `tstop` is smaller than a whole
interval. A flexible term in `tstop` could then fit those last rows apart from
the other rows, and the last rows hold the events. Under `tstart`, every row of
interval number ${k}$ sits at $({k} - 1)\,{w}$ weeks.

Danaei et al. (2013, p. 76) index follow-up by month of follow-up.
TrialEmulation indexes it by `followup_time`, the visit number (Su et al.
2024, p. 16). Both are interval numbers. The clipped width still enters the
offset through the person-weeks. Figure C shows two person-trials under both
readings.

```{r time-term-intervals, fig.width = 7, fig.height = 3.4, fig.cap = "Figure C. Bars are the rows of two person-trials, and the red row ends at an event in week 10. Dots place each row at tstart, and crosses place it at tstop."}
rows <- data.table(
  trial = c(rep("person-trial A", 4), rep("person-trial B", 3)),
  tstart = c(0, 4, 8, 12, 0, 4, 8),
  tstop = c(4, 8, 12, 16, 4, 8, 10),
  row = c(rep("whole interval", 6), "event, interval cut short")
)
rows[, y := as.numeric(factor(trial, levels = c("person-trial B", "person-trial A")))]
pts <- rbind(
  rows[, .(y = y + 0.2, time = tstart, reading = "time read at tstart")],
  rows[, .(y = y - 0.2, time = tstop, reading = "time read at tstop")]
)
ggplot(rows) +
  geom_vline(xintercept = seq(0, 16, 4), colour = "grey85") +
  geom_rect(aes(xmin = tstart, xmax = tstop, ymin = y - 0.08, ymax = y + 0.08, fill = row), colour = "black") +
  geom_point(aes(x = time, y = y, shape = reading), data = pts, size = 2.6) +
  scale_fill_manual(NULL, values = c("whole interval" = "grey70", "event, interval cut short" = "#8a1c1c")) +
  scale_shape_manual(NULL, values = c("time read at tstart" = 16, "time read at tstop" = 4)) +
  scale_x_continuous("Weeks since time zero", breaks = seq(0, 16, 4)) +
  scale_y_continuous(NULL, breaks = 1:2, labels = c("person-trial B", "person-trial A"), limits = c(0.6, 2.4)) +
  labs(title = "Where each row sits on the time axis") +
  tt_theme +
  theme(legend.box = "vertical")
```

#### The term ladder

flex(${x}$) depends on the number of distinct values of ${x}$ in the rows the
model fits.

- 10 or more, in a censoring model fit as a penalised GAM: a penalised spline.
  It asks for 10 basis functions, mgcv's default, so it needs 10 distinct
  values.
- 4 or more otherwise: a natural cubic spline of 3 degrees of freedom. With the
  intercept it has 4 parameters, so it needs 4 distinct values.
- 2 or 3: a factor, which needs 2.
- 1: no term.

The penalised spline fits only inside the GAM engine. Every outcome model and
the GLM censoring model of 1.9 therefore start the ladder at the natural
spline. A censoring model counts the distinct values in each stratum it fits,
by default each arm, after the rows of zero width leave. An outcome model
counts them in the rows it fits, so a subgroup IRR counts the rows of its
subgroup level.

#### Column names and stored objects

Earlier releases named the calendar-period column `trial_id`. On a follow-up
row that column held the calendar period of the row, and not the trial. The
name hid that the outcome model read the calendar period. swereg now names each
column by what it holds: `period_id` for the calendar period of a row, and
`enrollment_period_id` for the trial.

A saved R6 object runs the method bodies it was saved with. A stored
enrollment or plan from an earlier release would fit the earlier time terms
with no warning. swereg therefore refuses to read a stored enrollment below
schema 6 and a stored plan below schema 5. Rebuild the plan with s0, then run
s1, s2 and s3 again.

## 1.9 Sensitivity analyses

TARGET item 7h.ii.

The pipeline reports one sensitivity analysis by default: the per-protocol
IRR with untruncated weights. It is a required companion of the primary
analysis, not an alternative primary analysis. A material divergence between
the two estimates means the weights are under stress. It calls for two checks:
the raw weight distribution, and the positivity of treatment and censoring.
Where the extreme weights are structural, restrict the eligible population
rather than truncate harder.

The censoring model can also be refit as a generalised linear model, which
drops the penalised splines. Each time term of 1.8.2 is then a natural cubic
spline of 3 degrees of freedom with 4 or more distinct values, a factor with 2
or 3, and absent with 1. The numerator keeps the follow-up term alone (1.8.9).

## 1.10 Limitations

TARGET item 16 and RECORD item 19.1 (Benchimol et al. 2015).

- No grace periods and no cloning. Deviation censors at the right edge of the
  first discordant run that exceeds the tolerance of the arm (1.4).
- The estimand conditions on reaching time zero. It says nothing about people
  who die, or have an outcome, before time zero (1.6).
- Trials open once per enrollment period, so a person who becomes eligible
  inside an enrollment period waits for the next time zero.
- The censoring model carries no lagged treatment term, so it holds no
  adherence history.
- There is no as-treated estimand.
- The absolute risk is cause-specific. Death censors follow-up, and no
  competing-risk cumulative incidence is estimated (1.6).
- The bootstrap for the risk difference holds the weights fixed, so its
  interval leaves out the uncertainty of the weight models (1.8.6).
- The single imputation and the carry-forward do not propagate their
  uncertainty into the variance (1.8.7).
- Comparator sampling discards comparator information, which costs precision
  (1.3).
- The propensity model holds main effects only. The censoring model holds main
  effects and time terms, which are smooth only with enough distinct values
  (1.8.2). Non-linearities need derived variables.

Each study adds its own limitations, such as misclassification and
unmeasured confounding in the registers.

---

# 2. Manuscript methods

Replace each placeholder in the text below with the value of the study.

| Placeholder | Value |
|---|---|
| `[intervention]` | the intervention treatment |
| `[comparator]` | the comparator strategy |
| `[outcome]` | the outcome |
| `[confounders]` | the confounders |
| `[width]` | the width of the enrollment period, in weeks |
| `[washout]` | the new-user washout window |
| `[k]` | the number of comparators drawn per intervention person-trial, for example 2 |
| `[horizon]` | the analysis horizon |
| `[end]` | the administrative end of study |
| `[tolerance]` | the tolerance of each arm, in weeks |
| `[subgroups]` | the pre-specified subgroups |

We emulated a sequence of target trials of `[intervention]` versus
`[comparator]` for `[outcome]` (Hernán et al. 2008; Hernán and Robins 2016;
Cashin et al. 2025). The data came from the Swedish national health
registers.

## Eligibility, treatment strategies and assignment

Eligible individuals could start treatment at many calendar times. We
therefore emulated a sequence of trials rather than a single trial (Hernán et
al. 2008; Danaei et al. 2013; Caniglia et al. 2023). A new trial opened every `[width]`
weeks, and its enrollment period lasted `[width]` weeks. We assessed every
eligibility criterion in every week. Look-back windows counted calendar weeks,
not data rows. Each enrollment could add its own criteria to the global
criteria, including a range of calendar years for trial entry.

A new-user criterion excluded any week with use of `[intervention]` in the
`[washout]` before it. With a lifetime washout, each person started treatment
in at most one trial. With a finite washout, a person could start again in a
later trial after `[washout]` without treatment.

Assignment used the weeks of the enrollment period in which the individual was
eligible and held one of the two arm values. Individuals entered the
intervention arm if at least one of those weeks held `[intervention]`. They
entered the comparator arm if all of those weeks held `[comparator]`.
Individuals with no such week were ineligible for that trial.

All intervention person-trials were enrolled. To bound computation, we drew
comparators by `[k]`:1 incidence density sampling within each trial, with a
seed stated in the specification. The draw took `[k]` times the number of
intervention person-trials of that trial, or every remaining comparator where
fewer remained. It was stratified by the enrollment period and read no other
variable. It formed no matched sets, and no later step conditioned on one.

## Follow-up and outcomes

Time zero was a landmark: the first week after the enrollment period closed. A
person entered the trial only if they reached that week under observation and
free of every enrollment outcome. The enrollment period therefore contributed
no follow-up and no immortal time (Hernán and Robins 2016; Caniglia et al.
2023).

Follow-up ended at the earliest of `[outcome]`, loss of observation, the
administrative end of study at `[end]` and the horizon of `[horizon]`. Loss of
observation was the first week after time zero without observation, whether a
gap or the end of the record. No tolerance applied to it, and an outcome in
that week or later was not counted. Every stop was exact to the week. The risk
was cause-specific: death and the end of observation censored follow-up, and
we did not model them as competing risks.

Per-protocol follow-up also ended at protocol deviation: a run of consecutive
weeks off the assigned strategy longer than `[tolerance]` weeks. A missing
treatment status counted as a week off the strategy. When `[outcome]` fell in
the same follow-up interval as an earlier deviation, the person-trial ended
with the outcome. A deviation did not end intention-to-treat follow-up.

## Causal contrasts and identifying assumptions

We estimated two estimands (Danaei et al. 2013). Both conditioned on reaching
the time zero of each trial under observation and free of the study outcomes.
The observational analogue of the intention-to-treat effect compared the arms
as assigned, ignoring later changes in treatment. The per-protocol effect
compared sustained `[intervention]` with sustained `[comparator]`.

We reported both estimands as marginal incidence rate ratios (IRRs), with
weighted event counts and rates per 100,000 person-years by arm. We also
reported the risk difference at the end of follow-up. Beside it, we reported
the number needed to treat for benefit (NNTB) or for harm (NNTH) (Altman
1998).

Identification relied on consistency, positivity, and no unmeasured
confounding of assignment given `[confounders]`. The per-protocol effect also
relied on the censoring model capturing the common causes of deviation, loss
and the outcome.

## Statistical analysis

We adjusted for `[confounders]` by stabilised inverse probability of treatment
weights from a logistic model (Hernán et al. 2008). The model read each
confounder at the recruiting week, the first eligible week of the enrollment
period with an arm value. For the per-protocol effect, we added stabilised
inverse probability of censoring weights from discrete-time censoring models,
fitted separately by arm. These models included the most recent values of
`[confounders]`, the time since time zero and the calendar period of follow-up
(Danaei et al. 2013). Each time term was a penalised spline where enough
distinct values allowed one, and simpler otherwise. The numerator models
included the time since time zero only.

We truncated the weights at the 1st and 99th percentiles (Danaei et al.
2013). We fitted a weighted quasi-Poisson marginal structural model of the
event indicator on the assigned arm, with log person-time as the offset. It
included terms for the trial and for the time since time zero at the start of
each interval (Danaei et al. 2013; Su et al. 2024). Each was a natural spline
of 3 degrees of freedom with 4 or more distinct values, and a factor with 2 or
3. The exponentiated coefficient of the arm estimated the marginal IRR, pooled
across trials.

With rare events, the IRR approximated the marginal hazard ratio (Thompson
1977). Individuals contributed several observations within and across trials.
Confidence intervals therefore used survey-linearised (sandwich) standard
errors clustered on the person (Hernán et al. 2008; Danaei et al. 2013).

For the absolute scale, we estimated the survival of each arm with a weighted
product-limit estimator on the weekly grid of stop times. It used the same
weights as the IRR. The risk difference was the difference between the
arms in one minus survival. Its confidence interval, at the pre-specified
level, was the percentile interval of 500 bootstrap replicates. Each replicate
resampled persons, not person-trials, and one resample served both arms. The
weights were held fixed in the bootstrap.

No interval was reported at a time by which either arm had no weighted event.
The number needed to treat was the negative reciprocal of the risk difference,
so a positive value meant benefit. We reported its interval only when the
interval of the risk difference excluded zero.

A missing confounder value at the recruiting week was singly imputed by a
hot-deck draw. A missing value during follow-up was carried forward within the
person-trial. We assessed effect modification by `[subgroups]` with Wald tests
of the interaction between treatment and subgroup. We also estimated the IRR
within each subgroup level.

## Sensitivity analyses

We repeated the per-protocol analysis with untruncated weights, as a
pre-specified sensitivity analysis. We read a divergence between the two
estimates as a sign of unstable weights.

## Software

Analyses used R with the swereg package, which provided the sequential
enrollment, weighting and estimation described above. The censoring models
were fitted with mgcv, and the weighted outcome regression with
`survey::svyglm()`. The implementation was validated against simulated data
with known true effects, and against the TrialEmulation package (Su et al.
2024). The validation suite of the package ran in continuous integration.

---

# 3. Validation evidence

All numerical results in this section, in the prose as well as in the
tables and figures, are computed directly from a results artifact rather
than transcribed by hand. The artifact is produced by rerunning the
complete validation battery: the same data-generating processes, truth
calculations, and fit wrappers that the package's test suite enforces in
continuous integration. Section 4.2 maps each layer to its test file and
describes how to regenerate the artifact.

Provenance: generated `r ev$meta$generated_utc` with swereg
`r ev$meta$swereg`, TrialEmulation `r ev$meta$trialemulation`, under
`r ev$meta$r`.

## 3.1 Design of the validation battery

The battery is organised around one principle: an estimator is validated by
recovering a truth that is *known by construction*, not by agreeing with
another implementation. Agreement between two packages is used as
corroborating evidence only: two correct implementations of the same
estimand must agree, but two implementations can also agree while both miss
the truth, and the battery deliberately includes a scenario (informative
loss under the ITT estimand) that demonstrates exactly this.

Truth is therefore computed by direct counterfactual simulation. For each
scenario and estimand, 200,000 persons per arm are simulated under the forced
strategy (for the per-protocol truth, treatment is held at the assigned
value in every period; for the ITT truth, only the baseline value is forced
and subsequent treatment follows the scenario's natural switching process),
and the truth is the log ratio of first-event incidence rates, with events
counted until each person's first event and person-time accumulated only
while at risk. Loss to follow-up is never applied to the truth simulation:
loss is a nuisance the estimator must be robust to, not part of the estimand.
This first-event, person-time-at-risk construction matches the estimand
targeted by the weighted quasi-Poisson model (1.8.4) exactly; a
recurrent-event or fixed-denominator construction would target a different
quantity.

Four layers separate concerns, so that a failure localises to a pipeline
segment:

```{r}
layers <- data.table(
  layer = c(
    "Cross-package matrix (3.3)",
    "Stress matrix (3.4)",
    "Plan-layer truth matrix (3.5)",
    "Coverage calibration (3.6)"
  ),
  segment = c(
    "Enrollment-layer estimators (IPW, IPCW, weighted MSM) on person-period panels",
    "The same estimators at design extremes",
    "The complete production pipeline: specification, enrollment periods, sequential eligibility, the comparator draw, worker subprocesses, dual analysis files",
    "The sandwich variance estimator"
  ),
  question = c(
    "Do swereg and TrialEmulation each recover known truth where the estimand's assumptions hold, and fail identically where they do not?",
    "Does the estimator remain stable under rare outcomes, null and harmful effects, near-positivity violation, heavy informative attrition, and treatment-confounder feedback?",
    "Does the pipeline as a whole recover a planted constant-hazard truth, including the separation of PP from ITT under discontinuation?",
    "Do nominal 95% intervals cover the truth 95% of the time when the estimand is valid?"
  )
)
knitr::kable(
  layers,
  col.names = c("Layer", "Pipeline segment exercised", "Question answered"),
  align = "lll",
  caption = "Table 1. The four layers of the validation battery."
)
```

Interpreting single-dataset cells requires one calibration: at the sample
sizes used here, one simulated dataset carries Monte Carlo noise of roughly
0.03–0.05 on the log-IRR scale, so a single-run gap of that order is
indistinguishable from zero. All cells run at fixed seeds and are therefore
exactly reproducible; the multi-replicate cells (Tables 5, 8, and 13–17)
quantify bias and coverage across repeated draws, free of this caveat.

## 3.2 Enrollment-layer scenarios: data-generating processes

The enrollment-layer cells (3.3, 3.4) share one person-period
data-generating process. For person $i$ with standard-normal baseline
confounder $L_{0i}$ and periods $t = 0, \ldots, 19$:

$$
\begin{aligned}
\mathrm{logit}\,\mathrm{Pr}(A_{i0} = 1)
  &= -0.3 + \phi_A L_{0i}
  && \text{(baseline initiation)} \\
\mathrm{logit}\,\mathrm{Pr}(A_{it} = 1)
  &= -3.0 + \phi_S L_{0i} + 8\,A_{i,t-1}, \quad t \ge 1
  && \text{(switching with persistence)} \\
\mathrm{logit}\,\mathrm{Pr}(Y_{it} = 1)
  &= -3.5 + \theta A_{it} + \phi_Y L_{0i}
  && \text{(outcome)}
\end{aligned}
$$

with true contemporaneous treatment effect $\theta = -0.7$ unless a cell
varies it. The persistence coefficient of 8 keeps most initiators on
treatment (adherent person-time dominates) while still generating enough
switching to separate the PP and ITT truths. Loss to follow-up, when
present, is geometric dropout from a per-person hazard: constant at 0.06 per
period (independent loss), or $\mathrm{expit}(-1.4 + 0.9\,L_{0i})$
(informative loss, so that dropout selects on the confounder that also
drives treatment and outcome). The three standard scenarios switch the
nuisance parameters only, leaving the true effect identical:

```{r}
dgp <- data.table(
  scenario = c("s1", "s2", "s3"),
  phi_a = c("0", "0.6", "0.6"),
  phi_s = c("0", "0.4", "0.4"),
  phi_y = c("0", "0.4", "0.4"),
  loss = c("none", "independent (hazard 0.06)", "informative (expit(-1.4 + 0.9 L0))"),
  induces = c(
    "clean benchmark: no confounding, no selection",
    "baseline confounding plus outcome-independent attrition",
    "baseline confounding plus attrition that selects on the confounder"
  )
)
knitr::kable(
  dgp,
  col.names = c(
    "Scenario", "$\\phi_A$", "$\\phi_S$", "$\\phi_Y$",
    "Loss to follow-up", "What it induces"
  ),
  align = "lccccl",
  caption = "Table 2. Nuisance configuration of the three standard scenarios."
)
```

Each scenario dataset is simulated at 20,000 persons and 20 periods with a
fixed seed. Table 3 reports the realized characteristics of the exact
datasets analysed in 3.3.

```{r}
td <- copy(ev$triangle_desc)
knitr::kable(
  td[, .(
    scenario,
    n_persons = format(n_persons, big.mark = ","),
    person_periods = format(person_periods, big.mark = ","),
    pct_lost = sprintf("%.0f%%", pct_periods_lost),
    pct_init = sprintf("%.1f%%", pct_initiators),
    first_events = format(first_event_persons, big.mark = ","),
    risk = sprintf("%.2f%%", event_risk_follow_up_interval_pct)
  )],
  col.names = c(
    "Scenario", "Persons", "Person-periods", "Person-periods lost",
    "Initiators at baseline", "Persons with ≥1 event", "Event risk per period"
  ),
  align = "lrrrrrr",
  caption = "Table 3. Realized descriptives of the three scenario datasets."
)
```

## 3.3 Cross-package validation matrix

Each scenario dataset is fed through the full triangle (known
potential-outcome truth, swereg, and `TrialEmulation`) for both estimands.
Estimates are compared on a common rate-ratio scale: `TrialEmulation` reports
odds ratios from pooled logistic regression, which are converted with the
Zhang–Yu relation using the reference arm's per-period event risk from the
truth simulation (Zhang and Yu 1998; Section 3.7). `TrialEmulation` is a
peer required to recover the truth itself, not an oracle.

```{r}
tri <- copy(ev$triangle)
tri[, nuisance := fcase(
  scenario == "s1", "none",
  scenario == "s2", "confounding + independent loss",
  scenario == "s3", "confounding + informative loss"
)]
tri[, `:=`(
  sw = ci3(sw_est, sw_lo, sw_hi),
  te = ci3(te_est, te_lo, te_hi),
  bias_sw = sg3(sw_est - truth),
  bias_te = sg3(te_est - truth),
  gap = sg3(sw_est - te_est)
)]
knitr::kable(
  tri[, .(
    scenario, nuisance, estimand,
    truth = sprintf("%.3f", truth), sw, te, bias_sw, bias_te, gap
  )],
  col.names = c(
    "Scenario", "Nuisances", "Estimand", "True log-IRR",
    "swereg [95% CI]", "TrialEmulation [95% CI]",
    "swereg bias", "TE bias", "swereg − TE"
  ),
  align = "lllrrrrrr",
  caption = paste0(
    "Table 4. Cross-package validation matrix (N = 20,000, T = 20 periods, ",
    "fixed seed). swereg estimates use the primary truncated weights. ",
    "Log-IRR scale."
  )
)
x31 <- list(
  max_ok = tri[scenario != "s3" | estimand != "itt", max(abs(sw_est - truth))],
  itt_sw = tri[scenario == "s3" & estimand == "itt", sg3(sw_est - truth)],
  itt_te = tri[scenario == "s3" & estimand == "itt", sg3(te_est - truth)],
  itt_gap = tri[scenario == "s3" & estimand == "itt", sprintf("%.3f", abs(sw_est - te_est))]
)
```

In Table 4, every interval in a cell whose assumptions hold covers the
truth; the per-protocol estimator remains close to the truth in s3 precisely
because its censoring weights (1.8.2) model the informative loss; and in the
s3 ITT cell both packages miss on the same side (swereg `r x31$itt_sw`,
TrialEmulation `r x31$itt_te`) while agreeing with each other to within
`r x31$itt_gap`. A single dataset nonetheless provides limited evidence: at
N = 20,000 one estimate carries Monte Carlo noise of roughly 0.03–0.05 on
the log-IRR scale, so point estimates deviate visibly from the truth even
under a perfectly unbiased estimator, and the size of any one gap is
determined by sampling variation alone. A stronger assessment is obtained
by replication: the full
triangle is repeated on 20 independent datasets per scenario, which reduces
the Monte Carlo standard error of the estimated bias by a factor of
$\sqrt{20}$.

```{r}
tr_reps <- copy(ev$triangle_reps)
tr5 <- tr_reps[, .(
  truth = first(truth),
  R = .N,
  sw_bias = mean(sw_est - truth),
  sw_se = sd(sw_est - truth) / sqrt(.N),
  swu_bias = mean(sw_est_untrunc - truth),
  swu_se = sd(sw_est_untrunc - truth) / sqrt(.N),
  te_bias = mean(te_est - truth),
  te_se = sd(te_est - truth) / sqrt(.N),
  gap = mean(abs(sw_est - te_est))
), by = .(scenario, estimand)]
knitr::kable(
  tr5[, .(
    scenario, estimand, R,
    truth = sprintf("%.3f", truth),
    sw = sprintf("%+.3f (%.3f)", sw_bias, sw_se),
    swu = sprintf("%+.3f (%.3f)", swu_bias, swu_se),
    te = sprintf("%+.3f (%.3f)", te_bias, te_se),
    gap = sprintf("%.3f", gap)
  )],
  col.names = c(
    "Scenario", "Estimand", "Datasets", "True log-IRR",
    "swereg, truncated weights: mean bias (MC SE)",
    "swereg, untruncated weights: mean bias (MC SE)",
    "TrialEmulation: mean bias (MC SE)", "Mean |swereg − TE|"
  ),
  align = "llrrrrrr",
  caption = paste0(
    "Table 5. Replicated cross-package matrix: mean bias over 20 independent ",
    "datasets per scenario (N = 20,000 each), with the Monte Carlo standard ",
    "error of the mean. swereg is shown with its primary (1st/99th ",
    "percentile) weight truncation and with untruncated weights. ",
    "Log-IRR scale."
  )
)
mo <- tr5[scenario != "s3" | estimand != "itt", max(abs(sw_bias))]
x33r <- list(
  s1_max = tr5[scenario == "s1", sprintf("%.3f", max(abs(sw_bias)))],
  max_ok = sprintf("%.3f", mo),
  rel_ok = sprintf("%.0f%%", 100 * (exp(mo) - 1)),
  max_se = tr5[, sprintf("%.3f", max(sw_se))],
  s3_bias = tr5[scenario == "s3" & estimand == "itt", sprintf("%+.3f", sw_bias)],
  s3_z = tr5[scenario == "s3" & estimand == "itt", sprintf("%.0f", abs(sw_bias) / sw_se)],
  s3pp = tr5[scenario == "s3" & estimand == "pp", sprintf("%+.3f", sw_bias)],
  s3pp_u = tr5[scenario == "s3" & estimand == "pp", sprintf("%+.3f (MC SE %.3f)", swu_bias, swu_se)],
  s2pp = tr5[scenario == "s2" & estimand == "pp", sprintf("%+.3f", sw_bias)]
)
```

```{r, fig.width = 7, fig.height = 5, fig.cap = "Figure 1. Bias of the estimated log-IRR over 20 independent datasets per scenario (N = 20,000 each) in the scenarios whose assumptions every fit satisfies: s1 (no confounding, no loss) and s2 (confounding, outcome-independent loss). Faint points are individual datasets; solid points are the mean bias with its 95% Monte Carlo interval; the vertical line marks zero bias. All three fits (swereg with truncated weights, swereg with untruncated weights, and TrialEmulation) are centred on zero in s1 and in the per-protocol cells; the small displacement in the s2 intention-to-treat cell is shared by every fit (largest for TrialEmulation) and is the person-time-weighting residual discussed in the text, not a property of any one implementation. The informative-loss scenarios, where the fits genuinely differ, are shown in Figure 2."}
pd <- rbind(
  tr_reps[, .(scenario, estimand, rep, fit = "swereg, truncated weights", bias = sw_est - truth)],
  tr_reps[, .(scenario, estimand, rep, fit = "swereg, untruncated weights", bias = sw_est_untrunc - truth)],
  tr_reps[, .(scenario, estimand, rep, fit = "TrialEmulation", bias = te_est - truth)]
)[scenario != "s3"]
pd[, scenario_lab := factor(scen_lab(scenario), scen_lab(c("s1", "s2")))]
# each fit on its own row; truncated at the top of every panel (discrete y draws
# the first level at the bottom, so reverse the level order)
pd[, fit := factor(fit, rev(names(col_fit)))]
pm <- pd[, .(mean_bias = mean(bias), se = sd(bias) / sqrt(.N)),
  by = .(scenario, scenario_lab, estimand, fit)]
xr <- range(pd$bias)
f1_panel <- function(est, subtitle, xlab) {
  q <- ggplot(pd[estimand == est], aes(x = bias, y = fit, colour = fit))
  q <- q + geom_vline(xintercept = 0, colour = "grey30")
  q <- q + geom_point(
    size = 0.8, alpha = 0.25,
    position = position_jitter(height = 0.18, width = 0, seed = 1)
  )
  q <- q + geom_pointrange(
    data = pm[estimand == est],
    aes(x = mean_bias, xmin = mean_bias - 1.96 * se, xmax = mean_bias + 1.96 * se),
    size = 0.4, linewidth = 0.9
  )
  q <- q + facet_wrap(~scenario_lab, nrow = 1)
  q <- q + scale_colour_manual(values = col_fit, guide = "none")
  q <- q + coord_cartesian(xlim = xr)
  q <- q + labs(x = xlab, y = NULL, subtitle = subtitle)
  q <- q + theme_minimal()
  q
}
patchwork::wrap_plots(
  f1_panel("itt", "Intention-to-treat", NULL),
  f1_panel("pp", "Per-protocol", "Bias of the log-IRR (estimate − truth)"),
  ncol = 1
)
```

```{r, fig.width = 7.5, fig.height = 8, fig.cap = "Figure 2. Bias under informative loss to follow-up, the scenarios in which the estimators differ materially: s3 and its one-parameter variants (designs in Table 16, Section 3.8; 20 datasets for s3, 10 per variant). Top panel: for the intention-to-treat estimand the informative loss violates the estimand's own assumptions, and every fit is displaced; no estimation method corrects an invalid estimand. Bottom panel: for the per-protocol estimand the fits differ by how they correct the selection. TrialEmulation conditions on the baseline covariate that drives the loss, which is exact in these designs; swereg corrects the selection by censoring weights, and clipping them (truncation) adds an attenuation that grows with the informativeness of the loss; the ordering reverses in the reversed-selection and harmful-effect cells, and no fit has uniformly smaller bias (Section 3.8)."}
il_lab <- function(x) {
  fcase(
    x == "s3", "s3: informative loss (0.9·L0)",
    x == "mild", "mild (0.45·L0)",
    x == "harsh", "harsh (1.5·L0)",
    x == "reversed", "reversed (−0.9·L0)",
    x == "harmful", "harmful effect (+0.7)",
    x == "unmeasured_loss", "unmeasured loss driver (0.9·U)",
    x == "unmeasured_adherence", "unmeasured adherence (healthy-adherer)"
  )
}
tr3 <- ev$triangle_reps[scenario == "s3"]
pd <- rbind(
  tr3[, .(cell = "s3", estimand, rep, fit = "swereg, truncated weights", bias = sw_est - truth)],
  tr3[, .(cell = "s3", estimand, rep, fit = "swereg, untruncated weights", bias = sw_est_untrunc - truth)],
  tr3[, .(cell = "s3", estimand, rep, fit = "TrialEmulation", bias = te_est - truth)],
  ev$trunc_grid[cell != "base", .(cell, estimand = "pp", rep, fit = "swereg, truncated weights", bias = b_sw_trunc)],
  ev$trunc_grid[cell != "base", .(cell, estimand = "pp", rep, fit = "swereg, untruncated weights", bias = b_sw_untrunc)],
  ev$trunc_grid[cell != "base", .(cell, estimand = "pp", rep, fit = "TrialEmulation", bias = b_te)]
)
cell_order <- c("s3", "mild", "harsh", "reversed", "harmful", "unmeasured_loss", "unmeasured_adherence")
pd[, cell_lab := factor(il_lab(cell), il_lab(cell_order))]
# each fit on its own row; truncated at the top of every panel
pd[, fit := factor(fit, rev(names(col_fit)))]
pm <- pd[, .(mean_bias = mean(bias), se = sd(bias) / sqrt(.N)),
  by = .(cell, cell_lab, estimand, fit)]
xr <- range(pd$bias)
f2_panel <- function(est, subtitle, xlab, ncol) {
  q <- ggplot(pd[estimand == est], aes(x = bias, y = fit, colour = fit))
  q <- q + geom_vline(xintercept = 0, colour = "grey30")
  q <- q + geom_point(
    size = 0.8, alpha = 0.25,
    position = position_jitter(height = 0.18, width = 0, seed = 1)
  )
  q <- q + geom_pointrange(
    data = pm[estimand == est],
    aes(x = mean_bias, xmin = mean_bias - 1.96 * se, xmax = mean_bias + 1.96 * se),
    size = 0.4, linewidth = 0.9
  )
  q <- q + facet_wrap(~cell_lab, ncol = ncol)
  q <- q + scale_colour_manual(values = col_fit, guide = "none")
  q <- q + coord_cartesian(xlim = xr)
  q <- q + labs(x = xlab, y = NULL, subtitle = subtitle)
  q <- q + theme_minimal()
  q
}
# ITT has only s3, so give it a single half-width cell (top-left) that lines up
# with the left column of the two-column per-protocol grid below, rather than
# stretching one panel across the full width.
patchwork::wrap_plots(
  f2_panel("itt", "Intention-to-treat (s3 only)", NULL, 1),
  f2_panel("pp", "Per-protocol", "Bias of the log-IRR (estimate − truth)", 2),
  design = "A#\nBB\nBB\nBB\nBB"
)
```

Averaging over 20 datasets reduces the Monte Carlo standard error of the
estimated bias to roughly `r x33r$max_se`, fine enough to resolve
systematic effects that no single dataset can. Three magnitudes emerge from
Table 5 and Figures 1 and 2. In the clean scenario (s1) both packages are unbiased
within Monte Carlo error for both estimands (mean bias of swereg's
truncated-weight fit at most `r x33r$s1_max` in absolute value): the
estimation machinery itself introduces no bias. Where nuisances are present but the estimand remains
valid (s2, and the per-protocol estimand in s3), small systematic residuals
of up to `r x33r$max_ok` become resolvable in the truncated-weight fits:
at most a `r x33r$rel_ok` relative error on the rate-ratio scale, well
inside the tolerances the test suite enforces, and reported here
explicitly.

The truncated-versus-untruncated contrast localises the largest of these
residuals. In the s3 per-protocol cell, the truncated-weight mean bias of
`r x33r$s3pp` falls to `r x33r$s3pp_u` when the same replicates are refit
with untruncated weights: most of the displacement is attributable to
clipping the weights, the bias–variance tradeoff described in 1.8.3. Informative
dropout means the high-risk individuals still under observation late in
follow-up must carry large censoring weights to represent those who left;
the 1st/99th-percentile truncation caps precisely those weights, and the
under-corrected selection surfaces as bias toward the null. This is also
why the untruncated per-protocol results are exported as a sensitivity
analysis: a material divergence between the truncated and untruncated
estimates indicates that truncation is attenuating the correction. The event-priority convention (1.4) is excluded as a cause by
design contrast: the s2 per-protocol cell shares the identical switching,
censoring, and event-accounting machinery, differs only in that its loss is
non-informative, and shows no bias with the same truncated weights
(`r x33r$s2pp`). The remaining small
residuals (for example in the s2 ITT cell) are consistent with loss
truncating follow-up toward its early follow-up intervals, so that the
person-time-weighted working-model summary (1.6) no longer weights
follow-up exactly as the no-loss truth functional does.

The s3 ITT cell is of a different kind: truncated-weight mean bias
`r x33r$s3_bias`, roughly
`r x33r$s3_z` Monte Carlo standard errors from zero and present in both
packages. The displacement is one that replication sharpens rather than removes,
because the ITT estimand carries no loss weight and informative loss
therefore biases it in any correct implementation. Cross-package agreement
is not evidence of correctness, which is why every layer of this battery is
anchored to simulated truth.

## 3.4 Stress matrix

The stress cells reuse the Section 3.2 data-generating process with one or
two parameters pushed to an extreme, so that each cell probes a specific
failure mode. Table 5 specifies the designs; the cells then follow in order.

```{r}
sd_tab <- data.table(
  cell = c(
    "Rare outcome", "Null effect", "Informative attrition",
    "Harmful effect, depletion", "Near-positivity violation",
    "Treatment-confounder feedback", "Determinism"
  ),
  design = c(
    "Outcome intercept −6.0 (≈0.2% risk/period), N = 40,000, θ = −0.7",
    "θ = 0, independent loss, N = 20,000",
    "Dropout hazard expit(−1.3 + 0.9 L0): ≈73% of person-periods lost, selecting on the confounder; N = 30,000",
    "θ = +0.7, three independent seeds at N = 20,000, TrialEmulation cross-check",
    "φA = 1.5: propensity scores span 0–1; ITT fit at three truncation levels",
    "AR(1) confounder Lt = 0.7 Lt−1 − 0.4 At−1 + εt driving both switching (0.8 Lt) and outcome (0.5 Lt); N = 25,000",
    "Identical data, PP estimator fit twice"
  ),
  probes = c(
    "Sparse-event stability of the weighted MSM and the spline IPCW model",
    "False-positive effects (does the pipeline manufacture signal from noise?)",
    "IPCW under heavy selection; the ITT arm of this cell is expected to fail",
    "The person-time-weighted-average interpretation of the pooled IRR (1.6)",
    "The truncation bias-variance tradeoff (1.8.3)",
    "The residual-bias limit of single-model IPCW under feedback (1.7)",
    "Uncontrolled stochastic steps anywhere in the fit"
  )
)
knitr::kable(
  sd_tab,
  col.names = c("Cell", "Design deviation from the base DGP", "What it probes"),
  align = "lll",
  caption = "Table 6. Stress-cell designs. All other parameters as in Section 3.2; T = 20 periods, fixed seeds."
)
```

```{r}
st <- copy(ev$stress)
st[, covers := fifelse(truth >= lo & truth <= hi, "yes", "no")]
knitr::kable(
  st[, .(
    cell, estimand, truth = sprintf("%.3f", truth),
    est = ci3(est, lo, hi), bias = sg3(est - truth), covers, note
  )],
  col.names = c(
    "Cell", "Estimand", "True log-IRR", "Estimate [95% CI]",
    "Bias", "CI covers truth", "Note"
  ),
  align = "llrrrcl",
  caption = "Table 7. Stress cells, single dataset at fixed seed; swereg estimates use the primary truncated weights. Log-IRR scale."
)
x32a <- list(
  rare_pp = st[cell == "rare_outcome" & estimand == "pp", sg3(est - truth)],
  attr_pp = st[cell == "informative_attrition" & estimand == "pp", sg3(est - truth)],
  attr_itt = st[cell == "informative_attrition" & estimand == "itt", sg3(est - truth)]
)
```

Three observations from Table 7. At an event risk of roughly 0.2% per
period the per-protocol machinery, including the spline-based censoring
model, remains stable (bias `r x32a$rare_pp`). Under a true null the
estimate is small and its interval covers zero: the weighting and pooling
machinery does not manufacture an effect. And in the attrition cell, where
almost three quarters of person-periods are removed by confounder-driven
dropout, the per-protocol estimator stays at the truth
(bias `r x32a$attr_pp`) while the ITT estimator, which by construction
carries no loss weight, is displaced (bias `r x32a$attr_itt`) and its
interval excludes the truth: the designed failure that motivates the
estimand distinction in practice.

Determinism: refitting the per-protocol estimator on identical data
reproduced the estimate to a maximum absolute difference of
`r format(ev$stress_determinism$max_abs_delta)`; the pipeline has no
uncontrolled stochastic step.

```{r}
sh <- copy(ev$stress_harmful)
knitr::kable(
  sh[, .(
    seed, truth = sprintf("%.3f", truth),
    sw = sprintf("%.3f", sw_est), te = sprintf("%.3f", te_est),
    bias_sw = sg3(sw_est - truth), gap = sg3(sw_est - te_est)
  )],
  col.names = c(
    "Seed", "True log-IRR (cumulative-rate)", "swereg", "TrialEmulation",
    "swereg − truth", "swereg − TE"
  ),
  align = "lrrrrr",
  caption = paste0(
    "Table 8. Harmful effect (true log-IRR > 0) with strong depletion of ",
    "susceptibles, ITT with truncated weights, three seeds. Log-IRR scale."
  )
)
x32 <- list(
  mb = sh[, sg3(mean(sw_est - truth))],
  mg = sh[, sprintf("%.3f", max(abs(sw_est - te_est)))]
)
```

Under a harmful effect with strong depletion of susceptibles, the marginal
hazard ratio declines over follow-up, so the single pooled IRR, a
person-time-weighted average (1.6), legitimately lies above the
cumulative-rate truth, by a mean of `r x32$mb` across the three seeds. This
is a property of the estimand, not an implementation defect: swereg and
`TrialEmulation` agree to within `r x32$mg` on every seed because both
target the same weighted-average summary. Analyses in which the time path
of the effect matters should report follow-up-specific estimates.

```{r}
tc <- copy(ev$stress_trunc)
knitr::kable(
  tc[, .(
    trunc_percentiles, truth = sprintf("%.3f", truth),
    est = sprintf("%.3f", est), bias = sg3(est - truth),
    wmax = sprintf("%.0f", wmax_raw)
  )],
  col.names = c(
    "Truncation percentiles (%)", "True log-IRR", "Estimate",
    "Bias (attenuation)", "Max raw stabilised weight"
  ),
  align = "lrrrr",
  caption = paste0(
    "Table 9. Near-positivity violation: attenuation toward the null grows ",
    "monotonically with truncation severity. ITT, log-IRR scale."
  )
)
```

Table 9 quantifies the tradeoff stated in 1.8.3 on a design whose propensity
scores approach the boundary (maximum raw stabilised weight
`r tc[1, sprintf("%.0f", wmax_raw)]`): each tightening of the truncation
percentiles reduces variance at the cost of measurable bias toward the
null.
When extreme weights are structural rather than sporadic, the appropriate
response is to restrict the eligible population, not to truncate harder.

```{r}
tv <- copy(ev$stress_tv)
knitr::kable(
  tv[, .(
    fit, truth = sprintf("%.3f", truth), est = ci3(est, lo, hi),
    bias = sg3(est - truth), abs_bias = sprintf("%.3f", abs(est - truth))
  )],
  col.names = c(
    "Fit", "True log-IRR", "Estimate [95% CI]", "Bias", "|Bias|"
  ),
  align = "lrrrr",
  caption = paste0(
    "Table 10. Treatment–confounder feedback: per-protocol bias with the ",
    "censoring covariate time-updated versus frozen at baseline; ITT for ",
    "reference. All fits use the primary truncated weights. Log-IRR scale."
  )
)
x32tv <- list(
  up = tv[1, sprintf("%.3f", abs(est - truth))],
  fr = tv[2, sprintf("%.3f", abs(est - truth))],
  itt = tv[3, sg3(est - truth)]
)
```

The feedback cell delineates the limit of the per-protocol estimator's validity. When a
time-varying confounder is itself affected by treatment and drives both
adherence and the outcome, IPCW with time-updated covariates is strictly
less biased than IPCW with covariates frozen at baseline (|bias|
`r x32tv$up` versus `r x32tv$fr`), but a residual bias remains, part of
which is the working-model average under a time-ramping effect rather than
selection per se. The ITT estimand, which needs no censoring model against
this feedback, is near-unbiased in the same data (bias `r x32tv$itt`).
Where treatment–confounder feedback is central to the question, g-methods
beyond this pipeline are indicated, exactly as stated in 1.7.

## 3.5 Full-pipeline truth recovery (plan layer)

The layers above validate the estimators on pre-built person-period panels.
This layer validates everything that sits on top in production: the
machine-readable specification, enrollment-period assignment, sequential
eligibility with a lifetime new-user exclusion, the 2:1 comparator draw in
each enrollment period, the worker subprocess chain, the dual PP/ITT analysis files, and
the pooled weighted outcome model.

The data-generating process plants an exactly known truth in a realistic
skeleton. Persons are observed weekly from 2016-01-01 to 2021-06-30,
roughly 287 ISO weeks, deliberately spanning 2020's 53-week ISO year, and
are split into never-treaters and initiators; initiators start treatment at
an enrollment period drawn uniformly from the first 56 four-week enrollment
periods and, in the
discontinuation cell, stop after a geometric duration (4% weekly hazard).
The weekly outcome hazard is constant at 0.0025 untreated and doubled while
treated, so the marginal per-week incidence rate ratio among sustained
users is exactly 2.0. Scenario B adds a binary frailty carried by 30% of
persons that doubles both the initiation probability and the outcome
hazard, a genuine baseline confounder; mixture-averaging over the two
risk groups with first-event depletion attenuates the marginal truth to
1.982. Loss, when present, is geometric (2% weekly, or 1%/3% by risk group
for informative loss) and multiplies person-time equally in both arms, so
the truth is unchanged and loss is purely a nuisance the machinery must
tolerate. The ITT truth in the discontinuation cell
(`r sprintf("%.2f", ev$plan[cell == "DISC", truth_itt])`) is simulated
directly as the do(initiate)-versus-do(never) contrast with natural
discontinuation.

```{r}
pl <- copy(ev$plan)
knitr::kable(
  pl[, .(
    cell, scenario, loss,
    n = format(n_persons, big.mark = ","),
    pw = format(sk_person_weeks, big.mark = ","),
    tx_pw = format(sk_treated_person_weeks, big.mark = ","),
    events = format(sk_events_n, big.mark = ",")
  )],
  col.names = c(
    "Cell", "Scenario", "Loss", "Persons", "Person-weeks",
    "Treated person-weeks", "Events"
  ),
  align = "lllrrrr",
  caption = "Table 11. Skeleton descriptives per plan-layer cell (before eligibility and enrollment)."
)
```

```{r}
pl[, covers_pp := fifelse(pp_lo <= truth_pp & truth_pp <= pp_hi, "yes", "no")]
pl[, covers_itt := fifelse(itt_lo <= truth_itt & truth_itt <= itt_hi, "yes", "no")]
knitr::kable(
  pl[, .(
    cell, loss,
    truth_pp = sprintf("%.2f", truth_pp),
    pp = ci2(pp_irr, pp_lo, pp_hi), covers_pp,
    truth_itt = sprintf("%.2f", truth_itt),
    itt = ci2(itt_irr, itt_lo, itt_hi), covers_itt
  )],
  col.names = c(
    "Cell", "Loss", "PP truth", "PP IRR [95% CI]", "covers",
    "ITT truth", "ITT IRR [95% CI]", "covers"
  ),
  align = "llrrcrrc",
  caption = paste0(
    "Table 12. Plan-layer factorial (A = no confounding, B = baseline ",
    "confounding; each × no/independent/informative loss) plus the ",
    "discontinuation cell. PP and ITT estimates use the primary truncated ",
    "weights. IRR scale."
  )
)
x33 <- list(
  crude = pl[cell == "B_none", sprintf("%.2f", crude_rr)],
  ipw = pl[cell == "B_none", sprintf("%.2f", ipw_rr)],
  truth_b = pl[cell == "B_none", sprintf("%.2f", truth_pp)],
  disc_gap = pl[cell == "DISC", sg3(log(pp_irr) - log(itt_irr))],
  disc_truth_gap = pl[cell == "DISC", sg3(log(truth_pp) - log(truth_itt))]
)
```

Two rows warrant comment. In the confounded no-loss cell (B_none) the
frailty is doing real confounding work: the crude rate ratio in the enrolled
ITT panel is `r x33$crude`, the IPW-weighted rate ratio `r x33$ipw`, against
a marginal truth of `r x33$truth_b`: the weighting removes essentially all
of the planted confounding. In the discontinuation cell the two estimands
separate as designed, with PP − ITT = `r x33$disc_gap` on the log scale
against a true separation of `r x33$disc_truth_gap`: the per-protocol arm
censors at deviation and reweights back to the sustained-treatment truth of
2.0, while the ITT arm retains post-discontinuation person-time and
attenuates toward the do(initiate) truth. This cell also exercises the
event-priority convention (1.4), since events and deviations collide in the
same follow-up interval whenever discontinuers have events in their final
treated follow-up interval.

Because a single pipeline run at a fixed seed cannot distinguish bias from
draw-level noise, the two no-loss scenarios are repeated over eight
independent seeds at 6,000 persons each, rerunning the complete pipeline
per replicate:

```{r}
mc <- copy(ev$plan_mc)
mc[, covers_pp := fifelse(pp_lo <= truth & truth <= pp_hi, "yes", "no")]
mc[, covers_itt := fifelse(itt_lo <= truth & truth <= itt_hi, "yes", "no")]
knitr::kable(
  mc[, .(
    scenario, seed,
    truth = sprintf("%.2f", truth),
    pp = ci2(pp_irr, pp_lo, pp_hi), covers_pp,
    itt = ci2(itt_irr, itt_lo, itt_hi), covers_itt
  )],
  col.names = c(
    "Scenario", "Seed", "Truth", "PP IRR [95% CI]", "covers",
    "ITT IRR [95% CI]", "covers"
  ),
  align = "llrrcrc",
  caption = "Table 13. Plan-layer Monte Carlo, per replicate; truncated (primary) weights. IRR scale."
)
mcs <- rbind(
  mc[, .(
    estimand = "pp",
    mean_bias = mean(log(pp_irr) - log(truth)),
    mc_sd = sd(log(pp_irr)),
    covered = sum(pp_lo <= truth & truth <= pp_hi),
    n_seeds = .N
  ), by = scenario],
  mc[, .(
    estimand = "itt",
    mean_bias = mean(log(itt_irr) - log(truth)),
    mc_sd = sd(log(itt_irr)),
    covered = sum(itt_lo <= truth & truth <= itt_hi),
    n_seeds = .N
  ), by = scenario]
)
setorder(mcs, scenario, -estimand)
knitr::kable(
  mcs[, .(
    scenario, estimand, mean_bias = sg3(mean_bias),
    mc_sd = sprintf("%.3f", mc_sd),
    coverage = sprintf("%d/%d", covered, n_seeds)
  )],
  col.names = c(
    "Scenario", "Estimand", "Mean log bias", "MC sd", "95% CI coverage"
  ),
  align = "llrrr",
  caption = "Table 14. Plan-layer Monte Carlo, summarised over the eight seeds; truncated (primary) weights. Log-IRR scale."
)
```

The mean log-scale bias is within Monte Carlo error of zero in both
scenarios, and coverage is consistent with the nominal 95% at eight
replicates. Individual misses, visible in Table 13, are the expected
behaviour of honest intervals, not smoothed away.

## 3.6 Coverage calibration

The final layer asks whether the reported uncertainty can be trusted: over
200 replicate draws per scenario at 3,000 persons, each refit end to end,
what fraction of nominal 95% intervals cover the truth? The study uses the
per-protocol estimand estimated with the primary truncated weight (the
pipeline's default analysis exactly as reported). The per-protocol censoring
weights (1.8.2) target the sustained-treatment effect in all three scenarios,
including the informative loss in s3, so the three scenarios test whether the
interval calibration survives the same nuisance that biases the
intention-to-treat estimand.

```{r}
cv <- copy(ev$coverage)[estimand == "pp"]
cv[, nuisance := fcase(
  scenario == "s1", "none",
  scenario == "s2", "confounding + independent loss",
  scenario == "s3", "confounding + informative loss"
)]
knitr::kable(
  cv[, .(
    scenario, nuisance,
    reps = sprintf("%d/%d", n_fit, M),
    mean_bias = sg3(mc_mean_bias), mc_sd = sprintf("%.3f", mc_sd),
    coverage = sprintf("%d/%d (%.1f%%)", covered_n, n_fit, 100 * coverage)
  )],
  col.names = c(
    "Scenario", "Nuisances", "Replicates fit", "Mean log bias",
    "MC sd", "95% CI coverage"
  ),
  align = "llrrrr",
  caption = paste0(
    "Table 15. Per-protocol coverage calibration, M = 200 replicates per ",
    "scenario at N = 3,000, using the primary truncated per-protocol weight, ",
    "that is, the coverage of the pipeline's default per-protocol analysis as ",
    "reported. Log-IRR scale."
  )
)
x34 <- list(
  c1 = cv[scenario == "s1", sprintf("%.1f%%", 100 * coverage)],
  c2 = cv[scenario == "s2", sprintf("%.1f%%", 100 * coverage)],
  c3 = cv[scenario == "s3", sprintf("%.1f%%", 100 * coverage)],
  b3 = cv[scenario == "s3", sg3(mc_mean_bias)]
)
```

```{r, fig.width = 7, fig.height = 6, fig.cap = "Figure 3. Coverage calibration: all 200 replicate 95% confidence intervals per scenario (per-protocol estimand, primary truncated weights), sorted by point estimate, against the true log-IRR (horizontal line). Intervals that miss the truth are drawn in red. Because the per-protocol censoring weights correct the informative loss, the interval cloud straddles the truth in every scenario (s1, s2, and s3 alike), with only the sampling-expected few percent of misses and none of the wholesale downward displacement the intention-to-treat estimand shows under the same loss."}
pd <- copy(ev$coverage_reps)[estimand == "pp"]
pd[, scenario_lab := scen_lab(scenario)]
setorder(pd, scenario, est)
pd[, idx := seq_len(.N), by = scenario]
q <- ggplot(pd, aes(x = idx))
q <- q + geom_linerange(aes(ymin = lo, ymax = hi, colour = covered), linewidth = 0.35)
q <- q + geom_point(aes(y = est), size = 0.25, colour = "grey20")
q <- q + geom_hline(aes(yintercept = truth), linewidth = 0.4)
q <- q + facet_wrap(~scenario_lab, ncol = 1)
q <- q + scale_colour_manual(values = c(`TRUE` = "grey65", `FALSE` = "#C00000"), guide = "none")
q <- q + labs(x = "Replicate (sorted by point estimate)", y = "log-IRR, 95% CI")
q <- q + theme_minimal()
q
```

Across all three scenarios the per-protocol interval stays close to nominal:
`r x34$c1` in s1, `r x34$c2` under confounding with independent loss (s2),
and `r x34$c3` under informative loss (s3). The censoring weights (1.8.2) remove
the selection that informative loss induces, so the s3 point estimate carries
only `r x34$b3` mean bias (Table 15), so intervals of the correct width cover
the truth rather than missing systematically as the estimate distribution
shifts away from it. The mild departures in s1 and s2 are the expected
consequence of treating estimated weights as fixed (1.8.6). This is the payoff of being specific about
the estimand: under the same informative loss the intention-to-treat interval
degrades, because no variance estimator can repair a point estimate the
estimand itself leaves biased. The per-protocol interval, built
around an unbiased estimate, remains calibrated.

## 3.7 Marginal versus conditional estimands

swereg and `TrialEmulation` both remove baseline confounding, but by
different routes, producing two distinct and each valid estimands. swereg
weights and fits a covariate-free model: a marginal effect. `TrialEmulation`
conventionally adjusts the outcome model: a conditional effect. Rate ratios
are collapsible, so these coincide for the IRR; odds ratios are not, so the
`TrialEmulation` OR is converted with the Zhang–Yu relation
$\mathrm{RR} = \mathrm{OR} / (1 - p_0 + p_0\,\mathrm{OR})$, where $p_0$ is
the reference-arm per-period risk, before comparison. The conversion removes
the scale gap only; a residual conditional-versus-marginal difference
remains, visible in Table 4 as the small swereg − TE gaps in the confounded
scenarios (larger for ITT than for PP). The primary correctness guarantee is
each implementation's agreement with the known simulated truth on its own
scale. Section 3.8 measures where each route's advantage holds, and where
both end.

## 3.8 Boundary of validity: the truncation tradeoff across scenarios

The s3 per-protocol cell raised two questions a single scenario cannot
answer: is the conditional-adjustment route always the better one, and is
truncation always a cost? This section varies the design one knob at a time
around the s3 configuration: the strength of the loss's dependence on the
confounder (0.45, 0.9, 1.5 on $L_0$), its direction (−0.9, so that dropout
selects low-risk rather than high-risk person-time), and the direction of
the treatment effect (harmful, $+0.7$), together with two mechanisms in
which the selection is driven by an unmeasured prognostic factor $U$
(dropout on $U$, and a healthy-adherer mechanism in which treated
individuals with high $U$ discontinue preferentially), plus a separate
data-generating process in which censoring is driven by a time-varying
covariate that treatment itself affects. Per-protocol estimand throughout; ten paired
replicates per cell.

```{r}
tg <- copy(ev$trunc_grid)
tg_lab <- function(cl) {
  fcase(
    cl == "mild", "informative loss, mild (0.45·L0)",
    cl == "base", "informative loss, base (0.9·L0, = s3)",
    cl == "harsh", "informative loss, harsh (1.5·L0)",
    cl == "reversed", "informative loss, reversed (−0.9·L0)",
    cl == "harmful", "harmful effect (+0.7), informative loss",
    cl == "unmeasured_loss", "unmeasured loss driver (0.9·U)",
    cl == "unmeasured_adherence", "unmeasured adherence driver (healthy-adherer)"
  )
}
t16 <- tg[, .(
  R = .N,
  lost = sprintf("%.0f%%", mean(pct_lost)),
  truth = sprintf("%.3f", first(truth)),
  sw_t = sprintf("%+.3f (%.3f)", mean(b_sw_trunc), sd(b_sw_trunc) / sqrt(.N)),
  sw_u = sprintf("%+.3f (%.3f)", mean(b_sw_untrunc), sd(b_sw_untrunc) / sqrt(.N)),
  te = sprintf("%+.3f (%.3f)", mean(b_te), sd(b_te) / sqrt(.N))
), by = cell]
t16[, cell := tg_lab(cell)]
x16u <- list(
  loss_t = tg[cell == "unmeasured_loss", sprintf("%+.3f", mean(b_sw_trunc))],
  loss_u = tg[cell == "unmeasured_loss", sprintf("%+.3f", mean(b_sw_untrunc))],
  loss_te = tg[cell == "unmeasured_loss", sprintf("%+.3f", mean(b_te))],
  adh_t = tg[cell == "unmeasured_adherence", sprintf("%+.3f", mean(b_sw_trunc))],
  adh_u = tg[cell == "unmeasured_adherence", sprintf("%+.3f", mean(b_sw_untrunc))],
  adh_te = tg[cell == "unmeasured_adherence", sprintf("%+.3f", mean(b_te))]
)
knitr::kable(
  t16,
  col.names = c(
    "Cell", "Datasets", "Person-periods lost", "True log-IRR",
    "swereg truncated: mean bias (MC SE)",
    "swereg untruncated: mean bias (MC SE)",
    "TrialEmulation: mean bias (MC SE)"
  ),
  align = "lrrrrrr",
  caption = paste0(
    "Table 16. Truncation-tradeoff grid, per-protocol estimand: one design ",
    "parameter at a time around the s3 configuration, plus two cells in ",
    "which selection is driven by an unmeasured prognostic factor U ",
    "(N = 20,000 per dataset). Log-IRR scale."
  )
)
```

Some features of the s3 result generalise; the ranking does not. The
dose–response is systematic: swereg's truncated-weight bias grows
monotonically with the informativeness of the loss, and at the harshest
setting even the untruncated fit degrades, because under extreme selection
the censoring weights become difficult to estimate. The weighting route's
difficulty is therefore continuous in selection strength rather than a
truncation artifact alone. The conditional route
(`TrialEmulation`, no censoring weights, baseline covariate in the outcome
model) is unaffected across the dose–response, but only because this loss
is driven exactly by the covariate it conditions on. No uniform ranking
generalises. With the selection reversed, the
residual changes sign, partially cancels the truncation shift, and
truncated swereg lands nearest the truth of all three fits; with a harmful
effect, truncation partially offsets the downward drag that depletion of
susceptibles plus late-follow-up up-weighting produces, and the untruncated
fit is the worst of the three. Neither package, and neither weight variant,
dominates across the grid.

The two unmeasured-driver cells locate the boundary set by assumption (5)
of the analysis plan (1.7). In both, every fit is displaced together and
in the same direction, toward an exaggerated protective effect: with
dropout on the unmeasured factor, swereg truncated `r x16u$loss_t`,
untruncated `r x16u$loss_u`, TrialEmulation `r x16u$loss_te`; with the
unmeasured factor driving adherence, `r x16u$adh_t`, `r x16u$adh_u`, and
`r x16u$adh_te` respectively. No weighting or conditioning on measured
covariates corrects selection on an unobserved variable. Two further
observations. First, the displacement in the unmeasured-loss cell
contradicts the intuition that dropout independent of treatment should
cancel between arms in a ratio: events deplete high-risk person-time
faster in the comparator arm, so an identical dropout process interacts
differently with the two arms' risk sets, and the ratio does not escape.
Second, the truncated and untruncated estimates differ far less in these
cells than in the measured-covariate cells: the truncated-versus-
untruncated divergence responds to weight instability arising from
measured covariates and remains largely silent about unmeasured drivers,
whose detection requires design-based approaches (negative-control
outcomes, sensitivity analyses for unmeasured selection) rather than
weight diagnostics.

```{r}
fb <- copy(ev$feedback_grid)
t17 <- fb[, .(
  R = .N,
  truth = sprintf("%.3f", first(truth)),
  up = sprintf("%+.3f (%.3f)", mean(b_sw_updated), sd(b_sw_updated) / sqrt(.N)),
  fr = sprintf("%+.3f (%.3f)", mean(b_sw_frozen), sd(b_sw_frozen) / sqrt(.N)),
  te = sprintf("%+.3f (%.3f)", mean(b_te_baseline), sd(b_te_baseline) / sqrt(.N))
)]
knitr::kable(
  t17,
  col.names = c(
    "Datasets", "True log-IRR",
    "swereg IPCW, time-updated covariate: mean bias (MC SE)",
    "swereg IPCW, covariate frozen at baseline: mean bias (MC SE)",
    "TrialEmulation, baseline conditioning: mean bias (MC SE)"
  ),
  align = "rrrrr",
  caption = paste0(
    "Table 17. Feedback boundary, per-protocol estimand: censoring driven ",
    "by a time-varying covariate that treatment affects (the 1.7 regime); ",
    "both swereg fits use the truncated (primary) product weight. All approaches fail by ",
    "three to six times the largest bias in Table 16. Log-IRR scale."
  )
)
```

Table 17 is the boundary the SAP declares in 1.7, now measured: when the
determinants of censoring are time-varying and affected by treatment, every
configuration of either package (time-updated censoring weights, frozen
covariates, or baseline conditioning) is biased by an order of magnitude
more than anywhere else in this battery. All three estimates are therefore
unusable, and comparing them identifies only which approach fails least,
not an approach that works. No setting available within this pipeline
(different truncation percentiles, covariate sets, or censoring-model
specifications) repairs the problem, because the difficulty is structural.
When a time-varying confounder is itself affected by earlier treatment,
valid estimation requires methods designed for that feedback, such as the
parametric g-formula or g-estimation of structural nested models (Hernán
and Robins 2016), which this pipeline does not implement.

```{r}
tri_pp <- ev$triangle_reps[estimand == "pp", .(
  cell_lab = scen_lab(scenario),
  b_trunc = sw_est - truth,
  b_untrunc = sw_est_untrunc - truth,
  b_te = te_est - truth
)]
grid_pp <- tg[, .(
  cell_lab = tg_lab(cell),
  b_trunc = b_sw_trunc,
  b_untrunc = b_sw_untrunc,
  b_te = b_te
)]
comb <- rbind(tri_pp, grid_pp)
pdl <- melt(
  comb,
  id.vars = "cell_lab", variable.name = "variant", value.name = "bias"
)
pdl[, variant := fcase(
  variant == "b_trunc", "truncated (primary)",
  variant == "b_untrunc", "untruncated (sensitivity)",
  variant == "b_te", "TrialEmulation (reference)"
)]
sm <- pdl[, .(
  mean_bias = mean(bias),
  se = sd(bias) / sqrt(.N),
  spread = sd(bias),
  rmse = sqrt(mean(bias^2))
), by = .(cell_lab, variant)]
ord <- c(
  scen_lab(c("s1", "s2", "s3")),
  tg_lab(c(
    "mild", "base", "harsh", "reversed", "harmful",
    "unmeasured_loss", "unmeasured_adherence"
  ))
)
sm[, cell_lab := factor(cell_lab, rev(ord))]
```

```{r, fig.width = 7.5, fig.height = 6, fig.cap = "Figure 4. Mean bias of the per-protocol log-IRR across every validation cell (s1–s3: 20 datasets each; the Table 16 grid cells: 10 each), one panel per scenario, with 95% Monte Carlo intervals; the vertical line marks zero. Both swereg weight variants are shown together with TrialEmulation as the conditional-adjustment reference (a different estimation route, not a third weight variant: baseline covariate in the outcome model, no censoring weights, odds ratios converted to the rate-ratio scale). Truncation introduces bias where informative loss makes the censoring weights heavy-tailed (compare the mild, base, and harsh cells), has no measurable effect where the weight distribution is stable (s1, s2), and in the reversed-selection and harmful-effect cells the truncated fit is the least biased of the three."}
s4 <- copy(sm)
s4[, cell_lab := factor(as.character(cell_lab), ord)]
s4[, variant := factor(variant, rev(names(col_var)))]
q <- ggplot(s4, aes(x = mean_bias, y = variant, colour = variant))
q <- q + geom_vline(xintercept = 0, colour = "grey30")
q <- q + geom_pointrange(
  aes(xmin = mean_bias - 1.96 * se, xmax = mean_bias + 1.96 * se),
  size = 0.45, linewidth = 0.9
)
q <- q + facet_wrap(~cell_lab, ncol = 2)
q <- q + scale_colour_manual(values = col_var, guide = "none")
q <- q + labs(x = "Mean bias of the per-protocol log-IRR", y = NULL)
q <- q + theme_minimal()
q
```

```{r, fig.width = 7.5, fig.height = 6, fig.cap = "Figure 5. Spread of the same per-protocol estimates: the standard deviation across replicate datasets, the sampling noise an analyst running one study draws from. One panel per scenario, with bars anchored at zero; TrialEmulation is shown as the conditional-adjustment reference. Truncation reduces this component of error: the truncated fit has the smaller spread of the two swereg variants in every scenario, by up to a factor of three where the weights are most extreme (harmful-effect cell)."}
s5 <- copy(sm)
s5[, cell_lab := factor(as.character(cell_lab), ord)]
s5[, variant := factor(variant, rev(names(col_var)))]
q <- ggplot(s5, aes(x = spread, y = variant, colour = variant))
q <- q + geom_segment(aes(x = 0, xend = spread, yend = variant), linewidth = 0.7)
q <- q + geom_point(size = 1.5)
q <- q + facet_wrap(~cell_lab, ncol = 2)
q <- q + scale_colour_manual(values = col_var, guide = "none")
q <- q + labs(
  x = "Standard deviation of single-dataset per-protocol estimates (log-IRR)",
  y = NULL
)
q <- q + theme_minimal()
q
```

```{r, fig.width = 7.5, fig.height = 6, fig.cap = "Figure 6. Root-mean-squared error, combining bias (Figure 4) and spread (Figure 5) as the two components of the bias–variance tradeoff: the expected error of a single study's per-protocol estimate, and the criterion on which the primary analysis is chosen. One panel per scenario, with bars anchored at zero; TrialEmulation is shown as the conditional-adjustment reference. Of the two swereg variants, the truncated fit has lower or practically equal error in every scenario, with the largest advantage where the untruncated weights are unstable (harmful-effect cell); relative to the reference, no estimation route has uniformly lower error across scenarios."}
s6 <- copy(sm)
s6[, cell_lab := factor(as.character(cell_lab), ord)]
s6[, variant := factor(variant, rev(names(col_var)))]
q <- ggplot(s6, aes(x = rmse, y = variant, colour = variant))
q <- q + geom_segment(aes(x = 0, xend = rmse, yend = variant), linewidth = 0.7)
q <- q + geom_point(size = 1.5)
q <- q + facet_wrap(~cell_lab, ncol = 2)
q <- q + scale_colour_manual(values = col_var, guide = "none")
q <- q + labs(x = "Root-mean-squared error of the per-protocol estimate (log-IRR)", y = NULL)
q <- q + theme_minimal()
q
```

```{r}
x38 <- sm[, .(
  trunc_lower_spread_n = sum(
    spread[variant == "truncated (primary)"] <
      spread[variant == "untruncated (sensitivity)"]
  ),
  trunc_lower_rmse_n = sum(
    rmse[variant == "truncated (primary)"] <
      rmse[variant == "untruncated (sensitivity)"]
  ),
  cells_n = uniqueN(cell_lab)
)]
```

The recommendation follows from Figure 6, not from either ingredient
alone (Figures 4 and 5). Truncation lowers the spread of single-dataset estimates
in `r x38$trunc_lower_spread_n` of the `r x38$cells_n` per-protocol cells,
and has the lower RMSE in `r x38$trunc_lower_rmse_n` of them: neither
variant is uniformly better, and the cells in which each has the lower
error are those its mechanism predicts. The pipeline's convention is therefore
retained on the evidence: the truncated fit is the primary analysis (its
error is stable and bounded across every regime tested), the untruncated
fit is always exported alongside (1.8.3), and a material divergence between
the two indicates that the censoring weights are unstable. The appropriate
responses are then sensitivity analyses at looser truncation percentiles
(Table 9 quantifies the dose–response), restriction of the eligible
population where extreme weights are structural, or, when the censoring
drivers are time-varying and treatment-affected, the recognition that no
weighting scheme in this pipeline suffices (Table 17).

---

# 4. Implementation mapping

Section 1 names no code. This section names the function, argument, option,
column and test file behind each step of Section 1, and the source of the
validation evidence.

## 4.1 SAP step → code

| SAP | Step | Implementation |
|:--|:--|:--|
| Values, 1.1, 1.4 | Width of the enrollment period | `period_width` argument of `tteplan_from_spec_and_registrystudy()`, default `4L`. It also sets the width of each follow-up interval. `vignette("tte-timing")` states the timing rules with worked examples |
| 1.1 | Global and per-enrollment criteria | Global: `inclusion_criteria$isoyears`, `inclusion_criteria$criteria` and `exclusion_criteria`. Per enrollment: `additional_inclusion` (`age_range`, `isoyear_range`, `has_event`, washouts) and `additional_exclusion`, applied after the global criteria by `tteplan_apply_exclusions()`. `tteplan_read_spec()` refuses a second `age_range` in one enrollment, because every `age_range` writes the one column `eligible_age`. It also refuses a second `isoyear_range`, and an `isoyear_range` outside `inclusion_criteria$isoyears` (26.13.0). An `isoyear_range` writes a column of its own, so the age range still applies. Since 26.14.0, TARGET item 6a lists the criteria per enrollment, and the protocol table states an enrollment-level exclusion as "is TRUE" |
| 1.1 | Look-back windows | `window` takes a number of weeks, `"N year"` or `"N years"` (read as 52N weeks), or `lifetime_before_baseline`. A value of 99999 or more, `Inf` included, means lifetime. Since 26.12.0 the window counts calendar ISO weeks from `isoyearweek`, not rows (`any_events_prior_to()`). An exclusion with `window: lifetime_before_and_after_baseline` reads every row of the person, after baseline as well |
| 1.1, 1.8.7 | Washout / new-user exclusion | A washout rule in any of the four rule blocks. `type: no_prior_value` keeps a person-week when no prior week in the window holds `value`. A missing value in the window makes the rule `NA`, and the week is then ineligible. `type: only_prior_value` keeps a person-week when every prior week in the window that holds an observation holds `value`, so it skips missing weeks |
| 1.1 | Prevalent-user warning | `tteplan_validate_spec()` warns when no washout covers the enrollment's intervention level on the weekly rows of the first skeleton batch. A prevalent week is a week at that level after an earlier week of the same person at that level. A washout covers the enrollment when it makes every prevalent week ineligible. swereg skips the check and says so when no skeleton is loaded (`global_max_isoyearweek` supplied). Set `options(swereg.warn_prevalent_user = FALSE)` to silence it |
| 1.2, 1.4 | Arm values and the weekly status | `treatment.implementation`: `variable`, `intervention_value` and `comparator_value`. The pipeline writes `rd_intervention`: `TRUE` at the intervention value, `FALSE` at the comparator value, `NA` otherwise. It also writes `eligible_valid_treatment`, which is `TRUE` where `rd_intervention` is not `NA` |
| 1.3 | Arm classification, recruiting week | `R/tte_enrollment_periods.R`. Each confounder reaches the panel as `.tte_entry__<v>`, its value at the recruiting week. The panel column `enrollment_period_id` names the trial, and `period_id` names the calendar period of each follow-up interval |
| 1.3 | Comparator draw | `TTEPlan$s1_generate_enrollments_and_ipw()`. `comparator_to_intervention_ratio` (required; `tteplan_read_spec()` stops without it) and `seed` (required; `tteplan_read_spec()` stops without it) come from the YAML spec's `treatment.implementation`. The draw takes `round(ratio * n)` comparators, `by = enrollment_period_id` |
| 1.4 | Time-zero checks | `R/tte_landmark_qualify.R`. The attrition steps are `landmark_candidates` ("Candidate person-trials, before the time-zero checks"), `landmark_observed` ("Not under observation at time zero") and `landmark_event_free` ("Event before time zero"). The observation contract is `observed_var` on each enrollment |
| 1.4 | Arm tolerance | `intervention_tolerance_weeks` and `comparator_tolerance_weeks` on each enrollment, each a whole number of at least 0, default 0 |
| 1.4 | Follow-up stop events, event priority | The private `TTEEnrollment` method `s5_prepare_outcome()`, which `$s4_prepare_for_analysis()` calls. The horizon comes from `follow_up`. The administrative end comes from `global_max_isoyearweek` (default: the largest `isoyearweek` of the first skeleton file), passed to `TTEDesign` as `admin_censor_isoyearweek` |
| 1.4 | Loss of observation | `.tte_record_end_boundary()` writes `weeks_to_record_end`, and `.tte_observation_gap_boundary()` writes `weeks_to_observation_gap`, the first absent week after time zero (`.tte_first_gap_week()`). Under both estimands `.tte_gap_record_end()` moves the record end to that week, so the stop row gets `weeks_to_loss` and `censor_this_period = 1`. The event-priority rule clears `.deviation_clip` only, never the gap. `.tte_deviation_boundary()` holds discordant runs only, and ITT sets `weeks_to_protocol_deviation` to `NA`. A panel that `enroll()` built with `observed_var` and that lacks `weeks_to_observation_gap` was enrolled before swereg 26.15.0. Its weekly rows are gone, so the gap cannot be recomputed. swereg then warns once. The warning says that gaps in observation cannot be detected in the panel, and that an outcome after such a gap may be counted. A re-run of s1 removes the limitation. `qs2_read()` gives the warning when it reads such an enrollment from a file, and `.tte_gap_record_end()` gives it for a panel that was not read. An enrollment saved before 26.15.0 runs its saved method bodies and the old rules, so it warns on read only |
| 1.8.1 | Stabilised IPW | `TTEEnrollment$s2_ipw(stabilize = TRUE)`, on the `.tte_entry__<v>` columns |
| 1.8.2 | IPCW censoring model | The private `TTEEnrollment` method `s6_ipcw_pp()`, reached through `$s4_prepare_for_analysis(estimate_ipcw_pp_with_gam = TRUE, estimate_ipcw_pp_separately_by_treatment = TRUE)`. GAM engine `mgcv::bam(..., discrete = TRUE)`; `mgcv::s()` asks for 10 basis functions. Denominator `flex(tstart) + flex(period_id) + confounders`, numerator `flex(tstart)`, each with `offset(log(person_weeks))`. `flex()` is `.tte_time_term(var, n, gam = TRUE)`: `s(var)` with 10 or more distinct values, `splines::ns(var, df = 3)` with 4 to 9, `factor(var)` with 2 or 3, none with 1, counted per fitted stratum after zero-width rows leave. `$ipcw_formulas` records both formulas of each stratum |
| 1.8.3 | Weight truncation | In the pipeline, `TTEEnrollment$s3_truncate_weights(weight_cols = "ipw")` truncates the treatment weight at its defaults `lower = 0.01` and `upper = 0.99`, and writes `ipw_trunc` (ITT). The private `s6_ipcw_pp()` writes `analysis_weight_pp_trunc` (PP product weight) with 0.01 and 0.99 fixed in the code. Untruncated PP results are exported as a sensitivity sheet |
| 1.8.4, 1.8.6 | Outcome model and inference | `TTEEnrollment$irr(weight_col)`: `survey::svydesign(ids = ~person)` and `survey::svyglm(family = quasipoisson())` with `treatment + flex(tstart) + flex(enrollment_period_id) + offset(log(person_weeks))`. `flex()` is `.tte_time_term(var, n, gam = FALSE)`: `splines::ns(var, df = 3)` with 4 or more distinct values in the fitted rows, `factor(var)` with 2 or 3, none with 1. `tstart` is read through `.tte_interval_start()`. The result carries the formula in `attr(, "model_formula")`. The Wald interval uses `qnorm(1 - (1 - conf_level) / 2)`, and `$s3_analyze()` passes the study level `study.implementation.conf_level` |
| 1.6, 1.8.5, 1.8.6 | Risk difference, number needed to treat, bootstrap interval | `TTEEnrollment$risk_difference(weight_col, n_boot, seed, conf_level)`. `$s3_analyze()` runs it on every ETT at 500 replicates and seed 1, on `analysis_weight_pp_trunc` (stored as `rd_pp_trunc` and `rd_curve_pp_trunc`) and on `ipw_trunc` (stored as `rd_itt` and `rd_curve_itt`). The level comes from `study.implementation.conf_level` in the YAML spec, default 0.95. `TTEPlan$get_curves()` returns the stop time in weeks from time zero as the column `follow_up_interval`, and the stored risk-difference rows use the same name |
| 1.8.7 | Missing data | `TTEEnrollment$s1_impute_confounders(seed = 4)`. The method is the plan's `impute_fn`, whose default `tteenrollment_impute_confounders()` draws one hot-deck value. `TTEEnrollment$s1b_fill_followup_confounders()` carries a follow-up value forward, and `tteenrollment_fill_summary()` counts the filled rows |
| 1.8.8 | Subgroups and heterogeneity | `subgroups` in the YAML spec, each with `implementation$variable`. `$s3_analyze()` runs `TTEEnrollment$irr_by_subgroup()` and `TTEEnrollment$effect_modification_test()` for each subgroup and estimand. `$irr_by_subgroup()` and `$effect_modification_test()` take `conf_level`. The effect-modification model is `treatment * factor(subgroup) + flex(tstart) + flex(enrollment_period_id)`. `TTEEnrollment$heterogeneity_test(weight_col)` fits `treatment * splines::ns(enrollment_period_id, df = min(3, n - 1)) + flex(tstart)`, where `n` is the number of trials; `$s3_analyze()` does not call it |
| 1.9 | Sensitivity analyses | The untruncated PP IRR is stored as `irr_pp`. `estimate_ipcw_pp_with_gam = FALSE` in `$s2_generate_analysis_files_and_ipcw_pp()` fits the censoring model as a cloglog GLM, `.tte_time_term(var, n, gam = FALSE)` for `tstart` and `period_id`: `splines::ns(var, df = 3)` with 4 or more distinct values, `factor(var)` with 2 or 3, none with 1 |
| 1.8 | Pre-specification | YAML spec parsed by `tteplan_read_spec()`; full grid run by `TTEPlan$s1_…`/`s2_…`/`s3_analyze()` |

## 4.2 Where the validation numbers come from

The evidence layers in Section 3 are permanent, executable tests:

| Section | Layer | Test file | Gate |
|:--|:--|:--|:--|
| 3.3 | Cross-package triangle | `tests/testthat/test-tte_validation_matrix.R` | runs in CI |
| 3.4 | Stress matrix | `tests/testthat/test-tte_stress_matrix.R` | fast subset in CI; full battery `SWEREG_RUN_STRESS=true` |
| 3.5 | Plan-layer truth matrix | `tests/testthat/test-tteplan_truth_matrix.R` | reduced-N subset in CI; full factorial `SWEREG_RUN_PLAN_MATRIX=true` |
| 3.6 | Coverage calibration | `tests/testthat/test-tte_coverage.R` | opt-in only, `SWEREG_RUN_COVERAGE=true` |

The tables and figures themselves are rendered from
`vignettes/tte-validation-evidence.rds`, which
`dev/generate_validation_evidence.R` writes (in the source repository, not
the installed package). The script uses the same data-generating, truth and
fit helpers as the tests (`tests/testthat/helper-tte_*.R`). No test reads the
`.rds` file, so the vignette numbers and the test assertions can drift apart.
Rerun the script after any estimator change, and commit the refreshed file
with the change.

## References

- Hernán MA, Alonso A, Logan R, Grodstein F, Michels KB, Willett WC, Manson
  JE, Robins JM. Observational studies analyzed like randomized
  experiments: an application to postmenopausal hormone therapy and
  coronary heart disease. *Epidemiology* 2008;19(6):766–779.
  DOI 10.1097/EDE.0b013e3181875e61.
- Hernán MA, Robins JM. Using big data to emulate a target trial when a
  randomized trial is not available. *Am J Epidemiol* 2016;183(8):758–764.
  DOI 10.1093/aje/kwv254.
- Danaei G, García Rodríguez LA, Cantero OF, Logan R, Hernán MA.
  Observational data for comparative effectiveness research: an emulation
  of randomised trials of statins and primary prevention of coronary heart
  disease. *Stat Methods Med Res* 2013;22(1):70–96.
  DOI 10.1177/0962280211403603.
- Altman DG. Confidence intervals for the number needed to treat. *BMJ*
  1998;317(7168):1309–1312. DOI 10.1136/bmj.317.7168.1309.
- Benchimol EI, et al. The REporting of studies Conducted using
  Observational Routinely-collected health Data (RECORD) Statement. *PLoS
  Med* 2015;12(10):e1001885. DOI 10.1371/journal.pmed.1001885.
- Caniglia EC, Zash R, Fennell C, et al. Emulating target trials to avoid
  immortal time bias: an application to antibiotic initiation and preterm
  delivery. *Epidemiology* 2023;34(3):430–438.
  DOI 10.1097/EDE.0000000000001601.
- Dafni U. Landmark analysis at the 25-year landmark point. *Circ Cardiovasc
  Qual Outcomes* 2011;4(3):363–371. DOI 10.1161/CIRCOUTCOMES.110.957951.
- Cashin AG, Hansford HJ, Hernán MA, et al. Transparent Reporting of
  Observational Studies Emulating a Target Trial: the TARGET Statement.
  *JAMA* 2025;334(12):1084–1093. DOI 10.1001/jama.2025.13350. Also *BMJ*
  2025;390:e087179.
- Thompson WA Jr. On the treatment of grouped observations in life
  studies. *Biometrics* 1977;33(3):463–470.
- Su L, Rezvani R, Seaman SR, Starr C, Gravestock I. TrialEmulation: An R
  Package to Emulate Target Trials for Causal Analysis of Observational
  Time-to-event Data. arXiv:2402.12083, 2024.
- Zhang J, Yu KF. What's the relative risk? A method of correcting the odds
  ratio in cohort studies of common outcomes. *JAMA* 1998;280(19):1690–1691.
