---
title: "TTE methods: manuscript and statistical analysis plan text"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TTE methods: manuscript and statistical analysis plan text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(data.table)
library(ggplot2)
ev <- readRDS("tte-validation-evidence.rds")
ci3 <- function(est, lo, hi) sprintf("%.3f [%.3f, %.3f]", est, lo, hi)
ci2 <- function(est, lo, hi) sprintf("%.2f [%.2f, %.2f]", est, lo, hi)
sg3 <- function(x) sprintf("%+.3f", x)
scen_lab <- function(s) {
  fcase(
    s == "s1", "s1: no confounding, no loss",
    s == "s2", "s2: confounding, independent loss",
    s == "s3", "s3: confounding, informative loss"
  )
}
```

This vignette provides drop-in text describing the target trial emulation
(TTE) methodology implemented in the `swereg-TTE` family of functions
(`TTEEnrollment`, `TTEPlan`, and friends), for both the intention-to-treat
(ITT) and the per-protocol (PP) estimand. It has four sections:

- **Manuscript methods** (Section 1) — short, prose-only, suitable for the
  main body of a journal article. Copy, paste, and replace the `treatment` /
  `outcome` / `confounder` placeholders.
- **Statistical analysis plan** (Section 2) — detailed and
  implementation-agnostic: formulas, per-step model specifications,
  conventions, identifying assumptions, and known limitations, written so a
  statistician who has never seen this package could reimplement the
  estimator. Suitable for a pre-registered SAP, a protocol appendix, or a
  methods supplement.
- **Validation evidence** (Section 3) — the numbers: the design of the
  validation battery, the data-generating processes, and tables and figures
  of truth against estimate for every validation cell, rendered from a
  results artifact rather than asserted in prose.
- **Implementation mapping** (Section 4) — the code behind the SAP: which
  function, argument, option, and test file realises each SAP step, plus
  provenance notes on estimator changes.

Every formula and convention in Section 2 describes what the code actually
computes; where the implementation deviates from a canonical reference
construction, the deviation and its rationale are stated explicitly. The
implementation follows the sequential-trial-emulation literature (Hernán and
Robins 2008; Danaei et al. 2013; Hernán and Robins 2016; Caniglia et al.
2023; Cashin et al. 2025).

---

# 1. Manuscript methods

We applied target trial emulation, a framework for analysing observational
data under explicit protocols that mirror a hypothetical randomized trial, to
estimate the effect of `treatment` on `outcome` in the Swedish national health
registries (Hernán and Robins 2008, 2016; Cashin et al. 2025).

## Sequential trials design

Because eligible individuals can initiate treatment at many different calendar
times, we emulated a sequence of target trials rather than a single trial
(Hernán et al. 2008; Danaei et al. 2013; Caniglia et al. 2023). A new trial
opens every `period` weeks of calendar time. At each trial's baseline, all
eligibility criteria are re-evaluated; eligible individuals enter as
initiators (treatment begins in that trial's baseline period) or as
non-initiators (eligible and untreated). A new-user (washout) criterion
requires no use of the study treatment within a pre-specified washout window
before baseline (a fixed window, e.g. two years as in Danaei et al. 2013, or
the entire observable history for a never-user design), so each person
initiates in at most one trial while contributing eligible person-time as a
non-initiator to earlier trials. Anchoring time zero at eligibility and assignment — rather
than at eventual exposure — prevents immortal time bias (Hernán and Robins
2016). To bound computation, `k` non-initiators were sampled per initiator
within each trial; confounding adjustment is by weighting (below), not by
matching on covariates.

## Estimands

We report two estimands (Danaei et al. 2013). The observational analogue of
the intention-to-treat effect compares initiators with non-initiators as
classified at each trial's baseline, ignoring subsequent changes in
treatment. The per-protocol effect is the effect of sustained treatment
versus sustained non-treatment; for this estimand, follow-up is artificially
censored when a participant's treatment status first deviates from the
baseline-assigned strategy. Both are reported as marginal incidence rate
ratios (IRRs), with weighted event counts and rates per 100,000 person-years
by arm.

## Confounding and censoring adjustment

Baseline treatment assignment is not random: we adjusted for measured baseline
confounders (`confounders`) by stabilised inverse probability of treatment
weighting, estimated from a logistic model at each trial's baseline (Hernán
and Robins 2008). For the per-protocol estimand, artificial censoring at
protocol deviation is informative whenever time-varying factors predict both
adherence and the outcome; we therefore additionally applied stabilised
inverse probability of censoring weights from discrete-time censoring models
fit separately by assigned arm, with a smooth function of follow-up time and
the most recently updated covariate values (Hernán and Robins 2008; Danaei et
al. 2013). Weights were truncated at the 1st and 99th percentiles to limit
the influence of extreme values (Danaei et al. 2013).

## Outcome model and inference

We fit a weighted quasi-Poisson marginal structural model of the event
indicator on assigned baseline treatment with log person-time as offset,
including natural splines of follow-up time and of the trial (calendar)
index; the exponentiated treatment coefficient estimates the marginal IRR
pooled across sequential trials. Because individuals contribute repeated
observations within and across trials, confidence intervals use
cluster-robust (sandwich) standard errors clustered on the person (Hernán
and Robins 2008; Danaei et al. 2013). Effect heterogeneity across calendar
time and pre-specified subgroups was assessed by Wald tests of the
corresponding interaction terms.

## Software

Analyses used R with the `swereg` package, which implements the sequential
enrollment, weighting, and estimation pipeline described above;
censoring-weight models were fit with `mgcv`, and the final weighted
regression with cluster-robust variance with `survey::svyglm()`. The
implementation is validated against simulated data with known true effects
and against the `TrialEmulation` package (Su et al. 2024); the validation
suite runs in continuous integration.

---

# 2. Statistical analysis plan

This section specifies the estimators exactly as implemented, in pipeline
order, in implementation-agnostic terms; the mapping from each step to the
software that executes it is given in Section 4. Notation: individuals
$i = 1, \ldots, N$; sequential trials indexed by their calendar baseline band
$m$; follow-up bands within a trial $j = 0, 1, \ldots, K$, each of fixed
width $w$ weeks (four by default). Let $A_{i,m,j}$ indicate being on the
protocol-defined treatment in band $(m,j)$, with $A_{i,m,0}$ the assigned
baseline arm; $L_{i,m,j}$ the confounder vector as most recently updated at
band $(m,j)$, with $L_{i,m,0}$ its baseline value; $Y_{i,m,j}$ the outcome
indicator; and $C_{i,m,j}$ the indicator of artificial censoring (protocol
deviation or loss to follow-up) in band $(m,j)$.

## 2.1 Sequential enrollment, new-user requirement, and matching

Calendar time is partitioned into consecutive bands of width $w$; each band
opens one trial. Within a band, a person's arm is classified as intervention
if the person is on the intervention treatment in at least one eligible week
of that band, and as comparator if the person is on the comparator treatment
throughout; person-bands with treatment status outside the two protocol arms
are ineligible for that band. The band width is a bias–feasibility tradeoff:
coarser bands admit residual within-band immortal time (Caniglia et al.
2023), which shrinks as $w$ decreases.

Eligibility (inclusion windows, exclusion criteria with lifetime or
fixed-width look-back windows) is re-evaluated at every band. The design does
not impose a new-user rule automatically: the incident-user design is
produced by a protocol-specified washout exclusion on the treatment history,
either a finite look-back window (e.g. 104 weeks, the Danaei et al. 2013
convention) or the entire observable history, for a never-user design. A
lifetime washout makes each person eligible to initiate in at most one band
and removes them from later trials; a finite washout additionally allows
re-qualification after sufficient time off treatment. A protocol without any
washout exclusion enrols prevalent users as initiators at every band and
re-enrols discontinuers as comparators, a prevalent-user design that is
rarely the intended estimand; the software warns when a specification omits
the washout (Section 4.1).

Within each band, all intervention person-trials are enrolled and comparators
are randomly downsampled at a fixed matching ratio per initiator (2:1 by
default, with a pre-specified seed). This sampling bounds computation and is
not covariate matching; all confounding adjustment is deferred to the
weights (2.4). Each enrolled person-trial is expanded to $K$ follow-up bands;
within each band, confounders take their first-week value, outcomes are the
within-band maximum, and person-time is the number of observed source weeks
in the band, so that partially observed bands contribute their true
person-time.

## 2.2 Estimands

Both estimands are marginal incidence rate ratios, standardised over the
enrolled trials' baseline covariate distribution through the weights:

- *Intention-to-treat analogue*: the contrast of initiating versus not
  initiating at baseline, ignoring subsequent switching. Identified under
  baseline exchangeability given $L_{m,0}$, together with the assumptions in
  2.9; estimated with the treatment weight alone.
- *Per-protocol*: the contrast of sustained treatment versus sustained
  non-treatment. Follow-up is censored at the first deviation from the
  assigned strategy; identified under the additional assumption that the
  censoring model captures all joint determinants of deviation and outcome;
  estimated with the product of treatment and censoring weights.

*Interpretation under non-proportional hazards.* The reported IRR is the
coefficient of a proportional-rates working model. When the true marginal
rate ratio varies over follow-up (for example under depletion of
susceptibles, or with effects that accumulate over time), the single IRR is a
person-time-weighted average of the time-varying rate ratio — a well-defined
summary, but one that can differ from, say, the ratio of cumulative
incidences over the full horizon. Simulations with strongly time-varying
effects show that swereg and `TrialEmulation` produce the same
weighted-average summary (Section 3.4, Table 8). Where time-varying effects
are of scientific interest, follow-up-specific estimates (for example by
follow-up horizon) should be reported rather than the pooled IRR alone.

## 2.3 Follow-up construction and censoring events

For each person-trial, follow-up stops at the earliest of: (1) the first
outcome event; (2) the first protocol deviation (PP only; the ITT panel never
censors at switching); (3) the person's end of observed data, when that occurs
before any planned stop; (4) the pre-specified administrative end of study;
and (5) the pre-specified analysis horizon.

Deviation is the first band in which the observed treatment status differs
from the assigned arm (initiators off treatment; comparators on treatment);
a band with missing on-treatment status counts as deviation. There is no
grace period: deviation censors at the first mismatched band (a grace-period
design would require cloning, which this pipeline does not implement; Hernán
and Robins 2016).

*Event-priority convention.* If the first event and the first deviation
fall in the same band, the person-trial exits through the event: the
outcome is measured over the interval before within-interval censoring is
applied, so the band counts as an event, not a censoring, in both the
censoring model and the analysis data. The alternative — treating collision
bands as censorings — discards real events and undercounts the per-protocol
outcome in switching-heavy data; see the provenance note in Section 4.2.

Rows at and before the stop band are retained; censoring-event rows (their
band person-time included) are removed from the analysis data after the
censoring weights are estimated (2.5), so the analysis panel contains only
protocol-consistent, at-risk person-time.

## 2.4 Baseline treatment weights (IPW)

On the baseline row of each person-trial, a logistic regression of assignment
on the baseline confounders (main effects) is fit:

$$
\mathrm{logit}\,\mathrm{Pr}(A_{m,0} = 1 \mid L_{m,0})
  = \gamma_0 + \gamma^\top L_{m,0},
$$

and the stabilised weight uses the marginal initiation fraction
$\bar{p} = \widehat{\mathrm{Pr}}(A_{m,0}=1)$ as numerator:

$$
SW^A =
  A_{m,0}\,\frac{\bar{p}}{\widehat{ps}} +
  (1 - A_{m,0})\,\frac{1-\bar{p}}{1-\widehat{ps}},
\qquad \widehat{ps} = \widehat{\mathrm{Pr}}(A_{m,0}=1 \mid L_{m,0}).
$$

The weight is constant across a person-trial's follow-up rows. Missing
baseline confounders are singly imputed by hot-deck sampling from observed
values (fixed seed) before the model is fit; imputation uncertainty is not
propagated (2.8). The propensity model is main-effects only: if strong
non-linearity or interactions are suspected, they must be encoded as derived
confounder variables in the protocol specification.

## 2.5 Per-protocol censoring weights (IPCW)

Censoring ($C_{m,j} = 1$: deviation or loss in band $j$) is modelled on the
panel before censoring-event rows are dropped. The default censoring model,
fit separately by assigned arm $a$, is a discrete-time logistic
generalized additive model:

$$
\mathrm{logit}\,\mathrm{Pr}(C_{m,j} = 1 \mid \text{at risk},\ A_{m,0}=a)
  = s_a(j) + s_a(m) + \alpha_a^\top L_{m,j},
$$

with penalised-spline smooth functions $s_a(\cdot)$ of follow-up band and of
the trial index (a linear trial term when there are few bands; a fully
linear-in-time specification is available as a pre-specified sensitivity
option). The confounder columns carry their per-band updated values, so
time-varying confounders, where available in the source data, inform the
censoring model. Arm-specific fits fall back to the arm's marginal censoring
rate when a stratum has no (or all) censoring events or too few rows to
support the model.

The stabilised weight for the row in band $k$ is a ratio of cumulative
uncensored probabilities through band $k$ inclusive:

$$
SW^C_{m,k}
 = \frac{\prod_{j=0}^{k} \bar{q}_{a}(j)}
        {\prod_{j=0}^{k} \widehat{\mathrm{Pr}}(C_{m,j}=0 \mid \cdot)},
$$

where $\bar{q}_a(j)$ is the numerator: the marginal mean uncensored
probability at band $j$ within arm $a$ (by band only, when arms are pooled).

The construction deviates deliberately from the textbook version in two
respects:

- *Inclusive cumulative product.* Because the censoring-event row is
  subsequently removed, a row present at band $k$ exists if and only if the
  person-trial is uncensored through $k$; the weight therefore includes
  band $k$'s own uncensoring probability. (With the convention that censored
  bands stay in the risk set, the product would stop at $k-1$; the two
  conventions must not be mixed.)
- *Marginal numerator.* Canonical stabilisation (Danaei et al. 2013) uses a
  numerator model conditional on baseline covariates, which then requires
  those covariates in the outcome model. Here the outcome model is
  covariate-free (marginal MSM, 2.7), so the numerator is the marginal
  uncensored probability by band and arm. This preserves consistency of the
  marginal estimand; it stabilises slightly less aggressively when baseline
  covariates strongly predict censoring.

## 2.6 Final analysis weights and truncation

$$
W_{i,m,j} = SW^A_{i,m} \times SW^C_{i,m,j}
$$

for the per-protocol panel; $W_{i,m,j} = SW^A_{i,m}$ for the ITT panel.
Weights are truncated at percentiles (1st/99th by default) of the pooled
person-band rows: the ITT weight directly, and the PP weight as the truncated
product. Component-wise truncation is not applied, so extreme components
can offset; sensitivity analyses may truncate components separately.
Primary analyses use truncated weights; untruncated PP results are exported
alongside as a sensitivity analysis.

*Positivity and the truncation tradeoff.* Truncation trades variance for
bias toward the null: in a simulation with near-violation of positivity
(propensity scores spanning 0–1), attenuation of the log-IRR grows
monotonically with truncation severity (quantified in Section 3.4, Table 9).
The propensity-score overlap and the raw weight distribution should be
inspected; if extreme weights are structural (deterministic treatment within
a stratum), the eligible population should be restricted rather than the
truncation tightened.

## 2.7 Outcome model

The IRR is estimated by a weighted quasi-Poisson marginal structural model on
the analysis panel:

$$
\log \mathrm{E}[Y_{i,m,j}]
  = \beta_0 + \beta_1 A_{i,m,0}
  + \mathrm{ns}(j, 3)
  + f(m)
  + \log(\text{person-weeks}_{i,m,j}),
$$

fit by weighted quasi-Poisson regression with survey-linearised variance
(Section 4.1). $\mathrm{ns}(j, 3)$ is a natural cubic spline of follow-up
band (the discrete-time baseline-rate analogue); $f(m)$ is a natural spline
of the trial index with 3 df (linear when 2–4 bands; omitted for a single
band), adjusting smoothly for calendar trends while sharing one treatment
coefficient across trials (Danaei et al. 2013; Caniglia et al. 2023). No
confounders enter the outcome model: $\exp(\beta_1)$ is the marginal IRR
(Section 3.7 discusses marginal versus conditional estimands). Descriptive
weighted event counts, person-years (52.25 weeks/year), and rates per
100,000 person-years accompany each IRR.

## 2.8 Inference

Standard errors are survey-linearised (Huber–White sandwich) with clustering
on the person identifier, not the person-trial, accounting for repeated
person-trials and repeated bands within person (Hernán and Robins 2008;
Danaei et al. 2013; Su et al. 2024). Confidence intervals are Wald on the log
scale, $\exp(\hat\beta_1 \pm 1.96\,\widehat{se})$. Two caveats apply:

- The variance treats the estimated weights (and the hot-deck imputation) as
  fixed. For stabilised weights this is typically slightly conservative for
  the treatment coefficient, but it is not exact; a person-level bootstrap of
  the entire pipeline is the fuller alternative for definitive reporting.
- Monte Carlo calibration (Section 3.6, Table 15) shows near-nominal
  coverage where the estimand's assumptions hold, mild undercoverage under
  confounding with independent loss, and coverage degradation driven by
  bias, not by the variance estimator, when an estimand ignores informative
  loss.

## 2.9 Identifying assumptions

For the intention-to-treat analogue: (1) consistency; (2) no unmeasured
confounding of baseline assignment given the baseline confounders at each
trial's baseline; (3) positivity of assignment within confounder strata;
(4) loss to follow-up independent of the outcome. No censoring weights are
applied to the ITT panel; simulation shows the estimand holds under
independent loss and is biased under informative loss, in swereg and
`TrialEmulation` alike (Section 3.3).

For the per-protocol estimand, additionally: (5) the censoring model (2.5)
captures all joint determinants of protocol deviation/loss and the outcome,
including their time-varying values as materialised in the source data;
(6) positivity of continued adherence. Under strong treatment–confounder
feedback the single-model IPCW approach retains residual bias: time-updated
censoring covariates remove part, not all, of the deviation selection bias
relative to freezing them at baseline (quantified in Section 3.4, Table 10);
where feedback is central, g-methods beyond this pipeline are indicated.

## 2.10 Heterogeneity, subgroups, and small cells

Effect heterogeneity across calendar time is tested by a joint Wald test of
the treatment × trial-index spline interaction; effect modification by
pre-specified baseline subgroups by treatment × subgroup interaction, with
stratified IRRs per level. Zero-event strata return no estimate rather than
an unstable one. Enrollments and outcomes are pre-specified in a
machine-readable study specification; results tables report weighted events,
person-years, rates, IRR, CI, and p-value per estimand, plus CONSORT-style
attrition (unique persons and person-trials separately, per Cashin et al.
2025).

## 2.11 Known limitations

- No grace periods and no cloning; deviation censors at the first mismatched
  band (2.3).
- No as-treated estimand.
- Single hot-deck imputation of missing baseline confounders (no variance
  propagation).
- Comparator downsampling (2.1) discards comparator information (efficiency,
  not bias).
- The propensity and censoring models are main-effects (plus smooth time)
  specifications; non-linearities must be pre-encoded as derived variables.

---

# 3. Validation evidence

Every number in this section is computed, not asserted. The tables and
figures are rendered from a results artifact regenerated by rerunning the
complete validation battery — the same data-generating processes, truth
calculations, and fit wrappers that the package's test suite enforces in
continuous integration. Section 4.3 maps each layer to its test file and
describes how to regenerate the artifact.

Provenance: generated `r ev$meta$generated_utc` with swereg
`r ev$meta$swereg`, TrialEmulation `r ev$meta$trialemulation`, under
`r ev$meta$r`.

## 3.1 Design of the validation battery

The battery is organised around one principle: an estimator is validated by
recovering a truth that is *known by construction*, not by agreeing with
another implementation. Agreement between two packages is used as
corroborating evidence only — two correct implementations of the same
estimand must agree, but two implementations can also agree while both miss
the truth, and the battery deliberately includes a scenario (informative
loss under the ITT estimand) that demonstrates exactly this.

Truth is therefore computed by direct counterfactual simulation. For each
scenario and estimand, 200,000 persons per arm are simulated under the forced
strategy — for the per-protocol truth, treatment is held at the assigned
value in every period; for the ITT truth, only the baseline value is forced
and subsequent treatment follows the scenario's natural switching process —
and the truth is the log ratio of first-event incidence rates, with events
counted until each person's first event and person-time accumulated only
while at risk. Loss to follow-up is never applied to the truth simulation:
loss is a nuisance the estimator must be robust to, not part of the estimand.
This first-event, person-time-at-risk construction matches the estimand
targeted by the weighted quasi-Poisson model (2.7) exactly; a
recurrent-event or fixed-denominator construction would target a different
quantity.

Four layers separate concerns, so that a failure localises to a pipeline
segment:

```{r}
layers <- data.table(
  layer = c(
    "Cross-package matrix (3.3)",
    "Stress matrix (3.4)",
    "Plan-layer truth matrix (3.5)",
    "Coverage calibration (3.6)"
  ),
  segment = c(
    "Enrollment-layer estimators (IPW, IPCW, weighted MSM) on person-period panels",
    "The same estimators at design extremes",
    "The complete production pipeline: specification, banding, sequential eligibility, matching, worker subprocesses, dual analysis files",
    "The sandwich variance estimator"
  ),
  question = c(
    "Do swereg and TrialEmulation each recover known truth where the estimand's assumptions hold, and fail identically where they do not?",
    "Does the estimator remain stable under rare outcomes, null and harmful effects, near-positivity violation, heavy informative attrition, and treatment-confounder feedback?",
    "Does the pipeline as a whole recover a planted constant-hazard truth, including the separation of PP from ITT under discontinuation?",
    "Do nominal 95% intervals cover the truth 95% of the time when the estimand is valid?"
  )
)
knitr::kable(
  layers,
  col.names = c("Layer", "Pipeline segment exercised", "Question answered"),
  align = "lll",
  caption = "Table 1. The four layers of the validation battery."
)
```

Interpreting single-dataset cells requires one calibration: at the sample
sizes used here, one simulated dataset carries Monte Carlo noise of roughly
0.03–0.05 on the log-IRR scale, so a single-run gap of that order is
indistinguishable from zero. All cells run at fixed seeds and are therefore
exactly reproducible; the multi-replicate cells (Tables 5, 8, 13, 14, and 15)
quantify bias and coverage across repeated draws, free of this caveat.

## 3.2 Enrollment-layer scenarios: data-generating processes

The enrollment-layer cells (3.3, 3.4) share one person-period
data-generating process. For person $i$ with standard-normal baseline
confounder $L_{0i}$ and periods $t = 0, \ldots, 19$:

$$
\begin{aligned}
\mathrm{logit}\,\mathrm{Pr}(A_{i0} = 1)
  &= -0.3 + \phi_A L_{0i}
  && \text{(baseline initiation)} \\
\mathrm{logit}\,\mathrm{Pr}(A_{it} = 1)
  &= -3.0 + \phi_S L_{0i} + 8\,A_{i,t-1}, \quad t \ge 1
  && \text{(switching with persistence)} \\
\mathrm{logit}\,\mathrm{Pr}(Y_{it} = 1)
  &= -3.5 + \theta A_{it} + \phi_Y L_{0i}
  && \text{(outcome)}
\end{aligned}
$$

with true contemporaneous treatment effect $\theta = -0.7$ unless a cell
varies it. The persistence coefficient of 8 keeps most initiators on
treatment (adherent person-time dominates) while still generating enough
switching to separate the PP and ITT truths. Loss to follow-up, when
present, is geometric dropout from a per-person hazard: constant at 0.06 per
period (independent loss), or $\mathrm{expit}(-2.4 + 0.9\,L_{0i})$
(informative loss, so that dropout selects on the confounder that also
drives treatment and outcome). The three standard scenarios switch the
nuisance parameters only, leaving the true effect identical:

```{r}
dgp <- data.table(
  scenario = c("s1", "s2", "s3"),
  phi_a = c("0", "0.6", "0.6"),
  phi_s = c("0", "0.4", "0.4"),
  phi_y = c("0", "0.4", "0.4"),
  loss = c("none", "independent (hazard 0.06)", "informative (expit(-2.4 + 0.9 L0))"),
  induces = c(
    "clean benchmark: no confounding, no selection",
    "baseline confounding plus outcome-independent attrition",
    "baseline confounding plus attrition that selects on the confounder"
  )
)
knitr::kable(
  dgp,
  col.names = c(
    "Scenario", "$\\phi_A$", "$\\phi_S$", "$\\phi_Y$",
    "Loss to follow-up", "What it induces"
  ),
  align = "lccccl",
  caption = "Table 2. Nuisance configuration of the three standard scenarios."
)
```

Each scenario dataset is simulated at 20,000 persons and 20 periods with a
fixed seed. Table 3 reports the realized characteristics of the exact
datasets analysed in 3.3.

```{r}
td <- copy(ev$triangle_desc)
knitr::kable(
  td[, .(
    scenario,
    n_persons = format(n_persons, big.mark = ","),
    person_periods = format(person_periods, big.mark = ","),
    pct_lost = sprintf("%.0f%%", pct_periods_lost),
    pct_init = sprintf("%.1f%%", pct_initiators),
    first_events = format(first_event_persons, big.mark = ","),
    risk = sprintf("%.2f%%", event_risk_band_pct)
  )],
  col.names = c(
    "Scenario", "Persons", "Person-periods", "Person-periods lost",
    "Initiators at baseline", "Persons with ≥1 event", "Event risk per period"
  ),
  align = "lrrrrrr",
  caption = "Table 3. Realized descriptives of the three scenario datasets."
)
```

## 3.3 Cross-package validation matrix

Each scenario dataset is fed through the full triangle — known
potential-outcome truth, swereg, and `TrialEmulation` — for both estimands.
Estimates are compared on a common rate-ratio scale: `TrialEmulation` reports
odds ratios from pooled logistic regression, which are converted with the
Zhang–Yu relation using the reference arm's per-period event risk from the
truth simulation (Zhang and Yu 1998; Section 3.7). `TrialEmulation` is a
peer required to recover the truth itself, not an oracle.

```{r}
tri <- copy(ev$triangle)
tri[, nuisance := fcase(
  scenario == "s1", "none",
  scenario == "s2", "confounding + independent loss",
  scenario == "s3", "confounding + informative loss"
)]
tri[, `:=`(
  sw = ci3(sw_est, sw_lo, sw_hi),
  te = ci3(te_est, te_lo, te_hi),
  bias_sw = sg3(sw_est - truth),
  bias_te = sg3(te_est - truth),
  gap = sg3(sw_est - te_est)
)]
knitr::kable(
  tri[, .(
    scenario, nuisance, estimand,
    truth = sprintf("%.3f", truth), sw, te, bias_sw, bias_te, gap
  )],
  col.names = c(
    "Scenario", "Nuisances", "Estimand", "True log-IRR",
    "swereg [95% CI]", "TrialEmulation [95% CI]",
    "swereg bias", "TE bias", "swereg − TE"
  ),
  align = "lllrrrrrr",
  caption = paste0(
    "Table 4. Cross-package validation matrix (N = 20,000, T = 20 periods, ",
    "fixed seed). Log-IRR scale."
  )
)
x31 <- list(
  max_ok = tri[scenario != "s3" | estimand != "itt", max(abs(sw_est - truth))],
  itt_sw = tri[scenario == "s3" & estimand == "itt", sg3(sw_est - truth)],
  itt_te = tri[scenario == "s3" & estimand == "itt", sg3(te_est - truth)],
  itt_gap = tri[scenario == "s3" & estimand == "itt", sprintf("%.3f", abs(sw_est - te_est))]
)
```

In Table 4, every interval in a cell whose assumptions hold covers the
truth; the per-protocol estimator remains close to the truth in s3 precisely
because its censoring weights (2.5) model the informative loss; and in the
s3 ITT cell both packages miss on the same side (swereg `r x31$itt_sw`,
TrialEmulation `r x31$itt_te`) while agreeing with each other to within
`r x31$itt_gap`. A single dataset is nonetheless a weak demonstration: at
N = 20,000 one estimate carries Monte Carlo noise of roughly 0.03–0.05 on
the log-IRR scale, so point estimates sit visibly away from the truth even
under a perfectly unbiased estimator, and the size of any one gap is a
draw of luck. The definitive statement comes from replication: the full
triangle is repeated on 20 independent datasets per scenario, which reduces
the Monte Carlo standard error of the estimated bias by a factor of
$\sqrt{20}$.

```{r}
tr_reps <- copy(ev$triangle_reps)
tr5 <- tr_reps[, .(
  truth = first(truth),
  R = .N,
  sw_bias = mean(sw_est - truth),
  sw_se = sd(sw_est - truth) / sqrt(.N),
  swu_bias = mean(sw_est_untrunc - truth),
  swu_se = sd(sw_est_untrunc - truth) / sqrt(.N),
  te_bias = mean(te_est - truth),
  te_se = sd(te_est - truth) / sqrt(.N),
  gap = mean(abs(sw_est - te_est))
), by = .(scenario, estimand)]
knitr::kable(
  tr5[, .(
    scenario, estimand, R,
    truth = sprintf("%.3f", truth),
    sw = sprintf("%+.3f (%.3f)", sw_bias, sw_se),
    swu = sprintf("%+.3f (%.3f)", swu_bias, swu_se),
    te = sprintf("%+.3f (%.3f)", te_bias, te_se),
    gap = sprintf("%.3f", gap)
  )],
  col.names = c(
    "Scenario", "Estimand", "Datasets", "True log-IRR",
    "swereg, truncated weights: mean bias (MC SE)",
    "swereg, untruncated weights: mean bias (MC SE)",
    "TrialEmulation: mean bias (MC SE)", "Mean |swereg − TE|"
  ),
  align = "llrrrrrr",
  caption = paste0(
    "Table 5. Replicated cross-package matrix: mean bias over 20 independent ",
    "datasets per scenario (N = 20,000 each), with the Monte Carlo standard ",
    "error of the mean. swereg is shown with its primary (1st/99th ",
    "percentile) weight truncation and with untruncated weights. ",
    "Log-IRR scale."
  )
)
mo <- tr5[scenario != "s3" | estimand != "itt", max(abs(sw_bias))]
x33r <- list(
  s1_max = tr5[scenario == "s1", sprintf("%.3f", max(abs(sw_bias)))],
  max_ok = sprintf("%.3f", mo),
  rel_ok = sprintf("%.0f%%", 100 * (exp(mo) - 1)),
  max_se = tr5[, sprintf("%.3f", max(sw_se))],
  s3_bias = tr5[scenario == "s3" & estimand == "itt", sprintf("%+.3f", sw_bias)],
  s3_z = tr5[scenario == "s3" & estimand == "itt", sprintf("%.0f", abs(sw_bias) / sw_se)],
  s3pp = tr5[scenario == "s3" & estimand == "pp", sprintf("%+.3f", sw_bias)],
  s3pp_u = tr5[scenario == "s3" & estimand == "pp", sprintf("%+.3f (MC SE %.3f)", swu_bias, swu_se)],
  s2pp = tr5[scenario == "s2" & estimand == "pp", sprintf("%+.3f", sw_bias)]
)
```

```{r, fig.width = 7, fig.height = 5, fig.cap = "Figure 1. Bias of the estimated log-IRR over 20 independent datasets per scenario (N = 20,000 each): swereg with its primary truncated weights, swereg with untruncated weights, and TrialEmulation. Faint points are individual datasets; solid points are the mean bias with its 95% Monte Carlo interval; the vertical line marks zero bias. The truncated-versus-untruncated contrast isolates the cost of weight truncation: the two swereg fits are indistinguishable everywhere except the s3 per-protocol cell, where informative loss produces the large late-follow-up censoring weights that truncation clips, and untruncating removes most of the displacement. The intention-to-treat estimand under informative loss (s3) remains displaced in every fit — the estimand, not the software, fails there."}
pd <- rbind(
  tr_reps[, .(scenario, estimand, rep, fit = "swereg, truncated weights", bias = sw_est - truth)],
  tr_reps[, .(scenario, estimand, rep, fit = "swereg, untruncated weights", bias = sw_est_untrunc - truth)],
  tr_reps[, .(scenario, estimand, rep, fit = "TrialEmulation", bias = te_est - truth)]
)
# reversed levels: discrete y draws the first level at the bottom, and the
# panel should read s1 (top) -> s3 (bottom)
pd[, scenario_lab := factor(scen_lab(scenario), levels = scen_lab(c("s3", "s2", "s1")))]
pd[, estimand_lab := fifelse(estimand == "pp", "Per-protocol", "Intention-to-treat")]
pd[, fit := factor(fit, c(
  "swereg, truncated weights", "swereg, untruncated weights", "TrialEmulation"
))]
pm <- pd[, .(
  mean_bias = mean(bias), se = sd(bias) / sqrt(.N)
), by = .(scenario_lab, estimand_lab, fit)]
q <- ggplot(pd, aes(y = scenario_lab, colour = fit, group = fit))
q <- q + geom_vline(xintercept = 0, colour = "grey30")
q <- q + geom_point(
  aes(x = bias),
  position = position_dodge(width = 0.7), size = 0.9, alpha = 0.3
)
q <- q + geom_pointrange(
  data = pm,
  aes(x = mean_bias, xmin = mean_bias - 1.96 * se, xmax = mean_bias + 1.96 * se),
  position = position_dodge(width = 0.7), size = 0.45, linewidth = 0.9
)
q <- q + facet_wrap(~estimand_lab, ncol = 1)
q <- q + scale_colour_manual(values = c(
  "swereg, truncated weights" = "#0072B2",
  "swereg, untruncated weights" = "#56B4E9",
  "TrialEmulation" = "#D55E00"
))
q <- q + labs(x = "Bias of the log-IRR (estimate − truth)", y = NULL, colour = NULL)
q <- q + theme_minimal() + theme(legend.position = "bottom")
q
```

Averaging over 20 datasets reduces the Monte Carlo standard error of the
estimated bias to roughly `r x33r$max_se` — fine enough to resolve
systematic effects that no single dataset can. Three magnitudes emerge from
Table 5 and Figure 1. In the clean scenario (s1) both packages are unbiased
within Monte Carlo error for both estimands (swereg mean bias at most
`r x33r$s1_max` in absolute value): the estimation machinery itself
introduces no bias. Where nuisances are present but the estimand remains
valid (s2, and the per-protocol estimand in s3), small systematic residuals
of up to `r x33r$max_ok` become resolvable — at most a `r x33r$rel_ok`
relative error on the rate-ratio scale, well inside the tolerances the test
suite enforces, but reported here rather than rounded away.

The truncated-versus-untruncated contrast localises the largest of these
residuals. In the s3 per-protocol cell, swereg's mean bias of `r x33r$s3pp`
falls to `r x33r$s3pp_u` when the same replicates are refit with
untruncated weights: most of the displacement is the price of clipping the
weights, exactly the variance-for-bias trade described in 2.6. Informative
dropout means the high-risk individuals still under observation late in
follow-up must carry large censoring weights to represent those who left;
the 1st/99th-percentile truncation caps precisely those weights, and the
under-corrected selection surfaces as bias toward the null. This is also
why the untruncated per-protocol results are exported as a sensitivity
analysis: a material divergence between the truncated and untruncated
estimates is the working diagnostic that truncation is eating the
correction. The event-priority convention (2.3) is excluded as a cause by
design contrast: the s2 per-protocol cell shares the identical switching,
censoring, and event-accounting machinery, differs only in that its loss is
non-informative, and shows no bias (`r x33r$s2pp`). The remaining small
residuals (for example in the s2 ITT cell) are consistent with loss
truncating follow-up toward its early bands, so that the
person-time-weighted working-model summary (2.2) no longer weights
follow-up exactly as the no-loss truth functional does.

The s3 ITT cell is of a different kind: mean bias `r x33r$s3_bias`, roughly
`r x33r$s3_z` Monte Carlo standard errors from zero and present in both
packages — a displacement that replication sharpens rather than removes,
because the ITT estimand carries no loss weight and informative loss
therefore biases it in any correct implementation. Cross-package agreement
is not evidence of correctness, which is why every layer of this battery is
anchored to simulated truth.

## 3.4 Stress matrix

The stress cells reuse the Section 3.2 data-generating process with one or
two parameters pushed to an extreme, so that each cell probes a specific
failure mode. Table 5 specifies the designs; the cells then follow in order.

```{r}
sd_tab <- data.table(
  cell = c(
    "Rare outcome", "Null effect", "Informative attrition",
    "Harmful effect, depletion", "Near-positivity violation",
    "Treatment-confounder feedback", "Determinism"
  ),
  design = c(
    "Outcome intercept −6.0 (≈0.2% risk/period), N = 40,000, θ = −0.7",
    "θ = 0, independent loss, N = 20,000",
    "Dropout hazard expit(−1.3 + 0.9 L0): ≈73% of person-periods lost, selecting on the confounder; N = 30,000",
    "θ = +0.7, three independent seeds at N = 20,000, TrialEmulation cross-check",
    "φA = 2.5: propensity scores span 0–1; ITT fit at three truncation levels",
    "AR(1) confounder Lt = 0.7 Lt−1 − 0.4 At−1 + εt driving both switching (0.8 Lt) and outcome (0.5 Lt); N = 25,000",
    "Identical data, PP estimator fit twice"
  ),
  probes = c(
    "Sparse-event stability of the weighted MSM and the spline IPCW model",
    "False-positive effects (does the pipeline manufacture signal from noise?)",
    "IPCW under heavy selection; the ITT arm of this cell is expected to fail",
    "The person-time-weighted-average interpretation of the pooled IRR (2.2)",
    "The truncation bias-variance tradeoff (2.6)",
    "The residual-bias limit of single-model IPCW under feedback (2.9)",
    "Uncontrolled stochastic steps anywhere in the fit"
  )
)
knitr::kable(
  sd_tab,
  col.names = c("Cell", "Design deviation from the base DGP", "What it probes"),
  align = "lll",
  caption = "Table 6. Stress-cell designs. All other parameters as in Section 3.2; T = 20 periods, fixed seeds."
)
```

```{r}
st <- copy(ev$stress)
st[, covers := fifelse(truth >= lo & truth <= hi, "yes", "no")]
knitr::kable(
  st[, .(
    cell, estimand, truth = sprintf("%.3f", truth),
    est = ci3(est, lo, hi), bias = sg3(est - truth), covers, note
  )],
  col.names = c(
    "Cell", "Estimand", "True log-IRR", "Estimate [95% CI]",
    "Bias", "CI covers truth", "Note"
  ),
  align = "llrrrcl",
  caption = "Table 7. Stress cells, single dataset at fixed seed. Log-IRR scale."
)
x32a <- list(
  rare_pp = st[cell == "rare_outcome" & estimand == "pp", sg3(est - truth)],
  attr_pp = st[cell == "informative_attrition" & estimand == "pp", sg3(est - truth)],
  attr_itt = st[cell == "informative_attrition" & estimand == "itt", sg3(est - truth)]
)
```

Three observations from Table 7. At an event risk of roughly 0.2% per
period the per-protocol machinery — including the spline-based censoring
model — remains stable (bias `r x32a$rare_pp`). Under a true null the
estimate is small and its interval covers zero: the weighting and pooling
machinery does not manufacture an effect. And in the attrition cell, where
almost three quarters of person-periods are removed by confounder-driven
dropout, the per-protocol estimator stays at the truth
(bias `r x32a$attr_pp`) while the ITT estimator, which by construction
carries no loss weight, is displaced (bias `r x32a$attr_itt`) and its
interval excludes the truth — the designed failure that motivates the
estimand distinction in practice.

Determinism: refitting the per-protocol estimator on identical data
reproduced the estimate to a maximum absolute difference of
`r format(ev$stress_determinism$max_abs_delta)`; the pipeline has no
uncontrolled stochastic step.

```{r}
sh <- copy(ev$stress_harmful)
knitr::kable(
  sh[, .(
    seed, truth = sprintf("%.3f", truth),
    sw = sprintf("%.3f", sw_est), te = sprintf("%.3f", te_est),
    bias_sw = sg3(sw_est - truth), gap = sg3(sw_est - te_est)
  )],
  col.names = c(
    "Seed", "True log-IRR (cumulative-rate)", "swereg", "TrialEmulation",
    "swereg − truth", "swereg − TE"
  ),
  align = "lrrrrr",
  caption = paste0(
    "Table 8. Harmful effect (true log-IRR > 0) with strong depletion of ",
    "susceptibles, ITT, three seeds. Log-IRR scale."
  )
)
x32 <- list(
  mb = sh[, sg3(mean(sw_est - truth))],
  mg = sh[, sprintf("%.3f", max(abs(sw_est - te_est)))]
)
```

Under a harmful effect with strong depletion of susceptibles, the marginal
hazard ratio declines over follow-up, so the single pooled IRR — a
person-time-weighted average (2.2) — legitimately sits above the
cumulative-rate truth, by a mean of `r x32$mb` across the three seeds. This
is a property of the estimand, not an implementation defect: swereg and
`TrialEmulation` agree to within `r x32$mg` on every seed because both
target the same weighted-average summary. Analyses in which the time path
of the effect matters should report follow-up-specific estimates.

```{r}
tc <- copy(ev$stress_trunc)
knitr::kable(
  tc[, .(
    trunc_percentiles, truth = sprintf("%.3f", truth),
    est = sprintf("%.3f", est), bias = sg3(est - truth),
    wmax = sprintf("%.0f", wmax_raw)
  )],
  col.names = c(
    "Truncation percentiles (%)", "True log-IRR", "Estimate",
    "Bias (attenuation)", "Max raw stabilised weight"
  ),
  align = "lrrrr",
  caption = paste0(
    "Table 9. Near-positivity violation: attenuation toward the null grows ",
    "monotonically with truncation severity. ITT, log-IRR scale."
  )
)
```

Table 9 quantifies the tradeoff stated in 2.6 on a design whose propensity
scores approach the boundary (maximum raw stabilised weight
`r tc[1, sprintf("%.0f", wmax_raw)]`): each tightening of the truncation
percentiles buys variance at the price of measurable bias toward the null.
When extreme weights are structural rather than sporadic, the appropriate
response is to restrict the eligible population, not to truncate harder.

```{r}
tv <- copy(ev$stress_tv)
knitr::kable(
  tv[, .(
    fit, truth = sprintf("%.3f", truth), est = ci3(est, lo, hi),
    bias = sg3(est - truth), abs_bias = sprintf("%.3f", abs(est - truth))
  )],
  col.names = c(
    "Fit", "True log-IRR", "Estimate [95% CI]", "Bias", "|Bias|"
  ),
  align = "lrrrr",
  caption = paste0(
    "Table 10. Treatment–confounder feedback: per-protocol bias with the ",
    "censoring covariate time-updated versus frozen at baseline; ITT for ",
    "reference. Log-IRR scale."
  )
)
x32tv <- list(
  up = tv[1, sprintf("%.3f", abs(est - truth))],
  fr = tv[2, sprintf("%.3f", abs(est - truth))],
  itt = tv[3, sg3(est - truth)]
)
```

The feedback cell is the honest edge of the per-protocol estimator. When a
time-varying confounder is itself affected by treatment and drives both
adherence and the outcome, IPCW with time-updated covariates is strictly
less biased than IPCW with covariates frozen at baseline (|bias|
`r x32tv$up` versus `r x32tv$fr`), but a residual bias remains — part of
which is the working-model average under a time-ramping effect rather than
selection per se. The ITT estimand, which needs no censoring model against
this feedback, is near-unbiased in the same data (bias `r x32tv$itt`).
Where treatment–confounder feedback is central to the question, g-methods
beyond this pipeline are indicated, exactly as stated in 2.9.

## 3.5 Full-pipeline truth recovery (plan layer)

The layers above validate the estimators on pre-built person-period panels.
This layer validates everything that sits on top in production: the
machine-readable specification, trial-band assignment, sequential
eligibility with a lifetime new-user exclusion, per-band 2:1 comparator
matching, the worker subprocess chain, the dual PP/ITT analysis files, and
the pooled weighted outcome model.

The data-generating process plants an exactly known truth in a realistic
skeleton. Persons are observed weekly from 2016-01-01 to 2021-06-30 —
roughly 287 ISO weeks, deliberately spanning 2020's 53-week ISO year — and
are split into never-treaters and initiators; initiators start treatment at
a band drawn uniformly from the first 56 four-week bands and, in the
discontinuation cell, stop after a geometric duration (4% weekly hazard).
The weekly outcome hazard is constant at 0.0025 untreated and doubled while
treated, so the marginal per-week incidence rate ratio among sustained
users is exactly 2.0. Scenario B adds a binary frailty carried by 30% of
persons that doubles both the initiation probability and the outcome
hazard — a genuine baseline confounder; mixture-averaging over the two
risk groups with first-event depletion attenuates the marginal truth to
1.982. Loss, when present, is geometric (2% weekly, or 1%/3% by risk group
for informative loss) and multiplies person-time equally in both arms, so
the truth is unchanged and loss is purely a nuisance the machinery must
tolerate. The ITT truth in the discontinuation cell
(`r sprintf("%.2f", ev$plan[cell == "DISC", truth_itt])`) is simulated
directly as the do(initiate)-versus-do(never) contrast with natural
discontinuation.

```{r}
pl <- copy(ev$plan)
knitr::kable(
  pl[, .(
    cell, scenario, loss,
    n = format(n_persons, big.mark = ","),
    pw = format(sk_person_weeks, big.mark = ","),
    tx_pw = format(sk_treated_person_weeks, big.mark = ","),
    events = format(sk_events_n, big.mark = ",")
  )],
  col.names = c(
    "Cell", "Scenario", "Loss", "Persons", "Person-weeks",
    "Treated person-weeks", "Events"
  ),
  align = "lllrrrr",
  caption = "Table 11. Skeleton descriptives per plan-layer cell (before eligibility and enrollment)."
)
```

```{r}
pl[, covers_pp := fifelse(pp_lo <= truth_pp & truth_pp <= pp_hi, "yes", "no")]
pl[, covers_itt := fifelse(itt_lo <= truth_itt & truth_itt <= itt_hi, "yes", "no")]
knitr::kable(
  pl[, .(
    cell, loss,
    truth_pp = sprintf("%.2f", truth_pp),
    pp = ci2(pp_irr, pp_lo, pp_hi), covers_pp,
    truth_itt = sprintf("%.2f", truth_itt),
    itt = ci2(itt_irr, itt_lo, itt_hi), covers_itt
  )],
  col.names = c(
    "Cell", "Loss", "PP truth", "PP IRR [95% CI]", "covers",
    "ITT truth", "ITT IRR [95% CI]", "covers"
  ),
  align = "llrrcrrc",
  caption = paste0(
    "Table 12. Plan-layer factorial (A = no confounding, B = baseline ",
    "confounding; each × no/independent/informative loss) plus the ",
    "discontinuation cell. IRR scale."
  )
)
x33 <- list(
  crude = pl[cell == "B_none", sprintf("%.2f", crude_rr)],
  ipw = pl[cell == "B_none", sprintf("%.2f", ipw_rr)],
  truth_b = pl[cell == "B_none", sprintf("%.2f", truth_pp)],
  disc_gap = pl[cell == "DISC", sg3(log(pp_irr) - log(itt_irr))],
  disc_truth_gap = pl[cell == "DISC", sg3(log(truth_pp) - log(truth_itt))]
)
```

Two rows warrant comment. In the confounded no-loss cell (B_none) the
frailty is doing real confounding work: the crude rate ratio in the enrolled
ITT panel is `r x33$crude`, the IPW-weighted rate ratio `r x33$ipw`, against
a marginal truth of `r x33$truth_b` — the weighting removes essentially all
of the planted confounding. In the discontinuation cell the two estimands
genuinely separate, with PP − ITT = `r x33$disc_gap` on the log scale
against a true separation of `r x33$disc_truth_gap`: the per-protocol arm
censors at deviation and reweights back to the sustained-treatment truth of
2.0, while the ITT arm retains post-discontinuation person-time and
attenuates toward the do(initiate) truth. This cell also exercises the
event-priority convention (2.3), since events and deviations collide in the
same band whenever discontinuers have events in their final treated band.

Because a single pipeline run at a fixed seed cannot distinguish bias from
draw-level noise, the two no-loss scenarios are repeated over eight
independent seeds at 6,000 persons each, rerunning the complete pipeline
per replicate:

```{r}
mc <- copy(ev$plan_mc)
mc[, covers_pp := fifelse(pp_lo <= truth & truth <= pp_hi, "yes", "no")]
mc[, covers_itt := fifelse(itt_lo <= truth & truth <= itt_hi, "yes", "no")]
knitr::kable(
  mc[, .(
    scenario, seed,
    truth = sprintf("%.2f", truth),
    pp = ci2(pp_irr, pp_lo, pp_hi), covers_pp,
    itt = ci2(itt_irr, itt_lo, itt_hi), covers_itt
  )],
  col.names = c(
    "Scenario", "Seed", "Truth", "PP IRR [95% CI]", "covers",
    "ITT IRR [95% CI]", "covers"
  ),
  align = "llrrcrc",
  caption = "Table 13. Plan-layer Monte Carlo, per replicate. IRR scale."
)
mcs <- rbind(
  mc[, .(
    estimand = "pp",
    mean_bias = mean(log(pp_irr) - log(truth)),
    mc_sd = sd(log(pp_irr)),
    covered = sum(pp_lo <= truth & truth <= pp_hi),
    n_seeds = .N
  ), by = scenario],
  mc[, .(
    estimand = "itt",
    mean_bias = mean(log(itt_irr) - log(truth)),
    mc_sd = sd(log(itt_irr)),
    covered = sum(itt_lo <= truth & truth <= itt_hi),
    n_seeds = .N
  ), by = scenario]
)
setorder(mcs, scenario, -estimand)
knitr::kable(
  mcs[, .(
    scenario, estimand, mean_bias = sg3(mean_bias),
    mc_sd = sprintf("%.3f", mc_sd),
    coverage = sprintf("%d/%d", covered, n_seeds)
  )],
  col.names = c(
    "Scenario", "Estimand", "Mean log bias", "MC sd", "95% CI coverage"
  ),
  align = "llrrr",
  caption = "Table 14. Plan-layer Monte Carlo, summarised over the eight seeds. Log-IRR scale."
)
```

The mean log-scale bias is within Monte Carlo error of zero in both
scenarios, and coverage is consistent with the nominal 95% at eight
replicates. Individual misses — visible in Table 13 — are the expected
behaviour of honest intervals, not smoothed away.

## 3.6 Coverage calibration

The final layer asks whether the reported uncertainty can be trusted: over
200 replicate draws per scenario at 3,000 persons, each refit end to end,
what fraction of nominal 95% intervals cover the truth? The study uses the
ITT estimand, whose assumptions hold in s1 and s2 but are violated by the
informative loss in s3, so the three scenarios together probe both
calibration and the failure signature.

```{r}
cv <- copy(ev$coverage)
cv[, nuisance := fcase(
  scenario == "s1", "none",
  scenario == "s2", "confounding + independent loss",
  scenario == "s3", "confounding + informative loss"
)]
knitr::kable(
  cv[, .(
    scenario, nuisance,
    reps = sprintf("%d/%d", n_fit, M),
    mean_bias = sg3(mc_mean_bias), mc_sd = sprintf("%.3f", mc_sd),
    coverage = sprintf("%d/%d (%.1f%%)", covered_n, n_fit, 100 * coverage)
  )],
  col.names = c(
    "Scenario", "Nuisances", "Replicates fit", "Mean log bias",
    "MC sd", "95% CI coverage"
  ),
  align = "llrrrr",
  caption = paste0(
    "Table 15. ITT coverage calibration, M = 200 replicates per scenario at ",
    "N = 3,000. Log-IRR scale."
  )
)
x34 <- list(
  c1 = cv[scenario == "s1", sprintf("%.1f%%", 100 * coverage)],
  c2 = cv[scenario == "s2", sprintf("%.1f%%", 100 * coverage)],
  c3 = cv[scenario == "s3", sprintf("%.1f%%", 100 * coverage)],
  b3 = cv[scenario == "s3", sg3(mc_mean_bias)]
)
```

```{r, fig.width = 7, fig.height = 6, fig.cap = "Figure 2. Coverage calibration: all 200 replicate 95% confidence intervals per scenario, sorted by point estimate, against the true log-IRR (horizontal line). Intervals that miss the truth are drawn in red. In s1 and s2 the misses are the sampling-expected few percent, split across both tails; in s3 the interval cloud is displaced downward as a whole — the systematic bias of the ITT estimand under informative loss, which no variance estimator can repair."}
pd <- copy(ev$coverage_reps)
pd[, scenario_lab := scen_lab(scenario)]
setorder(pd, scenario, est)
pd[, idx := seq_len(.N), by = scenario]
q <- ggplot(pd, aes(x = idx))
q <- q + geom_linerange(aes(ymin = lo, ymax = hi, colour = covered), linewidth = 0.35)
q <- q + geom_point(aes(y = est), size = 0.25, colour = "grey20")
q <- q + geom_hline(aes(yintercept = truth), linewidth = 0.4)
q <- q + facet_wrap(~scenario_lab, ncol = 1)
q <- q + scale_colour_manual(values = c(`TRUE` = "grey65", `FALSE` = "#C00000"), guide = "none")
q <- q + labs(x = "Replicate (sorted by point estimate)", y = "log-IRR, 95% CI")
q <- q + theme_minimal()
q
```

Where the estimand's assumptions hold (s1), coverage is `r x34$c1`: the
sandwich variance is calibrated. Under confounding with independent loss
(s2) coverage dips mildly to `r x34$c2`, the typical price of treating
estimated weights as fixed (2.8). Under informative loss (s3) coverage
degrades to `r x34$c3`, and Table 15 together with Figure 2 shows why: the
mean bias (`r x34$b3`) displaces the entire distribution of estimates, so
intervals of the correct width miss the truth from one side. A correct
standard error cannot rescue a biased point estimate; the remedy is the
per-protocol estimand with censoring weights, not a wider interval.

## 3.7 Marginal versus conditional estimands

swereg and `TrialEmulation` both remove baseline confounding, but by
different routes, producing two distinct — each valid — estimands. swereg
weights and fits a covariate-free model: a marginal effect. `TrialEmulation`
conventionally adjusts the outcome model: a conditional effect. Rate ratios
are collapsible, so these coincide for the IRR; odds ratios are not, so the
`TrialEmulation` OR is converted with the Zhang–Yu relation
$\mathrm{RR} = \mathrm{OR} / (1 - p_0 + p_0\,\mathrm{OR})$, where $p_0$ is
the reference-arm per-period risk, before comparison. The conversion removes
the scale gap only; a residual conditional-versus-marginal difference
remains, visible in Table 4 as the small swereg − TE gaps in the confounded
scenarios (larger for ITT than for PP). The primary correctness guarantee is
each implementation's agreement with the known simulated truth on its own
scale.

---

# 4. Implementation mapping

The SAP (Section 2) is deliberately implementation-agnostic. This section
reveals the code: which function, argument, and option realises each step,
the provenance of estimator-behaviour changes, and where the validation
evidence comes from.

## 4.1 SAP step → code

| SAP | Step | Implementation |
|:--|:--|:--|
| 2.1 | Band width $w$ | `period_width` (default 4 weeks) in the trial-band assignment inside `TTEPlan` |
| 2.1 | Sequential eligibility, enrollment, matching | `TTEPlan$s1_generate_enrollments_and_ipw()`; `matching_ratio` and `seed` from the YAML spec's `treatment.implementation` |
| 2.1 | Washout / new-user exclusion | Spec-level exclusion: `type: no_prior_intervention` with `window: lifetime_before_baseline`, or a finite `window` in weeks |
| 2.1 | Prevalent-user warning | `tteplan_read_spec()` warns when a spec lacks a washout exclusion; silence with `options(swereg.warn_prevalent_user = FALSE)` |
| 2.3 | Follow-up stop events, event priority | `TTEEnrollment$s5_prepare_outcome()`; horizon from `follow_up`, administrative end of study from `admin_censor_isoyearweek` |
| 2.4 | Hot-deck imputation | `TTEEnrollment$s1_impute_confounders(seed = 4)` |
| 2.4 | Stabilised IPW | `TTEEnrollment$s2_ipw(stabilize = TRUE)` |
| 2.5 | IPCW censoring model | `TTEEnrollment$s6_ipcw_pp()` via `s4_prepare_for_analysis(estimate_ipcw_pp_with_gam = TRUE, estimate_ipcw_pp_separately_by_treatment = TRUE)`; GAM engine `mgcv::bam(..., discrete = TRUE)`; `estimate_ipcw_pp_with_gam = FALSE` gives the linear-in-time sensitivity variant |
| 2.6 | Weight truncation | `TTEEnrollment$s3_truncate_weights(lower = 0.01, upper = 0.99)`; truncated columns `ipw_trunc` (ITT) and `analysis_weight_pp_trunc` (PP product weight); untruncated PP results exported as a sensitivity sheet |
| 2.7–2.8 | Outcome model + inference | `TTEEnrollment$irr(weight_col)`: `survey::svydesign(ids = ~person)` + `survey::svyglm(family = quasipoisson())` with `splines::ns()` terms for follow-up and trial index |
| 2.10 | Pre-specification | YAML spec parsed by `tteplan_read_spec()`; full grid run by `TTEPlan$s1_…`/`s2_…`/`s3_analyze()` |

## 4.2 Provenance notes

- **Event-priority convention (2.3).** Enforced since swereg 26.7.3.
  Previously, a first event falling in the same band as the first protocol
  deviation was dropped as a censoring, which undercounted per-protocol
  events in switching-heavy data (≈10% of PP events in a high-switching
  simulation; negligible under strong persistence).
- **Per-person administrative censoring.** `admin_censor_var` (a per-person
  censoring column) is accepted by the constructor for backward compatibility
  but is not implemented in outcome preparation and now errors loudly rather
  than silently doing nothing; use `admin_censor_isoyearweek`.

## 4.3 Where the validation numbers come from

The evidence layers in Section 3 are permanent, executable tests:

| Section | Layer | Test file | Gate |
|:--|:--|:--|:--|
| 3.3 | Cross-package triangle | `tests/testthat/test-tte_validation_matrix.R` | runs in CI |
| 3.4 | Stress matrix | `tests/testthat/test-tte_stress_matrix.R` | fast subset in CI; full battery `SWEREG_RUN_STRESS=true` |
| 3.5 | Plan-layer truth matrix | `tests/testthat/test-tteplan_truth_matrix.R` | reduced-N subset in CI; full factorial `SWEREG_RUN_PLAN_MATRIX=true` |
| 3.6 | Coverage calibration | `tests/testthat/test-tte_coverage.R` | opt-in only, `SWEREG_RUN_COVERAGE=true` |

The tables and figures themselves are rendered from
`vignettes/tte-validation-evidence.rds`, regenerated by
`dev/generate_validation_evidence.R` (in the source repository, not the
installed package). The script reruns every cell through the same
DGP/truth/fit helpers the tests source (`tests/testthat/helper-tte_*.R`), so
the vignette's numbers and the suite's assertions cannot drift apart; rerun
it after any estimator change and commit the refreshed artifact alongside.

## References

- Hernán MA, Robins JM. Observational studies analyzed like randomized
  experiments: an application to postmenopausal hormone therapy and
  coronary heart disease. *Epidemiology* 2008;19(6):766–779.
- Hernán MA, Robins JM. Using big data to emulate a target trial when a
  randomized trial is not available. *Am J Epidemiol* 2016;183(8):758–764.
- Danaei G, García Rodríguez LA, Cantero OF, Logan R, Hernán MA.
  Observational data for comparative effectiveness research: an emulation
  of randomised trials of statins and primary prevention of coronary heart
  disease. *Stat Methods Med Res* 2013;22(1):70–96.
- Caniglia EC, et al. Emulating a sequence of target trials to avoid
  immortal time bias: an application in pregnancy. *Am J Epidemiol* 2023.
- Cashin AG, et al. Emulating a target trial — the TARGET statement.
  *JAMA* 2025.
- Su L, Rezvani R, Seaman SR, Bartlett JW. *TrialEmulation: An R package
  to emulate target trials for time-to-event data from electronic health
  records.* arXiv:2402.12083, 2024.
- Zhang J, Yu KF. What's the relative risk? A method of correcting the odds
  ratio in cohort studies of common outcomes. *JAMA* 1998;280(19):1690–1691.
